### 2020-07-22 AnalysisOutputLatex.tex
Initial Overleaf Import

%<*ClusterCorrelationSummary1>
Flat clusters were linearly correlated above chance levels (Fisher's method: $\chi^{2}(7844, N = 3922) = 15380.2, $ $p < .001$; Fig. 3e). Of the final angled clusters, 97.6\% (160/164) were positively correlated (Fig. 3f), suggesting the presence of a factor affecting response latency monotonically across the population. For the corresponding flat cluster of the criteria fulfilling angled clusters, spike pair differences were correlated with the spike time of the first unit in the pair (Fisher's method: $\chi^{2}(328, N = 164) = 8366.9, $ $p < .001$; Fig. 3g), showing that correlations were not explained by a model of the form $s_1 = s_0 + d + independent\_noise$ where $d$ is a fixed difference.
%</ClusterCorrelationSummary1>

%<*ClusterCorrelationSummary2>
For the final angled clusters, 81.7\% (134/164) of the corresponding flat clusters had linear regression p-values less than the p-values of linear regression applied to the unclustered first spike pairs (Fig. 3h). This implies that without clustering, linear regression would miss a large proportion of first spike correlations.
%</ClusterCorrelationSummary2>

%<*AngledClustersUniquePairsSummary>
Within this time window, the algorithm extracted a total of 164 angled clusters from 164 unique response distributions, such that there were no repeated or similar clusters.
%</AngledClustersUniquePairsSummary>

%<*OriginalAnglesSummary>
For the positively correlated angled clusters, Fig. 4a shows the distribution of the mean $\theta_{45}$ angles with 95\% confidence intervals. 66.5\% (109/164) of angles were significantly greater than $0^{\circ}$ (p-value < 0.025) and less than $45^{\circ}$ (p-value < 0.025). 
%</OriginalAnglesSummary>

%<*OriginalTestsPassedSummary>
Of the 164 angled clusters, 44 were determined to be stationary and sufficiently correlated by fulfilling the criteria described in the Methods \ref{methods:angle_and_factor_analysis_criteria}, such that an underlying factor could underlie the covariation.
%</OriginalTestsPassedSummary>

%<*OriginalTestsPassedAngleSummary>
Of the 44 criteria fulfilling angled clusters, 26 had estimated angles significantly different from both $0^{\circ}$ (p-value < 0.025; Fig. 4f) and $45^{\circ}$ (p-value < 0.025; Fig. 4f). Of the 44 criteria fulfilling angled clusters, normality was not rejected for 18 clusters (Henze-Zirkler p > 0.05), of which 11 had an estimated angle significantly different from both $0^{\circ}$ (p-value < 0.025) and $45^{\circ}$ (p-value < 0.025).
%</OriginalTestsPassedAngleSummary>

%<*OriginalTestsPassedFactorAnalysisSummary>
Factor analysis could be applied to the 44 criteria fulfilling angled clusters, to give an estimate of the factor-conditioned cluster response distributions for the stationary angled clusters.
%</OriginalTestsPassedFactorAnalysisSummary>

%<*ClusterSingleUnitKPSSStationaritySummary>
For the 164 angled clusters, 118 of the corresponding 328 ($=2*164$) single unit cluster first spike sequences were determined as KPSS non-stationary (p < 0.05; Fig. 4d).
%</ClusterSingleUnitKPSSStationaritySummary>

%<*ClusterSingleUnitTrialIndexLRSummary>
The cluster spike times of either 0, 1 or 2 of the units in a cluster pair were correlated with trial index (Fisher's method: $\chi^{2}(656, N = 328) = 2551.6, $ $p < .001$; Fig. 4b,c).
%</ClusterSingleUnitTrialIndexLRSummary>

%<*SelectivelyDifferencedSummary>
In total, 87 clusters had differencing applied to atleast one unit. 87 of these clusters fulfilled the criteria, giving a total of 87 criteria fulfilling angled clusters (including the undifferenced stationary clusters). 
%</SelectivelyDifferencedSummary>

%<*SelectivelyDifferencedCorrelatedSummary>
In total, 59 of the 87 criteria fulfilling clusters were linearly correlated following differencing (p-value < 0.05) (Fig. 6h). 26 of the correlated criteria fulfilling clusters had estimated angles significantly greater than $0^{\circ}$ (p-value < 0.025) and less than $45^{\circ}$ (p-value < 0.025). Of these, normality was not rejected for 15 clusters (Henze-Zirkler p > 0.05). Fig. 6j shows that the distribution of estimated angles for the successfully differenced clusters. Fig. 8? shows the difference between original and differenced estimated clusters?
%</SelectivelyDifferencedCorrelatedSummary>

%<*SelectivelyDifferencedTestsPassedNewSummary>
In total, 87 clusters had differencing applied to atleast one unit. 41 of these were correlated and fulfilled the criteria. 26 of these clusters had estimated angles significantly greater than $0^{\circ}$ (p-value < 0.025; Fig 6b) and less than $45^{\circ}$ (p-value < 0.025; Fig 6b), of which 14 normality was not rejected for (Henze-Zirkler p < 0.05). These angles are similar to the original angles before differencing (Fig. 6c), showing that the differencing operation does not largely alter cluster angle. Fig. 6d shows the amount of variance explained by the pairwise correlation after trend removal. 
%</SelectivelyDifferencedTestsPassedNewSummary>

%<*SelectivelyDifferencedTestsPassedEigenvalueSummary>
The factor analysis model from above can be applied to the 41 correlated criteria fulfilling differenced clusters, as the factor correlation matrices of each cluster only had one eigenvalue greater than 1. 
%</SelectivelyDifferencedTestsPassedEigenvalueSummary>

%<*SelectivelyDifferencedBoxJenkinsDifferencedSummary>
85 of the 87 differenced clusters had an ARMA model applied to atleast one unit. Almost all ARMA models fit for differenced clusters were of the MA type (Fig 7a), as expected for the negative autocorrelations introduced by differencing. 
%</SelectivelyDifferencedBoxJenkinsDifferencedSummary>

%<*SelectivelyDifferencedBoxJenkinsCorrelatedSummary>
After the described application of the Box-Jenkins method, there were 44 criteria fulfilling clusters. Normality was not rejected for 18 of these clusters (Henze-Zirkler p > 0.05). 26 of the correlated criteria fulfilling clusters had estimated angles significantly greater than $0^{\circ}$ (p-value < 0.025; Fig. 7b) and less than $45^{\circ}$ (p-value < 0.025; Fig. 7b), for which 9 normality was not rejected for (Henze-Zeckler p-value > 0.05). Fig. 7c shows the difference between original and ARIMA estimated cluster angles. 
%</SelectivelyDifferencedBoxJenkinsCorrelatedSummary>

%<*SelectivelyDifferencedBoxJenkinsTestsPassedEigenvalueSummary>
The factor analysis model from above can be applied to the 44 correlated criteria fulfilling clusters, as the factor correlation matrices of each cluster only had one eigenvalue greater than 1. 
%</SelectivelyDifferencedBoxJenkinsTestsPassedEigenvalueSummary>

### 2020-07-22 manuscript.tex
Initial Overleaf Import

\documentclass{article}
\usepackage{graphicx,subfigure}

\usepackage{arxiv}

\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography
\usepackage{lipsum}
\usepackage[dvipsnames]{xcolor}
\usepackage{textcomp}
\usepackage{catchfilebetweentags}

\usepackage[shortlabels]{enumitem}
\usepackage{amsmath}
\usepackage{bm}

\usepackage[colorinlistoftodos]{todonotes}

% \usepackage[nomarkers,notablist,nofiglist]{endfloat}
% \usepackage[nomarkers,notablist,nofiglist]{endfloat}
% \usepackage[colorinlistoftodos,disable]{todonotes}

\usepackage[english]{babel}
\usepackage[
backend=biber,
style=nature,
% citestyle=authoryear
]{biblatex}
\addbibresource{references.bib} %Imports bibliography file

\title{evidence of stationary and non-stationary clustered time-warped spike timing in cortical networks}

\author{
  James B. Isbister\\
  University of Oxford, United Kingdom
%   \thanks{Oxford Centre for Theoretical Neuroscience and Artificial Intelligence, University of Oxford, UK.}    
% \thanks{Blue Brain Project, EPFL, Switzerland.}\\
%   University of Oxford, UK.\\
%   EPFL, Switzerland.\\
%   \texttt{james.isbister@psy.ox.ac.uk}\\
%   \textit{isbisterjb@gmail.com}\\
  \And
              \\
              \\
  \And
  Vicente Reyes-Puerta\\
  Johannes Gutenberg University, Germany

%   \And
%   Irina Higgins\\
%   DeepMind

  \And
  Jyh-Jang Sun\\
  NERF, Belgium

  \And
  Heiko J. Luhmann\\
  Johannes Gutenberg University, Germany

}

\begin{document}
\maketitle

\begin{abstract}
% http://blogs.nature.com/naturejobs/2014/11/03/how-to-get-published-in-high-impact-journals-big-research-and-better-writing/
% \color{blue}`Do not undersell your research. “If you find something big, tell us!'\color{Black}

REL: Efficient population coding of naturalistic whisker motion in the ventro-posterior medial thalamus based on precise spike timing

Sensory stimulation induced patterns of action potentials recorded in adult rat cortex in vivo allow investigation of neural coding principles.

Patterns of action potentials induced by sensory stimulation in vivo allow investigation of neural coding principles. Trial-to-trial variability engenders debate over temporal and rate codes. Characterising distributions of stimulus responses enables discovery of response properties, which reliably encode the stimulus and may affect other processes. Non-stationarity (i.e. trial dependence) and noise correlations complicate such characterisation. A novel algorithm found correlated clusters above chance in first spike stimulus-response distributions for pairs of barrel cortex units. First spike times covaried at angles different from $45^{\circ}$ for a proportion of stationary and non-stationary (after removal of trial-dependence) clusters, suggestive of an underlying factor warping the distribution of decodable relative spike time differences from trial-to-trial. Factor analysis allowed estimation of factor-dependent and trial-independent stimulus response distributions, with a high estimated potential for discriminable stimulus encoding. The results suggest that consideration of clustering, factor-dependence and non-stationarity, together, are necessary to explain a large proportion of neural variability.

\end{abstract}

\keywords{Neural coding, precise spike timing, time-warp, feature-binding, barrel cortex, noise correlations, neural adaptation, in vivo recording}

\section{Introduction}
The neural coding problem asks how information is encoded through patterns of action potentials in the nervous system \cite{gerstner1997neural, johnson2000neural, stein2005neuronal, gerstner2014neuronal}. Responses following sensory stimulation can be analysed to explore principles of neural coding, with a classic dichotomy between temporal \cite{uzzell2004precision, johansson2004first, gollisch2008rapid, storchi2012comparison, reinagel2002precise, reyes2014laminar} and rate codes \cite{hong2016explicit, chang2017code, kar2019evidence, stringer2019high}. Temporal codes provide the possibility of a fast and efficient code. Whilst precise temporal patterns have been observed \cite{uzzell2004precision, johansson2004first, gollisch2008rapid, storchi2012comparison, reinagel2002precise, reyes2014laminar}, a large proportion of trial-to-trial variability remains unexplained. This high variability leads to common assumption of population firing rate codes, under the premise that variability cancels out across a population \cite{abbott1999effect, moreno2014information, stringer2019high}. Average firing rate codes are abstracted from neural mechanisms (see below), leaving a full understanding of neural coding and variability undetermined.

It is important to better characterise neural codes in order to understand a rich variety of dynamical, structural, homeostatic and cognitive processes. Characterising responses may also uncover response properties involved in behaviour. Due to response variability from trial-to-trial, characterising the neural representation of a single stimulus is a matter of characterising the \textit{distribution} of possible neural responses.

A multi-neuron stimulus response can be expressed as a point in multi-dimensional space, where each dimension represents a spike train metric. Responses from multiple presentations of a stimulus form a sample stimulus-response distribution in this space, which approximates an underlying stimulus-response distribution (Fig. 1a). If two stimuli have a behaviourally relevant difference, downstream neurons must be able to respond differently to samples from the response distribution of either stimulus. This is possible if the response distributions are highly non-overlapping in dimensions that downstream neurons are sensitive to, such as relative spike time differences (\cite{gasparini2006state, branco2010dendritic, branco2011synaptic}; Fig. 1a).

Two phenomena complicate stimulus response distribution characterisation. Firstly, responses of some neurons depend on trial index. That is, response distributions are \textit{non-stationary} due to mechanisms of adaptation \cite{dragoi2002dynamics, ulanovsky2003processing, sharpee2006adaptive, ahissar2000transformation, kheradpezhouh2017response, ahissar2001temporal, khatri2009stimulus, martin2014tactile, barros2019response}. It follows that the region from which stimulus-responses are drawn from at a single point in time is smaller than the region from which samples are drawn from over multiple trials (Fig. 1b).  Adaptation occurs across species and modalities \cite{dragoi2002dynamics, ulanovsky2003processing, sharpee2006adaptive}, and in barrel cortex to repetitive whisker stimulation for both spike count \cite{ahissar2000transformation, ahissar2001temporal, kheradpezhouh2017response, khatri2009stimulus, martin2014tactile, barros2019response} and spike latency \cite{ahissar2000transformation, ahissar2001temporal, kheradpezhouh2017response}.

\begin{figure}[t!]
\centering
\includegraphics[width=\textwidth]{Figure1.pdf}
\caption{Explanatory Figures. 
(a) Left: For two neurons (x and y-axis), illustration of sample (dots) and boundary of underlying (circle) first spike stimulus-response distributions for two hypothetical stimuli (red and gold). Right: Illustration of corresponding relative first spike differences. 
(b) Illustration of trial dependence of a sampled response distribution (dots). Circle illustrates the boundary of a possible trial-dependent response distribution on a single trial. 
(c) Illustration of boundary of factor-dependent first spike response distribution for a single factor value. The green line shows the mean points that would be predicted for a continuous set of factor values.
(d) Illustration of superposition of two single spike pattern feature representations.
(e) Illustration of `stationary clustered time-warped precise spike timing'. Left: Responses form several clear clusters. The main green cluster is correlated with non-unitary gradient (i.e. different from $45^{\circ}$). Under the assumption of stationarity, the green line and circle are the same as in (c). The orange ellipses illustrate the boundary of an underlying response distribution under the assumption of independence and normality for the sample (solid) and another stimulus (dashed). Centre: Illustration of spike time differences for values of the factor. The green circle illustrates the boundary of a factor-dependent distribution of spike time differences. Right: Conceptual illustration of time warp relative differences.}
\label{fig:fig1}
\end{figure}

Secondly, firing correlations between spatially distanced neurons (see \cite{reid2012functional, feldmeyer2013barrel} for review) suggest the possible presence of a \textit{continuous} underlying state covarying neural responses. Response samples on a single trial (when the continuous state has a certain value) would then come from a state-dependent stimulus-response distribution (i.e. a smaller region of response distribution space; Fig. 1c).
If receiving neurons could respond to a state-modulated signal in a state-dependent manner (contrary to common neural coding assumptions \cite{moreno2014information, stringer2019high}), state-dependent encoding would increase the capacity for discriminable stimulus encoding (Fig. 1c). The question is whether correlations between spatially distanced neurons are evidence of a shared state, which receiving neurons could also be modulated by [New Fig?].
% Correlations in the barrel cortex up to a distance of $\sim1.5mm$ (vertically and horizontally) \cite{reyes2015long} suggest it is possible that afferently connected neurons could also be state-modulated and respond in a state-dependent manner. 
Subthreshold membrane potentials are a candidate for such shared state, as they affect action potential initiation \cite{petersen2003interaction} and are correlated across areas \cite{poulet2008internal}.
% State information could alternatively be explicitly encoded in state modulated population activity.

Estimating state-dependent stimulus-response distributions at a single snapshot in time is made difficult by heterogeneity of non-stationarity and response distribution structures (Supplementary Vid. 1), leading to use of averaging and coarse spike train metrics. As \textit{spike counts} show high trial-to-trial variability, single units cannot reliably encode many different stimuli. Instead, time, trial or population averaged scalar \textit{firing rates} \cite{gerstner2014neuronal} are calculated, with capacity beyond the original single trial discrete spike count, under the assumption that noise cancels out across a population \cite{abbott1999effect, moreno2014information, stringer2019high}.

Trial-averaged firing rate population codes (often calculated over windows $>100ms$) allow decoding of object \cite{hong2016explicit} and face \cite{chang2017code} properties from macaque higher visual areas. 
% Such long integration windows are discordant with behavioural speed \cite{thorpe1996speed} and the fast propagation of activity through cortical layers and areas \cite{perrett1982visual}. 
Object category can be decoded population trial averaged multi-unit activity (MUA) over short 10ms time windows \cite{kar2019evidence}. Decoders do not generalise to later windows \cite{kar2019evidence}, excluding the notion of sustained firing rate representations. Higher encoding capacity is conferred by high spike counts of unsorted MUA \cite{hong2016explicit, kar2019evidence} or for persisting visual stimuli \cite{hong2016explicit, chang2017code, kar2019evidence}. Dimensionality reduction is commonly used to find dynamical population firing rate patterns \cite{ahrens2012brain, mante2013context, cunningham2014dimensionality}. Spike count and firing rates do not superimpose (Fig 1d\todo[color=red!40]{To do}), although competition schemes have been presented for simultaneously active overlapping representations \cite{bao2018representation}. 
% Rate and timing codes may be aligned to internal oscillations \cite{fries2007gamma, havenith2011synchrony, vinck2010gamma, vinck2013attentional}.

The topographic representation of each single whisker on the snout of a rodent in the somatosensory cortex, the so-called barrel cortex, represents an excellent model to study neuronal processing of defined sensory stimulation \cite{feldmeyer2013barrel}.
The majority of barrel cortex neurons respond 0 or 1 times to single whisker deflections \cite{reyes2014laminar}. First spike patterns allow direct study of the relationship between known entities (e.g. stimulus location, intensity, frequency) and neuronal activity. A recent optogenetics study showed that early visual cortex patterns of single spikes are sufficient to enable visual discrimination \cite{resulaj2018first}, implicating single spike patterns as a candidate for atomic representation.

First spike times can have millisecond order precision relative to stimulus onset in sensory organs \cite{johansson2004first, uzzell2004precision, gollisch2008rapid}, thalamic areas \cite{reinagel2002precise, storchi2012comparison} and early sensory cortices \cite{reyes2014laminar}, with some evidence of continuum-like encoding relative to onset \cite{gollisch2008rapid, storchi2012comparison} and oscillations \cite{havenith2011synchrony}. Relative differences \cite{izhikevich2006polychronization} observed experimentally \cite{johansson2004first, gollisch2008rapid, reyes2014laminar} have been shown to encode information when trial averaged \cite{gollisch2008rapid} and do not require reference to an onset signal.
% as neurons are sensitive to the spatial and temporal pattern of synaptic activations \cite{gasparini2006state, branco2010dendritic, branco2011synaptic}.
Action potential instantaneity suggests that relative spike codes could superimpose (Fig. 1d).

Assumptions about stimulus-response distribution structure constrain the potential encoding power of timing-based single spike patterns.
Independent and normally distributed first spike-times would confer a low discriminability potential (as will be seen), dependent on single neuron variability (Fig. 1e). 
% Over typical 10-40ms ranges of mean first spike times for short stimuli \cite{reyes2014laminar, resulaj2018first}, 
% Observed first spike time standard deviations \cite{reyes2014laminar} would imply that very few unique stimuli/features could be encoded with high discriminability by a neuron pair. 
Spike time differences between two neurons with independent and normal first spike times would have a standard deviation of \(\sigma_{diff} = \sqrt{\sigma_{a}^2 + \sigma_{b}^2}\).
Spike times could covary with unitary gradient under the assumption that relative spike times have a factor-independent mean. Precise relative differences have been reported in the eye \cite{gollisch2008rapid} and barrel cortex \cite{reyes2014laminar}. Response distributions may form multiple clusters, implying stochasticity in the cluster from which representations come (Fig. 1d; \cite{reyes2014laminar}).
% , which could be determined by some underlying factor or non-stationarity.

% Characterisation techniques for sparse (low trial count) datasets are often used, which make assumptions about distribution structure, risking fundamental distribution properties and encoding forms being overlooked.

% The spike time of either neuron would be correlated with the difference \todo[color=green!40]{Figure?}(Fig. X) but due to independence there would be no process underlying this correlation, and thus no means by which downstream neurons could interpret the difference in a state-dependent manner (other than an event signal). Moreover, the discriminability of one neuron's spike time and the difference would be the same as that of the two neuron's spike times \todo[color=green!40]{Same figure}(Fig. X).

% Projection of the stimulus-response distribution to a first spike time vs difference space would not decrease the area of the distribution, and thus would not improve discriminability.

% The standard deviation of spike times and spike time differences would not capture clustering structure, thus underestimating discriminability. If trial-wise cluster was determined by some state, then discriminability given a state, would equal the corresponding cluster's discriminability.

\section{Results}

\subsection*{Overview}
% 
This paper analyses first-spike stimulus-response distributions for pairs of barrel cortex units in response to repeated single whisker deflections. In first-spike stimulus-response distributions for pairs of units, correlated clusters were found above chance-levels. Clustering was necessary for the detection of a large proportion of correlations.
Cluster spike times of both units were stationary for X\%\todo[color=red!40]{Fill} of correlated clusters. X\% of these stationary correlated clusters had non-unitary (different from $45^{\circ}$) correlation angles, suggestive of an internal underlying factor co-varying first spike times and thus \textit{warping} the distribution of (decodable) relative spike time differences from trial-to-trial (Fig. 1d). Estimated factor-conditioned response distributions for these stationary clusters showed a high potential encoding capacity. 
% Algorithms for detecting correlated clusters and estimating correlation angles were developed for this study and verified on surrogate control datasets.

Time series analysis was applied to the non-stationary units of the remaining correlated clusters, to remove neural adaptive trends. A large proportion of detrended clusters were correlated with non-unitary gradient, suggestive of an underlying factor with an improved encoding potential. The results suggest that consideration of clustering, factor-dependence and non-stationarity, together, are necessary to explain a large proportion of neural variability, and could explain the failure to find a fast and efficient ubiquitous spike code that utilises the high sensitivity of neural architecture and plasticity. In this study, such consideration gave an estimate of the potential resolution of neural stimulus representation. 
% Although the study analyses responses to single stimuli in isolation, it also considers the proposed and previously suggested neural codes with respect to feature-binding and continua encoding.

The characterised response distribution structure is referred to as \textbf{stationary and non stationary clustered time-warped spike timing}. This differs from previous `time-warp' studies, which aim to align spike trains over trials and do not consider discriminable encoding, clustering, non-stationarity and different warping functions for different neurons \cite{williams2020discovering}. One previous study has modelled time-warp-invariant integration \cite{gutig2009time}, although neurons needn't respond identically to warps of the same pattern. Rather, neurons should process relative spike patterns in a factor-dependent manner.

\subsection*{Neurophysiology}

The dataset used for this paper is reused from an earlier barrel cortex study, of which analyses and data collection methods have previously been reported \cite{reyes2014laminar}. In summary, extracellular recordings were made in 15 anaesthetised Wistar rats using 16 or 128 electrode silicon probes, inserted perpendicularly into the barrel cortex. Single whisker deflections were made at various frequencies (0.1-10Hz) in blocks, each containing typically 200 trials (sweeps) with an intertrial interval of 5 s (occasionally 10, 15, or 30 s). This was repeated for each of the selected target whiskers (maximum 4). Experimental sessions varied in terms of electrode placement, stimulated whiskers, stimulation frequencies and number of trials per unique stimulus. 7 experimental sessions used single shanks with 16 electrodes, whilst 8 experimental sessions used 8 neighbouring shanks (16 electrodes per shank) covering 2-3 neighbouring barrels. Spike sorting classified units by E/I type and cortical layers L2/3, L4, L5A, L5B (see Methods \ref{methods:spike_sorting}).

\begin{figure}[t!]
\centering
\includegraphics[width=\textwidth]{Figure2.pdf}
\caption{Initial response and first spike time non-stationarity. (a) Spike count histograms (not including 0 spike count) for the 6 neuron groups over 25ms (left) and 100ms (right) following cortical response (responses pooled over all stimuli and trials). (b) Normalised PSTHs following stimulus onset of 6 neuron groups averaged over all stimuli and trials. Cortical onset is marked by the green line. (c) Spike count mean vs variance for all stimuli and unit combinations. (d) First spike rasters following cortical onset for five unit stimulus combinations. (e) P-values of linear regressors fit to predict first spike time from trial index for all stimulus and unit combinations. (f) $R$ (green) and $R^2$ (blue) of linear regressors fit to predict first spike time from trial index for all stimulus and unit combinations. (g) First spike time KPSS p-values for all stimulus and unit combinations. Implementation of p-value estimation rounds all p-values $>0.1$ to 0.1.}
\label{fig:universe}
\end{figure}

\subsection*{Low spike counts}

The shortness of whisker deflections reduces ambiguity in what temporal aspect of a stimulus neural activity represents, offering relatively controlled study of the neural representation of atomic entities. Excluding non-spiking responses over 25 and 100ms following cortical onset (5.2ms following stimulus onset; Supplementary Fig. 1), responses to single whisker deflections in barrel cortex are largely single spike responses (Fig. 2a). Spike count encoding therefore has a low capacity for discriminable stimulus encoding, made worse by high spike count variance (Fig. 2b). It is possible that double spike responses are examples of superpositions of single spike representations. The responses of each unit-type and layer group (averaged over stimuli) demonstrate a clear initial response, followed by a tailing-off (Fig. 2c). The paper aims to better characterise this initial (largely single spike) response.

\subsection*{First spike non-stationarity}

Fig 2d. shows the first spike times of 5 single units in response to repeated presentations of single stimuli. The unit of Fig. 2d(i) demonstrates depressive latency adaptation. The units of Fig. 2d(ii, iii) demonstrate no obvious latency adaptation. The units of Fig. 2d(iv, v) demonstrate more complex response-distribution structure and possible non-linear non-stationarity. For repeated presentations single stimuli, single unit first spike times are linearly correlated with trial number above chance 
(Fisher's method: $\chi^{2} (1786, N = 893) = 3672.69, p<.001$; Fig. 2e).
% (Fisher's method: $\chi^{2} (1786, N = 893) = 3672.69, p = 1.027\textrm{e-132}$; Fig. 2e). 
Of the 63 first spike response distributions that were significantly correlated with trial index (p<.005), $90\%$ (57/63) were positively correlated (Fig. 2f), supporting depressive latency adaptation \cite{ahissar2000transformation, ahissar2001temporal, kheradpezhouh2017response}. The KPSS null hypothesis of stationarity was rejected for p-value < 0.05 for 19\% (168/893) single stimulus single unit first spike response distributions (Fig. 2g).

Autocorrelations of single-unit single-stimulus first spike times were above chance for lags 1-5 (Supplementary Fig. 2), supportive of general adaptive trends. Only the lag-2 partial autocorrelation was slightly above chance (Supplementary Fig. 2), suggestive of some additional structure in the trial-to-trial sequence of first spike times for some units.
% after accounting for general adaptive trends.

\subsection*{First spike pair response distribution example}

\begin{figure}[t!]
\centering
\includegraphics[width=\textwidth]{Figure3.pdf}
\caption{Clustered time-warp example and correlated clusters. (a) Sampled first spike response distribution for two units (L5A Inh, L4 Inh) over a 50ms window following cortical onset (1 outlier out of view). Point colours illustrate DBSCAN cluster (EPS = 1.35). Black points are outliers. Dashed ellipses illustrate $3\sigma$ bounding ellipse calculated for each DBSCAN cluster. Green flat ellipse illustrates $4\sigma$ boundary of green flat cluster. Green angled ellipse illustrates $4\sigma$ boundary of green angled cluster. Orange ellipse illustrates $4\sigma$ ellipse of points under the assumption of independence and normality. (b) Distribution of spike pair differences for response distribution of (a). (c) Histogram of flat cluster end points (mean + $4\sigma$) for all criteria fulfilling angled clusters. (d) The blue histogram represents the ratio of cluster spike pairs to the number of trials of the stimulus, for each angled cluster. The green histogram represents the ratio of cluster spike pairs to the number of trials that both neurons spike to the stimulus, for each angled cluster. (e) Linear regression p-values for all flat clusters. (f) R (green) and $R^2$ (blue) values of all angled clusters. (g) Linear regression p-values of first unit cluster spike times and spike difference. (h) Histogram of flat cluster linear regression p-values minus conjunctive linear regression p-values, for all criteria fulfilling angled clusters.}
\label{fig:fig3}
\end{figure}

Fig. 3a shows the distribution of first spike pairs in response to a 4Hz stimulus for two units (L5A Inh, L4 Inh) over a 50ms window following cortical onset. Only responses on `conjunctive trials' (trials on which both units spiked) are shown. The orange Gaussian is centered at the mean and has widths equal to 4 times the first spike time standard deviations of each unit (on conjunctive trial), illustrating the region from which 99.9\% of first spike pairs would be predicted to fall under the assumption of independence and normality. The assumptions fail to capture the distribution structure of many spike pair distributions (Supplementary Vid. 1) and would allow very few stimuli to be encoded with high-discriminability. The spike time differences have a standard deviation of X (Fig. 3b), which is also poor for discriminable encoding.

First spike pairs in Fig. 3a appear to form two clusters (blue and green dots). The green cluster appears correlated with non-unitary gradient, suggestive of an underlying factor, which covaries first spike times, such that distributions of differences would be smaller when factor-conditioned, offering a higher power for discriminable encoding.

\subsection*{Clustering algorithm summary}
A two-stage algorithm was developed to quantify the occurrence and angle of correlated clusters. 
Stage 1 extracted well isolated \textit{`flat clusters'} defined by the $4\sigma$ boundary of flat 2D Gaussian ellipses (flat solid green ellipse in Fig. 3a; see Methods \ref{methods:flat_clusters} for details).  
% The algorithm applied the unsupervised DBSCAN clustering method to first spike response distributions for a range of values of the DBSCAN EPS free parameter. Repeated clusterings for different values of the EPS free parameter were removed from analysis. For DBSCAN clusters that were well isolated and had more than 20 elements, a bounding ellipse was defined from the means and standard deviations of points in the DBSCAN cluster (flat solid green ellipse in Fig. 3a). Points in this bounding ellipse were assigned to the `flat cluster'. 
To assess the occurrence of correlated clusters without sample bias, repeated and similar flat clusters produced by non-identical DBSCAN clusterings were left in the final group of flat clusters (Methods \ref{methods:flat_clusters}). When the algorithm was verified on surrogate control datasets (see Methods \ref{methods:correlation_surrogate_control_datasets}) correlated clusters were found at chance levels (Supplementary Fig. 3).

Stage 2 of the clustering algorithm aimed to better estimate the shape of the distributions underlying correlated flat clusters (p-value < 0.005), by estimating an `angled cluster' ellipse (angled solid green ellipse in Fig. 3a; see Methods \ref{methods:angled_clusters}). Repeats, similar and non-isolated angled clusters were discarded. 
\ExecuteMetaData[100msAnalysisOutputLatex.tex]{ClusterTimeSpanSummary} \ExecuteMetaData[AnalysisOutputLatex.tex]{AngledClustersUniquePairsSummary}
X\%\todo[color=red!40]{Fill} of angled clusters contained spikes from all conjunctive trials (Fig. 3d).

\subsection*{Correlation analysis}
\ExecuteMetaData[AnalysisOutputLatex.tex]{ClusterCorrelationSummary1}
\ExecuteMetaData[AnalysisOutputLatex.tex]{ClusterCorrelationSummary2}

\subsection*{Cluster angle}
\label{results:cluster_angle}

\begin{figure}[t!]
\centering
\includegraphics[width=\textwidth]{Figure4.pdf}
\caption{(a) Mean $\theta_{45}$ angles with 95\% confidence intervals for criteria fulfilling positively correlated angled clusters. Angles are coloured blue if they are significantly greater than $0^{\circ}$ (p-value < 0.025) and less than $45^{\circ}$ (p-value < 0.025), and red otherwise. [Gap shown by orange circle to be removed.] (b) Correlation p-values of angled cluster single unit spike times with trial index. (c) Correlation p-values of angled cluster single unit spike times with trial index (unit 0 vs unit 1). (d) $R^2$ values of multi-output regression (trial index vs spike pairs). (e) Cluster single unit KPSS p-values. Implementation of p-value estimation rounds all p-values $>0.1$ to 0.1. (f) Estimated angles of criteria fulfilling angled clusters. (g) Linear regression $R^2$ values for criteria fulfilling angled clusters.}
\label{fig:universe}
\end{figure}

An algorithm was developed to estimate a mean angle and confidence interval for each angled cluster (see Methods \ref{methods:cluster_angle_confidence_interval} for details), to test whether the angles of angled clusters were significantly different from $45^{\circ}$ (possibly indicative of factor or trial-dependent spike time differences). The mean angles, $\theta$, of positively correlated angled clusters (between $0^{\circ}$ and $90^{\circ}$) were of interest. $\theta_{45} = min(\theta, 90-\theta)$ were calculated for analysis, as only the ordering of the units in the pair determined whether $\theta$ was above or below $45^{\circ}$.

\ExecuteMetaData[AnalysisOutputLatex.tex]{OriginalAnglesSummary}
The algorithm was tested on a surrogate control dataset to validate that the detection of such differences from $45^{\circ}$ were not artifacts of the algorithm (see Methods \ref{methods:angle_surrogate_control_dataset}; Supplementary Fig. 6). The following subsections aim to determine when trial number and/or another factor underlie such non-unitary covariation.

\subsection*{Cluster stationarity and normality}

\ExecuteMetaData[AnalysisOutputLatex.tex]{ClusterSingleUnitTrialIndexLRSummary}
Only a small amount of variance was explained by the linear models fit to predict the joint spike pair times from trial number (Fig. 4d), suggesting that neural adaptation may not alone explain the covarying cluster spike times. 
% More complex first spike time relationships with trial index were also observed. 
\ExecuteMetaData[AnalysisOutputLatex.tex]{ClusterSingleUnitKPSSStationaritySummary}

Beside general adaptive trends, some units exhibited observable short trial-by-trial sequences (3-4 trials) of highly correlated first spike times, before losing such correlations. Lagged autocorrelations were significantly above chance for a high number of lags (Supplementary Fig. 4) supportive of general adaptive trends. Only the lag-2 partial autocorrelation was significantly above chance (p-value < 0.005; Supplementary Fig. 4), representative of the additional trial-by-trial structure observed for very few single units.

The cluster spike times of cluster unit pairs exhibit lagged cross-correlations above chance for lags 1-5 (Supplementary Fig. 5), as expected for general adaptive trends. Only the lag-2 partial cross-correlation was above chance (Supplementary Fig. 5). This suggests that the individual adaptive trends of cluster pairs explains the majority of above chance cross-correlations, whilst there may be some finer sequential structure between a few unit pairs.

% Normality was rejected for X of Y single unit cluster spike times (Shapiro-Wilks p < 0.05; Supplementary Fig. 7) and X out of Y 2-dimensional angled clusters (Henze-Zirkler p < 0.05; Supplementary Fig. 7).

\subsection*{Analysis of stationary clusters}
\ExecuteMetaData[AnalysisOutputLatex.tex]{OriginalTestsPassedSummary}
\ExecuteMetaData[AnalysisOutputLatex.tex]{OriginalTestsPassedAngleSummary}
Fig. 4g shows the amount of variance explained by the stationary angled cluster correlations. Autocorrelations, cross-correlations, partial autocorrelations and partial cross-correlations were at chance levels for the stationary clusters (Supplementary Fig. 7).

\begin{figure}[t!]
\centering
\includegraphics[width=\textwidth]{Figure5.pdf}
\caption{(a) Sampled first spike response distribution for two units (L5B Exc, L5A Exc). Point colours illustrate DBSCAN cluster (EPS = 5.0). Black points are outliers. Green flat ellipse illustrates $4\sigma$ boundary of green flat cluster. Green angled ellipse illustrates $4\sigma$ boundary of green angled cluster. Yellow ellipse illustrates $4\sigma$ ellipse of points under the assumption of independence and normality. Purple circle illustrates the $4\sigma$ boundary of the factor-conditioned response distribution predicted by factor analysis for a single factor value. The purple line shows the mean points predicted for the factor values [-4, 4] (i.e. with variance equal to 0). (b) PCA angles vs factor analysis angles for the criteria fulfilling clusters. (c) The standard deviations of the flat clusters vs standard deviations of the factor-dependent response distributions predicted by factor analysis (for units $i = 0,1$ separately). (d) The standard deviations of the conjunctive trial first spike times vs the standard deviations of the factor-dependent response distributions predicted by factor analysis (for units $i = 0,1$ separately). (e) The standard deviation of differences on conjunctive trials vs the factor-dependent standard deviation of differences predicted by factor analysis.}
\label{fig:universe}
\end{figure}

\ExecuteMetaData[AnalysisOutputLatex.tex]{OriginalTestsPassedFactorAnalysisSummary} The following sections apply time series analysis techniques to remove the neural adaptive trends of the non-stationary units of the remaining clusters. The factor correlation matrix of all criteria fulfilling angled cluster had only one eigenvalue greater than 1, suggesting that a single factor underlied the pairwise cluster covariations. The following single factor model was therefore fit to the elements of each (zero-centered) cluster: $s_i = \lambda_i \eta + \sigma_i$ for $i \in [0, 1]$ 
% and $s_1 = \lambda_1 \eta + \sigma_1$, 
where $\eta$ is some underlying factor, 
% $\lambda_i$ is the gradient at which the factor affects the spike time of unit $i$, 
and $\sigma_i$ is the predicted standard deviation of $s_i$ given $\eta$. Bootstrapping (1000 iterations) was used to calculate means of these values, which were used in subsequent analyses.

The purple line in Fig. 5a shows the values predicted by the model for $\eta = [-4, 4]$ and $\sigma_i = 0$, re-centered around the cluster mean. $\theta_{45}$ angles of the factor lines were estimated using the same method used for PCA component angle estimation (Methods \ref{methods:cluster_angle_confidence_interval}) and were similar to the corresponding PCA $\theta_{45}$ angles (Fig. 5b). The purple ellipse in Fig. 5a has widths equal to $4\sigma_0, 4\sigma_1$ where  $\sigma_0, \sigma_1$ are the standard deviations of the predicted factor ($\eta$) dependent response distribution. This allows visual comparison of the estimated factor-conditioned response distribution to the larger $4\sigma$ areas under the assumptions of normality and independence (orange), normal and independent clusters (green flat cluster) and normal correlated clusters (green angled cluster).

The areas of the two-dimensional (i.e. two unit) factor dependent response distributions show a striking improvement in potential encoding power (Supplementary Vid. 1; Supplementary Fig. 8). The standard deviations $\sigma_i$ of the factor analysis predicted response distributions for units $i = 0,1$ separately are also strikingly smaller than those assumed by the simpler assumptions (Fig. 5c,d). The factor dependent predicted standard deviations, $\sigma$, are less than 1 for X\% of units and less than 2 for Y\% of units\todo[color=red!40]{Fill}. 
% This gives an estimate of potential encoding power for single units, given by the factor dependent response distributions. 
% It is necessary to consider the results in terms of relative spike timing differences, which can be explicitly processed by downstream neurons.
As the factor analysis model assumes that noise for each unit given $\eta$ is normally distributed and independent between units, the state-dependent difference standard deviation predicted by the model is equal to \(\sigma_{diff} = \sqrt{\sigma_{0}^2 + \sigma_{1}^2}\).
% , where  $\sigma_0, \sigma_1$ are the standard deviations predicted by factor analysis given state $\eta$. 
The estimated $\sigma_{diff}$ values are strikingly lower than the original standard deviation of differences (Fig. 5e).

% The factor analysis model therefore predicts factor-dependent distributions of differences with a higher potential for discriminable continua encoding than both unclustered spike time differences and clustered but factor-independent spike time differences.

\begin{figure}[t!]
\centering
\includegraphics[width=\textwidth]{Figure6.pdf}
\caption{(a) This subfigure analyses the spike times of the green cluster from Fig. 3a for which the spike times of one unit were non-stationary. The columns correspond to (1) the original angled cluster spike times, (2) the cluster after the non-stationary unit cluster spike times were differenced, and (3) the result of applying ARMA modelling to the differenced cluster. Row 1 shows the cluster spike times of both units. Row 2 shows the corresponding cluster, with point colour illustrating trial index (from dark to light). The clusters fulfil the criteria after differencing (autocorrelation criteria relaxed) and ARIMA modelling. As such the $4\sigma$ angled bounding ellipse and factor analysis line and circle are presented. Rows 3 and 4 present the autocorrelation and partial autocorrelation plots respectively with 95\% confidence intervals. (b) Predicted angles and confidence intervals of the criteria fulfilling differenced clusters. (c) Original PCA angles vs differenced PCA angles for the criteria fulfilling differenced clusters. (d) Linear regression $R^2$ values for criteria fulfilling differenced clusters. (e) PCA angles vs factor analysis angles for the criteria fulfilling differenced clusters. (f) The standard deviations of the original flat clusters vs standard deviations of the factor-dependent response distributions predicted by factor analysis for the criteria fulfilling differenced and scaled clusters (for units $i = 0,1$ separately). (g) The standard deviation of the original differences on conjunctive trials vs the factor-dependent standard deviation of differences predicted by factor analysis for the criteria fulfilling differenced and scaled clusters.}
\label{fig:fig6}
\end{figure}

\subsection*{Differencing}

To test whether a shared state could underlie covariation in the remaining non-stationary angled clusters, the non-stationary single unit cluster spike times were differenced and rescaled (a standard time-series analysis technique) to remove neural adaptive trends by subtracting the spike time on trial t by the spike time on trial t-1.
% Clusters which did not fulfil the criteria after differencing were removed from further analysis. 
% Differencing a time series reduces the number of samples by 1 (Methods \ref{methods:angle_and_factor_analysis_criteria}). 
The green angled cluster of Fig. 3a fulfils the criteria after the spike times of its non-stationary unit are differenced (Fig. 6a). Application of the angle estimation technique estimated that the angle was X\todo[color=red!40]{Fill} plus minus Y (Fig. 6a), similar to the original angle and suggestive of an underlying factor after the removal of trial dependence. As expected by differencing, the spike times of the differenced unit show negative lag-1 autocorrelations and negative lag-2 partial autocorrelations (Fig. 6a). 
% Such negative autocorrelations are expected by differencing, and are removed in the next section by fitting ARIMA models.

Due to the expected negative autocorrelations introduced by differencing (Supplementary Fig. 9), the criteria for cluster unit autocorrelation p-values < 0.05, was relaxed for the differenced clusters (criterion 6, Methods \ref{methods:angle_and_factor_analysis_criteria}).
\ExecuteMetaData[AnalysisOutputLatex.tex]{SelectivelyDifferencedTestsPassedNewSummary}

\ExecuteMetaData[AnalysisOutputLatex.tex]{SelectivelyDifferencedTestsPassedEigenvalueSummary}
Again, the factor analysis angles show close similarity to the PCA angles (Fig. 6e).
The standard deviations of the factor-dependent response distributions (estimated by factor analysis) again offer a strikingly higher potential for encoding than simpler neural coding assumptions (Fig. 6a; Supplementary Vid 1; Fig. 6f,g; Supplementary Fig. 10).

% Negative lag-1 autocorrelations were found above chance levels but at chance for lags 2-12 (Supplementary Fig. X). Negative partial autocorrelations were found for lags 1-4 above chance levels, and at chance levels for lags 5-12 (Supplementary Fig. X). Such negative autocorrelations are expected by differencing. Positive autocorrelations and partial autocorrelations were found at or below chance for lags 1-12 (Supplementary Fig. X). It was inappropriate to assess cross correlations, which could be spuriously due to autocorrelations.

% \ExecuteMetaData[AnalysisOutputLatex.tex]{SelectivelyDifferencedSummary}

% Differencing removed the correlations of both units' spike times with trial index, as expected by selective differencing criterion 1 (Fig 6b). Selective difference spike times of both units are stationary according to the KPSS test, as expected by selective differencing criterion 2 (Fig 6c).

% Some Shapiro-Wilks (Fig. 6f). Some Henze-Zeckler normal (Fig. 6g). Most pass spherecity test (Supplementary Fig. X)

% \ExecuteMetaData[AnalysisOutputLatex.tex]{SelectivelyDifferencedCorrelatedSummary}
% \ExecuteMetaData[AnalysisOutputLatex.tex]{SelectivelyDifferencedTestsPassedSummary}

\subsubsection*{ARIMA modelling}

\begin{figure}[t!]
\centering
\includegraphics[width=\textwidth]{Figure7.pdf}
\caption{(a) Order of AR and MA models fit to the single unit cluster spike times of differenced clusters. (b) Predicted angles and confidence intervals of the criteria fulfilling ARIMA clusters. (c) Original PCA angles vs ARIMA PCA angles for the criteria fulfilling ARIMA clusters. (d) Linear regression $R^2$ values for criteria fulfilling ARIMA clusters. (e) PCA angles vs factor analysis $\theta_{45}$ angles for the criteria fulfilling ARIMA clusters. (f) The standard deviations of the original flat clusters vs standard deviations of the factor-dependent response distributions predicted by factor analysis for the criteria fulfilling ARIMA clusters (for units $i = 0,1$ separately). (g) The standard deviation of the original differences on conjunctive trials vs the factor-dependent standard deviation of differences predicted by factor analysis for the criteria fulfilling ARIMA clusters.}
\label{fig:universe}
\end{figure}

To remove autocorrelations in cluster unit time series (particularly negative autocorrelations introduced by differencing) and to return the differenced time series to the original scale without division by 2, autoregressive (AR) and/or moving average (MA) models were fit to differenced correlated criteria fulfilling angled clusters, using an automated version of the Box-Jenkins method (Methods \ref{methods:automated_box_jenkins}).

\ExecuteMetaData[AnalysisOutputLatex.tex]{SelectivelyDifferencedBoxJenkinsDifferencedSummary}
As expected, the model residuals showed autocorrelations, partial autocorrelations, cross correlations and partial cross correlations at and below chance levels (Supplementary Fig. 11). 
% Overall\todo{What type of clusters? Check.}, a very low amount of variance was explained by the linear models fit to predict spike pair times from trial number (Fig. 7e).

% The following number of models were fit to single unit cluster time series: MA(1): X/Y, MA(2): X/Y, MA(3): X/Y, AR(1): X/Y, AR(2): X/Y, AR(3): X/Y. 
% Differencing and ARMA removed the correlations of both units' spike times with trial index, as expected by selective differencing criterion 1 (Fig 7c). Selective difference spike times of both units are stationary according to the KPSS test, as expected by selective differencing criterion 2 (Fig 7d). 
% Shapiro (Supplementary Fig. X). Some Henze-Zeckler normal (Fig. 7f). Most pass spherecity test (Supplementary Fig. X)

\ExecuteMetaData[AnalysisOutputLatex.tex]{SelectivelyDifferencedBoxJenkinsCorrelatedSummary}
Fig. 7d shows the amount of variance explained by the pairwise correlation to the criteria fulfilling ARIMA cluster.
Fig. [ToDo] shows the difference between scaled differenced and ARIMA estimated cluster angles, showing that the scaled difference angles gave a good estimation of the angle.

\ExecuteMetaData[AnalysisOutputLatex.tex]{SelectivelyDifferencedBoxJenkinsTestsPassedEigenvalueSummary} 
The angles estimated for the factor analysis components were similar to the angles estimated by PCA (Fig. 7e). The standard deviations of the factor-dependent response distributions (estimated by factor analysis) again offer a strikingly higher potential for encoding than simpler neural coding assumptions (purple ellipses in row 2 column 3 of Supplementary Vid 1; Fig. 7f,g; Supplementary Fig. X).

\subsection*{Stimuli, unit-types and unit-locations}

Fig. 8a shows the spatial distribution of unit pairs which exhibited criteria fulfilling correlated clusters, pooled for stationary clusters and non-stationary clusters made stationary by ARIMA modelling. Correlations are found, which span multiple (up to 4) cortical layers and barrels for both E and I types. The most correlations are found between E-type units in L5A and L5B of the principle column and neighbouring columns. Fig. 8b shows a histogram over stimulus frequencies of criteria fulfilling correlated clusters both for stationary clusters and non-stationary clusters made stationary by ARIMA modelling.

\begin{figure}[t!]
\centering
\includegraphics[width=\textwidth]{Figure8.pdf}
\caption{Spatial organisation and eliciting stimuli of criteria fulfilling correlated clusters. (a) Line colour and thickness indicates the number of criteria fulfilling correlated clusters found for pairs of units of different E/I-type and in different layers and barrels. Red and blue correspond to E and I unit types respectively. For example, the thick green line illustrates that there were 10 criteria fulfilling clusters, for which both units were in L5B, 1 unit was in the principle column and 1 was in a neighbouring column. The colour and thickness of rings surrounding the red or blue circles indicate the number of pairs of the same E/I type, layer and barrel. The left most column is shown for pairs in which units in a pair are on opposite sides of the principle column. Counts are pooled for stationary clusters and non-stationary clusters made stationary by ARIMA modelling. (b) Histogram over stimulus frequencies of criteria fulfilling correlated clusters both for stationary clusters and non-stationary clusters made stationary by ARIMA modelling.}
\label{fig:universe}
\end{figure}

% Fig. 8a,b,c show the distribution of pairwise locations for unique correlated secondary clusters for E/E, I/I and E/I unit pairs respectively. It is clear that such pairwise correlations occur across cortical layers and barrels, with X correlations occurring across 4 cortical barrels\todo[color=red!40]{Ellaborate}. This gives primary evidence that a shared state underlying the correlations may at least be shared across Xmm of cortex, making it possible that downstream neurons could process/interpret pairwise differences in a state-dependent manner. The majority of unique secondary correlated clusters occured for low-frequency stimulations (0-4Hz) (Fig. 8d). It is important to note that the heterogeneity in the experiments (i.e. number of trials per stimulus, stimulation frequencies, stimulated whiskers, barrels recorded from, heterogeneity in type and selectivity of units detected) makes it difficult to make full quantitative conclusions as to the spatial distribution of correlated clusters and the stimuli which they are elicited by.

\section{Discussion}

\subsection*{Summary of results}
[TO DO]

\subsection*{Underlying state}
[TO DO]

% For example,  The local field potential (LFP) reflects changes in the subthreshold membrane potential [REF] and is strongly correlatated across cortical layers [REF], suggesting that a subthreshold membrane potential state could be shared across layers.

\subsection*{The feature-binding problem (FBP) and space-time continua}
[TO DO]
% The \textit{presence} of a sensory feature (i.e. pattern) may signal the \textit{opportunity} to undertake particular \textit{types} of actions, whilst feature \textit{properties} (i.e. transformation \textit{motor-specifics} of performing such actions. Over stages of sensory hierarchies, cell assemblies (i.e. groups of neurons) respond to increasingly complex features \cite{hubel1962receptive, hinkle2002three, yamane2008neural} with increased selectivity over larger ranges of transformations \cite{rust2010selectivity}. A \textit{`feature-presence cell assembly'} defines a group of neurons which responds selectively to a feature over multiple transformations. The feature-binding problem (FBP) originally asked how, in the presence of multiple high-level features, feature-properties were associated with the correct feature-presence cell assemblies \cite{Rosenblatt_Principles_1961, Treisman_Illusory_1982} (Fig. 1e). Historically proposed solutions failed to solve the FBP (see Discussion).

% Stimuli are likely to correspond with response distributions that exist on a continua in response distribution space. That is, small changes in stimulus space are likely to correspond with small changes in response distribution space. The relevance of such `space-time continua' to behaviour, experience and generalisation are later discussed. It follows from above that space-time continua would be non-stationary and factor-dependent. Consideration of candidate neural codes should additionally account for (i) how such continua emerge through learning, (ii) how spatially overlapping stimulus representations could superimpose (\cite{Rosenblatt_Principles_1961, von1999and}; Fig. 1X), and (iii) how representations can be correctly parsed with minimal information loss.

% Characterisation of single stimulus-response distributions which account for non-stationarity and factor-dependence are a necessary step towards characterising the nature of space-time continua in the brain.

% It also follows that a postsynaptic neuron's action potential timings \textit{could} vary continuously with continuous (spatial and temporal) input variation, such that cell assemblies forming spike-time continuums \textit{may} elicit downstream spike-time continuums. Spike-time-dependent plasticity (STDP) \cite{markram1997regulation} suggests spike-timing continuums could self-organise.

% The FBP is simplified by evidence that cell assemblies can encode feature properties through their activity \textit{pattern} \cite{gollisch2008rapid, storchi2012comparison, chang2017code} (Fig. X). 
% % Feature presence and property information could then be encoded jointly, 
% obviating posthoc association of action-opportunities and specifics. The definitive nature of property encoding remains open, as is the question of how the brain would learn to parse input into such representations. Candidate neural codes are considered w.r.t. the following proposed requirements of a feature-property encoding form:

% \textit{(i) Discriminable encoding} as discussed.

% \textit{(ii) Continua encoding.} For dimensions corresponding to the candidate neural code, small (continuous) changes in feature-transform should correspond with small (continuous) changes to stimulus response-distributions (Fig. 1f).

% \textit{(iii) Learned continua.}
% If continua arise through experience such that some regions on a continua represent transforms experienced during learning, the region representing a novel transform on the continua would implicitly encode the transform \textit{relative} to previously experienced feature transforms (Fig. 1f).

% % Analysis of space-time continua is beyond the scope of the current paper but it is interesting to note that if such continua are non-stationary, then the encoding of previously experienced stimuli, and new stimuli relative to such previously encoded stimuli, would also be non-stationary.\color{black}

% % This would describe semantic meaning and subjective experience as comparison to past experience, and has useful properties for generalisable action (Fig. X)\todo[color=green!40]{Same figure?}. The discussion posits how feature presence-encoding cell assemblies with property encoding continuums could emerge for a novel feature from experience of only a few example transformations.

% \textit{(iv) Parsing and selectivity.} Correct activation of feature-presence cell assemblies is constrained by the representation of low and high-level features, and the mechanisms by which downstream neurons learn and integrate information. 
% % Fig. 1e illustrates how spike-timing based continua could support such selectivity using neural coincidence detection.

% \textit{(v) Superposition.} During co-activation of multiple disjoint or overlapping feature-presence assemblies, the occurrence of downstream areas `interpreting' incorrect features/feature-transforms should be minimised (\cite{Rosenblatt_Principles_1961, von1999and}; Fig. 1g\todo[color=red!40]{Add to figure.}).
% % Several firing rate competition schemes have been presented for simultaneously active overlapping representations [REF]. A recent NHP study recording from face-specific and body-specific neurons in IT showed that... \cite{bao2018representation}.

% \textit{(vi) Compositionality.}
% If property-encoding forms were general for low and high-level representations, such representations could be \textit{compositional}. Particularly, low-level property information could be considered during parsing and passed on in the output representation of the higher-level representation.

% % Candidate neural codes would ideally be considered in a trial-conditioned manner. 
% % for example by making characterisations for inferred trial-conditioned stimulus-response distribution projections rather than for non-trial-conditioned distributions.

% % Spike times may also be measured relative to oscillatory activity including population spiking activity or subthreshold membrane oscillations, which themselves could indirectly encode the onset of stimuli, internal event or temporal reference points. Here we present evidence of such possible codes.

\subsection*{Non-stationarity}

Future studies of non-stationary time-warped continua should carefully consider non-stationarity. Adaptation describes the `continuous recalibration of the sensory system to compensate for the changes in the statistics of the input stream' \cite{kheradpezhouh2017response}, with specific functional roles including gain modulation \cite{ahissar2000transformation, heiss2008shift, khatri2009stimulus, mohar2013opposite}, efficiency \cite{sharpee2006adaptive} and detectability of novel stimuli \cite{dragoi2002dynamics, ulanovsky2003processing}. The rate and degree of adaptation depends on stimulus parameters, neuron location and neuron properties (see \cite{kheradpezhouh2017response} for review). Adaptation modulates correlations and response heterogeneity (see \cite{kheradpezhouh2017response} for review), and depends on brain state and anaesthesia \cite{castro2004absence, katz2012trial}. Adaptation from repetitive stimulation changes excitatory-inhibitory balance \cite{higley2006balanced, heiss2008shift, malina2013imbalance} thought to be caused by short term synaptic depression (due to neurotransmitter depletion) observed in trigeminal complex \cite{mohar2013opposite}, thalamus \cite{deschenes2003relay} and cortex \cite{higley2006balanced, heiss2008shift, malina2013imbalance} (mostly at thalamocortical synapses \cite{chung2002short}).

In mouse barrel cortex, spiking statistics change through development \cite{van2017layer}. Functional imaging in adult mice shows day-to-day maintenance of population activity \cite{margolis2012reorganization, mayrhofer2015sparse} but variability in the day-to-day stability of single unit representations \cite{mayrhofer2015sparse}. Functional imaging shows that in addition to tuning sensory maps, plasticity and sensory experience affects response gain, variability, firing correlations, and top-down modulation by task context (see \cite{lemessurier2018plasticity}).

Mechanisms of attention \cite{cohen2010neuronal} and plasticity \cite{lemessurier2018plasticity}.
% Collectively, these studies characterise the barrel cortex as a highly dynamic system, with representations continuously changing.

% There are several obvious possibilities for the form of such an underlying state with possible explanations of how they could cause variability and how downstream neurons could interpret relative spike patterns differentially w.r.t. such underlying states:

% determines integration velocity, and thus secondary spatiotemporal pattern is warped in a state space too.

% High trial-to-trial variability and the curse of dimensionality suggests that characterisation of stimulus-response distribution structure may be difficult for large neuronal groups (Section X).

% Under each of these encoding forms,

% \cite{williams2020discovering} Neuron paper
% \begin{itemize}
%     \item Nicely sets up. Discussion of variability. Idea of averaging over trials etc.
%     \item Like phrase `ground truth variability'
%     \item Cool that doesn't require reference to event
%     \item Does it take into account neural adaptation. Some neurons change over the course of an experiment. Whilst others don't and are still time-warped
%     \item Could look to see if there is some order in the warping functions (i.e. slopes) over trials. Could all be down to adaptation! (Although probably not).

%     \item `place a penalty on the area between each warping function and the identity line'. Could be that some are getting warped (or adapted) and some are not, and identity is a `half-way' house.

%     \item Anything to stop different conditions getting warped to same PSTH? Perhaps a bit useless then? Could have tested decoding of reaches?

%     \item Could you actually fit different models for different neurons using the one-held out model? `This approach fits the neural response templates to a subset of trials(Figure 2J, shaded blue regions) and fits the warping functions to a subset of the neurons (Figure 2J, shaded red regions). We refer to the intersection of these two sets as the training set (Figure 2J,shaded purple regions), while the intersection of the complement sets is used for validation and testing (Figure 2J, unshaded' regions). This procedure is then repeated many times withdifferent randomized partitions of the data.

%     \item `Unlike in trial-averaged analyses,R2valueswill typically be very low because spike trains exhibit substantialsingle-trial variability'
% \end{itemize}

% Not sure I understand:
% \begin{itemize}
%     \item `roughness penalty on the response template, which directly encourages the model to identify smooth firing rates'
% \end{itemize}

% A recent optogenetics study showed that mice could discriminate visual stimuli with 76\% accuracy with V1 activity limited to 80ms, during which \textasciitilde85\% of `discriminating units' spiked 0 or 1 times \cite{resulaj2018first}, demonstrating an initial stimulus-representation of single spikes.

% \cite{rolls1995responses,zoccolan2005multiple}.

% Firing rates in visual experiments are often calculated over periods of hundreds of milliseconds, leading to the common misconception of sustained consistent responses.

% Another study showed that the encoding mechanism of two simultaneously presented features? in V1 was dependent on the relative contrast between the two features \cite{}. [Desimone?].

% Gamma, other? \cite{fries2007gamma, havenith2011synchrony, vinck2010gamma, vinck2013attentional}
% \textit{Rank order coding} suggests information is encoded in the order in which neurons spike [REF]. This has a high combinatorial capacity but does not alone address the requirements. 
% \textit{Precise interspike intervals} have been observed but would require longer integration timescales than patterns of single spikes.

% Capsule networks

% \textit{Feature Integration Theory} \cite{Treisman_Illusory_1982} suggested that features within an attenional locus are associated to a single object, contradicting multiple object perception. \textit{Binding-by-synchrony} (BBS) \cite{milner1974model, von1981correlation} suggested feature-property representations are associated by firing in synchrony. The limited number of high-level features that distinct phases of proposed gamma range synchrony would allow discords with our detailed perception. BBS was supported by local synchronous MUA pair oscillations \cite{gray1989oscillatory, brecht1998correlation} but requires distant localised coordination. \textit{Communication through coherence (CTC)} \cite{fries2005mechanism, fries2007gamma, fries2015rhythms} suggests periodic synchronization focuses excitability to short windows \cite{fries2015rhythms}, allowing competing presynaptic groups to establish effective connectivity \cite{womelsdorf2007modulation}. A mechanism for modulation of selective coherence (supported experimentally \cite{bosman2012attentional, fries2015rhythms}) was recently proposed \cite{fries2015rhythms}. The \textit{Thousand Brains} hypothesis does not account for multiple objects and contradicts sensory maps by suggesting that objects are encoded by 1000s of grid cell systems encoding object location w.r.t. sensor patches \cite{hawkins2019}.

% \textit{Population codes} \cite{hawkins2017theory} do not address the requirements but solve the \textit{`the combination problem'} \cite{von1999and}: that a near infinite number of cells would be needed to encode every possible feature-presence/property combination individually.

% Relative spike patterns can be detected at different timescales without reference to events \cite{russo2017cell}.

% The occurrence of positively correlated clusters (Sec. 3.5) are suggestive of a factor (or factors) affecting first spike times of multiple neurons monotonically. Non-unitary covariation suggests that spike timing relationship w.r.t. the factor is more complex than either consistent alignment to oscillatory activity plus independent noise or variation in the precise onset of the stimulus (which would vary pairs of first spike times with unitary gradient). Instead, non-unitary gradients are suggestive of shared underlying factors (possibly just one) that vary the relative timing differences between neurons. The following subsections aim to test the degree to which trial number and/or an unknown hidden factor underlie such non-unitary covariation.

% Representation of transform relative to previous experience has several useful properties for generalisable action (Fig. 1e) and could describe semantic meaning and subjective experience as comparison to past experience.

% \subsection{Working notes on points still to address}

% Maybe in this paper, or in future:
% \begin{enumerate}
%     \item Are other clusters definitely from other units?
%         \begin{itemize}
%             \item Should consider that current work was interested only interested in finding and analysing correlated clusters, rather than detecting all definitive clusters. This would be easier to do with more trials per stimulus (as clusters hopefully more defined) and can look at disjointness of trials.
%             \item Worth at least thinking about. 
%         \end{itemize}
%     \item 3+ unit correlations?
%     \item Does LFP explain clustering or within cluster variance?
%         \begin{itemize}
%             \item Clustering hard.
%         \end{itemize}
%     \item Does population and/or single unit spiking explain clustering and/or variance?
%         \begin{itemize}
%             \item Multi-trial history
%             \item Preceding activity.
%         \end{itemize}
% \end{enumerate}

% Maybe some analysis of secondary spikes as support from same unit. Think. Consider spike sorting.

% \todo[color=red!40]{Plan key points first}

\section{Materials and Methods}
\label{sec:headings}

\subsection{Spike sorting summary}
\label{methods:spike_sorting}

See original paper for details of spike sorting \cite{reyes2014laminar}.
Spikes were detected using groups of 2 to 4 channels defined as `virtual tetrodes' \cite{harris2000accuracy, einevoll2012towards}. Spike detection used amplitude-thresholding \cite{gray1995tetrodes}. Spike waveform features were used for sorting by KlustaKwik (http://klustakwik. sourceforge.net) and Klusters (http://klustakwik.sourceforge.net) \cite{harris2000accuracy, hazan2006klusters}. Unit quality was verified using several additional steps. Cells were classified as putative INH and EXC neurons based on the mean spike waveform \cite{sirota2008entrainment, sakata2009laminar, royer2012control}. Histology, VSD imaging and voltage density analysis were used to determine the three-dimensional location of electrodes so that each unit could be assigned to a cortical layer (L2/3, L4, L5A, L5B).

\subsection{Clustering Stage 1: Flat clusters}

% \subsubsection*{\textit{Stage 1 - DBSCAN Clustering.}}
% \subsubsection*{Stage 1 - Flat clusters}
\label{methods:flat_clusters}

\textit{1. DBSCAN.} The DBSCAN unsupervised clustering algorithm was used as a first step to detect clusters in  first spike pair response distributions.
% (dendrogram grouping produced similar results not shown here). 
The DBSCAN EPS parameter has the effect of varying cluster size. Clustering was performed using the following DBSCAN EPS values sequentially: [0.4, 0.45, ..., 5.0]. The colour of points in Fig. 3a represents the clusters to which points were assigned by DBSCAN for EPS = 1.35. If the clusters produced for an EPS value were identical to those produced for the previous EPS value, then the clusters produced for the current EPS value were not included in further analysis. First-stage clusters with less than 20 elements were also not included in further analysis. Checks for similar or identical \textit{individual} clusters produced for different EPS values were not used at this stage to avoid sampling bias, which could affect the uniformity of surrogate control tests. Checks for similar clusters are introduced in the next stage. The DBSCAN `minimum number of elements in a cluster' parameter was set to 2 for reasons given in the next step.

\textit{2. Isolation test.} For each DBSCAN cluster produced in step 1, an `intersection ellipse' was created (dashed ellipses in Fig. 3a), with centre equal to the cluster mean and widths equal to $w_i = 3*max(0.5, \sigma_i)$ for units $i=0,1$, where $\sigma_i$ is the standard deviation of samples in the cluster for unit i (dashed ellipses in Fig. 3a). As a control for cluster quality, any initial cluster with an intersection ellipse which intersected with or was within at least one other intersection ellipse were excluded from further analysis (i.e. excluded from the set of DBSCAN clusters). In the example of Fig. 3a, this corresponded to the blue, red and light blue dashed clusters. Setting the DBSCAN `minimum number of elements in a cluster' parameter of 2 increased the cluster quality control, as intersection ellipses were created for 2 element clusters rather than assigning them as outliers.

\textit{3. Flat clusters.} For each remaining DBSCAN cluster, a flat bounding ellipse was created (solid green ellipse in Fig. 3a) with centre equal to the DBSCAN cluster mean, and widths equal to $w_i = 4*\sigma_i$ for units $i=0,1$, where $\sigma_i$ is the standard deviation of samples in the DBSCAN cluster for unit i. All elements within the bounding ellipse (including those not assigned to the original DBSCAN cluster) were assigned to the `flat cluster'.

\subsection{Clustering Stage 2: Angled clusters}
\label{methods:angled_clusters}

\textit{4. Angled clusters.} `Correlated flat clusters' refer to `flat clusters' from the previous stage with correlation p-values < 0.005. An angled ellipse was estimated for points in each correlated flat cluster using a 3-stage bootstrap procedure. Bootstrap stage 1 used elements in the correlated flat cluster. A number of elements equal to the number of elements in the flat cluster were sampled from the flat cluster with replacement. PCA was performed on the sample to attain a sample mean and covariance matrix. This constituted a single bootstrap \textit{`iteration'}. 1000 such iterations were made.

The sample means and covariance matrices were then averaged, from the 1000 sample means and covariance matrices, to construct a new bounding ellipse (similar to the rotated green ellipse in Fig. 3a) with centre equal to the bootstrapped mean, rotation equal to the angle that the principal component made with the positive x-axis, and widths equal to $4\sigma_i$ where $\sigma_i$ is the square root of the variance explained by each principle component. The bootstrap procedure was then repeated by sampling the elements within this new ellipse (bootstrap stage 2), with the averaged mean and covariance matrix used to construct a third ellipse for sampling in bootstrap stage 3. Elements within the final bounding ellipse defined by the averaged mean and covariance matrix of the third bootstrap stage were assigned to the `angled cluster'.

\textit{5. Isolation test.} The step tested whether scaled versions of the angled ellipses from the step above ($3\sigma_i$ instead of $4\sigma_i$) were well isolated (similarly to the Stage 1). If a scaled angled ellipse intersected with or was within the flat bounding ellipses of other DBSCAN clusters, it was removed from further analysis.

\textit{6. Cluster similarity check.} Angled clusters which shared at least 15 elements to previous angled clusters (produced by lower EPS values for the same response distribution) were excluded from further analysis. Combined with the ascending order of the EPS sequence, this meant that no clusters were included, which contained well isolated `sub-clusters' produced by lower EPS values.
% with similar mean points (difference norm < 0.1) and similar covariance matrices (difference norm < 0.25) to previous angled clusters

\subsection{Correlation surrogate control datasets}
\label{methods:correlation_surrogate_control_datasets}

To test that correlations were not artifacts of the clustering algorithm and were above chance levels, the algorithm was applied to the following surrogate control datasets in which correlated clusters appeared at chance levels:

\textit{(i) All shuffled.} Pairwise shuffling of elements of first spike pair distribution.

\textit{(ii) Single Gaussian.} Spike pairs drawn from an independent bivariate normal distribution with the same mean and standard deviation as the original spikes. The number of samples drawn from the distribution was equal to the number of original samples.

% \textit{(iii) Correlated clusters shuffled.} Pairwise shuffling of elements of correlated secondary clusters.

% \textit{(iv) Flattened Gaussian clusters.} Elements of correlated secondary clusters were replaced by an equal number of samples from an independent bivariate normal distribution with the same mean and standard deviations as the cluster.

\subsection{Cluster angle confidence interval}
\label{methods:cluster_angle_confidence_interval}

Response stochasticity and resultant cluster boundary stochasticity, meant that applying the angle and confidence interval algorithm to the means and covariance matrices of the first bootstrap stage was not accurate in estimating the true cluster angle. The 1000 sample first principle components produced in bootstrap stage 3 were used to find the average angle that the first principle component makes with the positive x-axis (with confidence intervals). To account for symmetry, sample principal components with negative y-components were replaced by their anti-correlated vector so that all vectors had angles with the positive x-axis between 0 and $180^{\circ}$. To allow averaging and angular difference calculations without the effects of discontinuity, the vectors were rotated about the origin such that the angles of the vectors with the x-axis were between 0 and $360^{\circ}$ (double their original angle). The resulting mean vector was then rotated by half of its anti-clockwise angle with the positive x-axis so that it was between 0 and $180^{\circ}$. The standard deviation of the angle between the mean first principle component vector and sample first principal component vectors (in $360^{\circ}$ space) were calculated before dividing the value by 2 to create a value at the correct scale. This standard deviation was then used to create angle confidence intervals for the first principle component.

\subsection{Angle surrogate control dataset}
\label{methods:angle_surrogate_control_dataset}

To validate that such differences from $45^{\circ}$ were not artifacts of the algorithm, a surrogate control dataset was created by rotating each positively correlated angled clusters that had an angle significantly different from $45^{\circ}$, such that the principal component of the cluster was $45^{\circ}$. The algorithm was rereun on this control dataset. Clusters were removed, however, if the number of cluster samples was 5 greater than the original rotated cluster. This controlled for instances in which a cluster had been rotated close to another cluster of spikes. This made sure the surrogate control dataset did not contain clusters different from $45^{\circ}$. X\todo[color=red!40]{Fill} positively correlated angled clusters were detected for the surrogate dataset (compared to Y for the original dataset) suggesting robustness in secondary correlated cluster detection. Only X\% of the correlated secondary clusters were significantly different from $45^{\circ}$ for p-value < 0.05 (below chance) (Supplementary Fig. 6). 
% X\% of p-values were actually greater than 0.95 demonstrating similarity to $45^{\circ}$.

\subsection{Criteria cluster angle and factor analysis}
\label{methods:angle_and_factor_analysis_criteria}

Clusters were determined to be stationary and sufficiently correlated such that an underlying factor could underlie the covariation if they passed the following 5 tests: (1) LR p-value < 0.05, r-value > 0.3, (2) Bartlett spherecity p-value > 0.05, (3) multi-output $r^2 < 0.05$, (4) both units trial index LR p-value > 0.05, (5) both units KPSS p-value > 0.05. (6) Both units lag-1 autocorrelation p-value > 0.05 (this requirement was relaxed for the differenced clusters but not ARIMA clusters). Bartlett spherecity tests that the correlation matrix is significantly different from the identity, and is a standard pre-test for factor analysis. The Kaiser-Meyer-Olkin (KMO) factor analysis pre-test for factor analysis was unnecessary, as there were only two units, the partial correlation matrix and the correlation matrix are the same.

\subsection{Differencing}
\label{methods:differencing}

Differencing a time series reduces the number of samples by 1.
For clusters for which the spike times of only one unit was differenced, the first sample of the undifferenced unit was removed, so that the number of samples was equal for both units. In this case a spike time for an undifferenced trial t was matched with the difference between the spike times on trial t and trial t - 1.

Differencing increases the standard deviation of a time series $S = (s_1, ..., s_t)$ by approximately 2. This is shown by considering differencing as approximate to negating two independent random variables $S_{(1, t-1)} = (s_1, ..., s_{t-1})$ and $S_{(2, t)} = (s_2, ..., s_t)$, where $t$ is the number of samples in the original time series. As the standard deviation of $S_{(0, t-1)}$ and $S_{(1, t)}$ are approximate to the standard deviation of the original series $S$, the standard deviation of the differences is on the order of \(\sigma_{differenced} \approx \sqrt{\sigma_{(1, t-1)}^2 + \sigma_{(2, t)}^2} \approx  \sqrt{2\sigma^2} \).  Differenced time series were therefore divided by 2, which later comparison to ARIMA models shows to be reasonable. The assessment of `raw' differences, rather than the ARIMA residuals, however, allows a direct assessment of cluster angles after trend removal.

\subsection{Automated Box-Jenkins Method}
\label{methods:automated_box_jenkins}

Working on improving.

\section{Author Contributions}

% JBI wrote the initial draft and analyses.

% JBI developed the theories, analyses, literature review and ...
% VRP conducted the original rat barrel cortex data collection,

\section{Funding}

JBI’s contribution to this work was supported by the Oxford Foundation for Theoretical Neuroscience and Artificial Intelligence, the UK Economic and Social Science Research Council (ESRC) (grant no. ES/J500112/1) and the UK Engineering and Physical Science Research Council (EPSRC) (grant no. EP/N509711/1).

[To do]

\section{Acknowledgements}

% Arxiv
% \bibliographystyle{unsrt}  
% \bibliography{references}

% Nature
\printbibliography

% \textbf{Cortical onset}

% Average cortical onset was determined as the time that the PSTH summed over all stimuli, units and experimental sessions rose above 3 standard deviations of PSTH before stimulus onset. This was determined as 6ms. Spike trains were considered in a 20ms window following this time point, at which point the average PSTH had returned to

% Due to the variable number of trials for different stimuli and experimental sessions (Supplementary material x), and the sparseness neuronal responses

% \section{Draft cover letter}

% Dear Editor,

% \begin{itemize}
%     \item Paper aims to target a wide audience. With particular focus in neuroscience, artificial intelligence and mathemeatics.
%     \item Where mathematical notation was unneccesary for the description of concepts, we have chosen to description instead. In order not to obfusticate principles and widen reader audience.
% \end{itemize}

% \section{Questions for readers}

% \begin{enumerate}
%     \item Better to use `neuron group' or `cell assembly'.
%     \item Missed any Feature-binding types?
%     \item Missed any neural coding types?
%     \item Was it clear?
% \end{enumerate}

\end{document}

### 2020-07-22 manuscript.tex
Banging!

@@
 \section{Introduction}
-The neural coding problem asks how information is encoded through patterns of action potentials in the nervous system \cite{gerstner1997neural, johnson2000neural, stein2005neuronal, gerstner2014neuronal}.
-Responses following sensory stimulation can be analysed to explore principles of neural coding, with a classic dichotomy between temporal \cite{uzzell2004precision, johansson2004first, gollisch2008rapid, storchi2012comparison, reinagel2002precise, reyes2014laminar} and rate codes \cite{hong2016explicit, chang2017code, kar2019evidence, stringer2019high}.
-Temporal codes provide the possibility of a fast and efficient code.
+The neural coding problem asks how information is encoded through patterns of action potentials in the nervous system \cite{gerstner1997neural, johnson2000neural, stein2005neuronal, gerstner2014neuronal}, with a dichotomy between temporal \cite{uzzell2004precision, johansson2004first, gollisch2008rapid, storchi2012comparison, reinagel2002precise, reyes2014laminar} and rate codes \cite{hong2016explicit, chang2017code, kar2019evidence, stringer2019high}.
+% Temporal codes provide the possibility of a fast and efficient code. 
 Whilst precise temporal patterns have been observed \cite{uzzell2004precision, johansson2004first, gollisch2008rapid, storchi2012comparison, reinagel2002precise, reyes2014laminar}, a large proportion of trial-to-trial variability remains unexplained.
-This high variability leads to common assumption of population firing rate codes, under the premise that variability cancels out across a population \cite{abbott1999effect, moreno2014information, stringer2019high}.
-Average firing rate codes are abstracted from neural mechanisms (see below), leaving a full understanding of neural coding and variability undetermined.
+Such variability leads to the common assumption of average firing rate codes, under the premise that variability cancels out across a population \cite{abbott1999effect, moreno2014information, stringer2019high}.
+Continuous firing rates, averaged over time, trial or population, have capacity beyond the single trial discrete spike count which they abstract, leaving a full understanding of neural coding and variability undetermined.
+It is important to better characterise neural codes in order to understand a rich variety of dynamic, structural, homeostatic and cognitive processes, as well as properties of action potential patterns directly involved in behaviour.
+% Characterising responses may also uncover response properties involved in behaviour.
 
-It is important to better characterise neural codes in order to understand a rich variety of dynamical, structural, homeostatic and cognitive processes.
-Characterising responses may also uncover response properties involved in behaviour.
-Due to response variability from trial-to-trial, characterising the neural representation of a single stimulus is a matter of characterising the \textit{distribution} of possible neural responses.
-
-A multi-neuron stimulus response can be expressed as a point in multi-dimensional space, where each dimension represents a spike train metric.
-Responses from multiple presentations of a stimulus form a sample stimulus-response distribution in this space, which approximates an underlying stimulus-response distribution (Fig.
+Responses following sensory stimulation can be analysed to explore principles of neural coding.
+Due to response variability, characterising the neural representation of a single stimulus is a matter of characterising the \textit{distribution} of possible neural responses for the stimulus.
+A multi-neuron response can be expressed as a point in multi-dimensional space, where each dimension represents a spike train metric.
+Responses from multiple presentations of a stimulus form a sample stimulus-response distribution in this space (Fig.
 1a).
-If two stimuli have a behaviourally relevant difference, downstream neurons must be able to respond differently to samples from the response distribution of either stimulus.
-This is possible if the response distributions are highly non-overlapping in dimensions that downstream neurons are sensitive to, such as relative spike time differences (\cite{gasparini2006state, branco2010dendritic, branco2011synaptic}; Fig.
-1a).
-
-Two phenomena complicate stimulus response distribution characterisation.
-Firstly, responses of some neurons depend on trial index.
-That is, response distributions are \textit{non-stationary} due to mechanisms of adaptation \cite{dragoi2002dynamics, ulanovsky2003processing, sharpee2006adaptive, ahissar2000transformation, kheradpezhouh2017response, ahissar2001temporal, khatri2009stimulus, martin2014tactile, barros2019response}.
-It follows that the region from which stimulus-responses are drawn from at a single point in time is smaller than the region from which samples are drawn from over multiple trials (Fig.
-1b).
-Adaptation occurs across species and modalities \cite{dragoi2002dynamics, ulanovsky2003processing, sharpee2006adaptive}, and in barrel cortex to repetitive whisker stimulation for both spike count \cite{ahissar2000transformation, ahissar2001temporal, kheradpezhouh2017response, khatri2009stimulus, martin2014tactile, barros2019response} and spike latency \cite{ahissar2000transformation, ahissar2001temporal, kheradpezhouh2017response}.
+% which approximates an underlying stimulus-response distribution.
 
@@
 
-Secondly, firing correlations between spatially distanced neurons (see \cite{reid2012functional, feldmeyer2013barrel} for review) suggest the possible presence of a \textit{continuous} underlying state covarying neural responses.
-Response samples on a single trial (when the continuous state has a certain value) would then come from a state-dependent stimulus-response distribution (i.e.
-a smaller region of response distribution space; Fig.
+If two stimuli have a behaviourally relevant difference, downstream neurons must be able to respond differently to samples from the underlying response distribution of either stimulus.
+This is possible if the response distributions are highly non-overlapping in dimensions that downstream neurons are sensitive to, such as relative spike time differences (\cite{gasparini2006state, branco2010dendritic, branco2011synaptic}; Fig.
+1a).
+Characterising the distribution from which a stimulus response is `drawn' from on a single trial is complicated by state dependence, neural adaptation and heterogeneity of sample response distribution structure.
+
+Firing correlations between spatially distanced neurons (see \cite{reid2012functional, feldmeyer2013barrel}) suggest the possible presence of a \textit{continuous} underlying state covarying neural responses.
+Response samples on a single trial, when the continuous state has a certain value, would then come from a state-conditioned stimulus-response distribution.
+This would be a smaller region of stimulus-response space than the region from which all samples come (Fig.
 1c).
-If receiving neurons could respond to a state-modulated signal in a state-dependent manner (contrary to common neural coding assumptions \cite{moreno2014information, stringer2019high}), state-dependent encoding would increase the capacity for discriminable stimulus encoding (Fig.
-1c).
-The question is whether correlations between spatially distanced neurons are evidence of a shared state, which receiving neurons could also be modulated by [New Fig?].
-% Correlations in the barrel cortex up to a distance of $\sim1.5mm$ (vertically and horizontally) \cite{reyes2015long} suggest it is possible that afferently connected neurons could also be state-modulated and respond in a state-dependent manner.
-Subthreshold membrane potentials are a candidate for such shared state, as they affect action potential initiation \cite{petersen2003interaction} and are correlated across areas \cite{poulet2008internal}.
-% State information could alternatively be explicitly encoded in state modulated population activity.
+Therefore, if downstream neurons could respond in a state-dependent manner (contrary to common neural coding assumptions \cite{moreno2014information, stringer2019high}), state-dependent encoding would increase the capacity for discriminable stimulus encoding.
 
-Estimating state-dependent stimulus-response distributions at a single snapshot in time is made difficult by heterogeneity of non-stationarity and response distribution structures (Supplementary Vid.
-1), leading to use of averaging and coarse spike train metrics.
-As \textit{spike counts} show high trial-to-trial variability, single units cannot reliably encode many different stimuli.
-Instead, time, trial or population averaged scalar \textit{firing rates} \cite{gerstner2014neuronal} are calculated, with capacity beyond the original single trial discrete spike count, under the assumption that noise cancels out across a population \cite{abbott1999effect, moreno2014information, stringer2019high}.
-
-Trial-averaged firing rate population codes (often calculated over windows $>100ms$) allow decoding of object \cite{hong2016explicit} and face \cite{chang2017code} properties from macaque higher visual areas.
-% Such long integration windows are discordant with behavioural speed \cite{thorpe1996speed} and the fast propagation of activity through cortical layers and areas \cite{perrett1982visual}. 
-Object category can be decoded population trial averaged multi-unit activity (MUA) over short 10ms time windows \cite{kar2019evidence}.
-Decoders do not generalise to later windows \cite{kar2019evidence}, excluding the notion of sustained firing rate representations.
-Higher encoding capacity is conferred by high spike counts of unsorted MUA \cite{hong2016explicit, kar2019evidence} or for persisting visual stimuli \cite{hong2016explicit, chang2017code, kar2019evidence}.
-Dimensionality reduction is commonly used to find dynamical population firing rate patterns \cite{ahrens2012brain, mante2013context, cunningham2014dimensionality}.
-Spike count and firing rates do not superimpose (Fig 1d\todo[color=red!40]{To do}), although competition schemes have been presented for simultaneously active overlapping representations \cite{bao2018representation}.
-% Rate and timing codes may be aligned to internal oscillations \cite{fries2007gamma, havenith2011synchrony, vinck2010gamma, vinck2013attentional}.
+Characterisation of response distribution structure is also made complicated by neural adaptation, which occurs across species and modalities \cite{dragoi2002dynamics, ulanovsky2003processing, sharpee2006adaptive}, and in barrel cortex to repetitive whisker stimulation for spike count \cite{ahissar2000transformation, ahissar2001temporal, kheradpezhouh2017response, khatri2009stimulus, martin2014tactile, barros2019response} and spike latency \cite{ahissar2000transformation, ahissar2001temporal, kheradpezhouh2017response}.
+That is, the response of some neurons are \textit{non-stationary} and change as a function of trial index.
+It is therefore important to control for neural adaptation when searching for correlative structure indicative of a possible underlying state.
 
@@
 stimulus location, intensity, frequency) and neuronal activity.
-A recent optogenetics study showed that early visual cortex patterns of single spikes are sufficient to enable visual discrimination \cite{resulaj2018first}, implicating single spike patterns as a candidate for atomic representation.
+A recent optogenetics study showed that early visual cortex patterns of single spikes are sufficient to enable visual discrimination \cite{resulaj2018first}, implicating single spike patterns as a candidate for atomic representation of information.
 
@@
 Relative differences \cite{izhikevich2006polychronization} observed experimentally \cite{johansson2004first, gollisch2008rapid, reyes2014laminar} have been shown to encode information when trial averaged \cite{gollisch2008rapid} and do not require reference to an onset signal.
+Response distributions may form multiple clusters, implying stochasticity in the cluster from which representations come (Fig. 1d; \cite{reyes2014laminar}).
 % as neurons are sensitive to the spatial and temporal pattern of synaptic activations \cite{gasparini2006state, branco2010dendritic, branco2011synaptic}.
-Action potential instantaneity suggests that relative spike codes could superimpose (Fig. 1d).
-
-Assumptions about stimulus-response distribution structure constrain the potential encoding power of timing-based single spike patterns.
-Independent and normally distributed first spike-times would confer a low discriminability potential (as will be seen), dependent on single neuron variability (Fig. 1e). 
-% Over typical 10-40ms ranges of mean first spike times for short stimuli \cite{reyes2014laminar, resulaj2018first}, 
-% Observed first spike time standard deviations \cite{reyes2014laminar} would imply that very few unique stimuli/features could be encoded with high discriminability by a neuron pair. 
-Spike time differences between two neurons with independent and normal first spike times would have a standard deviation of \(\sigma_{diff} = \sqrt{\sigma_{a}^2 + \sigma_{b}^2}\).
-Spike times could covary with unitary gradient under the assumption that relative spike times have a factor-independent mean.
-Precise relative differences have been reported in the eye \cite{gollisch2008rapid} and barrel cortex \cite{reyes2014laminar}.
+Action potential instantaneity suggests that relative spike codes could superimpose (Fig.
+1d).
 Response distributions may form multiple clusters, implying stochasticity in the cluster from which representations come (Fig.
 1d; \cite{reyes2014laminar}).
-% , which could be determined by some underlying factor or non-stationarity.
 
-% Characterisation techniques for sparse (low trial count) datasets are often used, which make assumptions about distribution structure, risking fundamental distribution properties and encoding forms being overlooked.
+Using a large data set of spike sorted recordings in anesthetized rat barrel cortex, pairwise first spike response distributions were first observed for repeated presentations of single stimuli.
+That is for repeated presentations of a single stimulus, the first spike times of two units were plotted against each other for trials on which they both spiked.
+The reader is strongly recommended to view these plots in Supplementary Vid.
+1.
+Lines, colours and ellipses can be ignored at this stage.
+Simple observation of these plots shows that clusters of first spike pair responses appear correlated, suggesting that first spike pairs are correlated for pairs of neurons on a subset of trials of a stimulus.
 
-% The spike time of either neuron would be correlated with the difference \todo[color=green!40]{Figure?}(Fig.
-X) but due to independence there would be no process underlying this correlation, and thus no means by which downstream neurons could interpret the difference in a state-dependent manner (other than an event signal).
-Moreover, the discriminability of one neuron's spike time and the difference would be the same as that of the two neuron's spike times \todo[color=green!40]{Same figure}(Fig.
-X).
+Moreover, the correlation angle of some of the correlated clusters appears to be different from $45^{\circ}$ (i.e.
+has a gradient different from 1).
+Such correlations with `non-unitary gradient' may be indicative that correlations are due to an internal process (i.e.
+state) rather than variation in the precise timing of stimulus onset.
+However, this must be considered after checking and controlling for neural adaptive trends.
 
-% Projection of the stimulus-response distribution to a first spike time vs difference space would not decrease the area of the distribution, and thus would not improve discriminability.
+A novel algorithm extracted the observed correlated clusters and found that they appeared above chance levels.
+To control for adaptive trends, the cluster spike times of each neuron in the pair were tested for stationarity.
+The cluster spike times of both units were stationary for X\% of correlated clusters.
+For the remaining clusters, which had at least one unit with non-stationary cluster spike times, time series analysis was applied to the cluster spike times of the unit to remove the non stationarity (i.e.
+neural adaptive trend).
 
-% The standard deviation of spike times and spike time differences would not capture clustering structure, thus underestimating discriminability.
-If trial-wise cluster was determined by some state, then discriminability given a state, would equal the corresponding cluster's discriminability.
+% That is, each axis corresponds to the first spike time of a different recorded unit.
+The scatter points therefore represent the first spike times of both units for the trials (of a single stimulus) on which both units spiked.
 
-\section{Results}
+when the is spike times are correlate. Such first spike response distribution structure can be seen in the top left plots of Supplementary Vid. 1, The the reader is strongly recommended to view.
 
-\subsection*{Overview}
-% 
-This paper analyses first-spike stimulus-response distributions for pairs of barrel cortex units in response to repeated single whisker deflections.
-In first-spike stimulus-response distributions for pairs of units, correlated clusters were found above chance-levels.
-Clustering was necessary for the detection of a large proportion of correlations.
+Simple observation .  It is clear
+
+It is important for neurophysiology to determine whether correlations between spatially distanced neurons are evidence of a shared continuous state, and whether this shared state also modulates downstream decoding neurons.
+Subthreshold membrane potentials are a candidate for such state, as they affect action potential initiation \cite{petersen2003interaction} and are correlated across areas \cite{poulet2008internal}.
+
+From observation of first spi
+
+This paper presents a novel algorithm, which characterises the occurrence and nature of correlations between the first spike times of pairs of neurons on subsets of trials.
+This offers (1) a precise understanding of stimulus responses and correlative variability [Vicente Refs], and (2) as a necessary step towards either rejecting state-dependent coding, or confirming its precise nature.
+
+% It follows that the region of response-distribution space from which a stimulus-response is drawn from at a single point in time is smaller than the region from which samples are drawn from over all trials (Fig.
+1b).
+
+% The paper presents a novel technique for uncovering correlated first spike times for pairs of neurons in noisy response distributions.
+
+% As \textit{spike counts} show high trial-to-trial variability, single units cannot reliably encode many different stimuli, leading to the use of firing rate codes (as discussed above and in the the Discussion).
+
+% Assumptions about stimulus-response distribution structure constrain the potential encoding power of timing-based single spike patterns.
+% Independent and normally distributed first spike-times would confer a low discriminability potential (as will be seen), dependent on single neuron variability (Fig. 1e). 
+% % Over typical 10-40ms ranges of mean first spike times for short stimuli \cite{reyes2014laminar, resulaj2018first}, 
+% % Observed first spike time standard deviations \cite{reyes2014laminar} would imply that very few unique stimuli/features could be encoded with high discriminability by a neuron pair. 
+% Spike time differences between two neurons with independent and normal first spike times would have a standard deviation of \(\sigma_{diff} = \sqrt{\sigma_{0}^2 + \sigma_{1}^2}\), where \(\sigma_{i}\) is the standard deviation of neuron $i$ ($i=0,1$).
+% Spike times could covary with unitary gradient under the assumption that relative spike times have a factor-independent mean. 
+% Precise relative differences have been reported in the eye \cite{gollisch2008rapid} and barrel cortex \cite{reyes2014laminar}.
+
+The paper uncovers interesting correlative properties, which offer a novel description of neural representations and are suggestive (but not confirmatory) of an internal continuous state varying responses from trial-to-trial.
+
+The characterisation offers a prediction of the encoding resolution of this state-dependent coding form, which would surpass previous unconditioned coding forms.
+Testing the decoding performance of the uncovered response properties would require explicit knowledge of the underlying state so is left for future studies.
+
+% Using a large data set of spike sorted recordings in anesthetized rat barrel cortex, the paper aims to address the following questions:
+
+\textit{1.} To repeated repeated trials of single stimuli (i.e. single whisker deflections at a particular velocity) do the first spike times of pairs of units form clear clusters?
+
+\textit{2.}
+
+first spike responses of two units
+
+Firstly, for repeated trials of single stimuli (i.e.
+single whisker deflections at a particular velocity), are the first spike times of pairs of units correlated?
+This analysis should take into account outliers and the possibility that responses come .
+
+In first-spike stimulus-response distributions for pairs of units, correlated clusters were found above chance-levels. Clustering was necessary for the detection of a large proportion of correlations.
+
+\textit{2.} For sets of correlated spike pairs, are the angles of correlation significantly different from $45^{\circ}$ (i.e.
+is the gradient of correlation different from 1)?
+This question is of interest for two reasons.
+Firstly such .
+This leads to the
+
+\textit{3.} Does this offer a potential improvement in encoding capacity
+
+\textit{4.} For pairs of units in which
+
+analyses first-spike stimulus-response distributions for pairs of barrel cortex units in response to repeated single whisker deflections. The
+
+we addressed the following questions / hypotheses. (1) …. (2) …. (3)…
+
+To our knowledge this is the first report demonstrating that
+
 Cluster spike times of both units were stationary for X\%\todo[color=red!40]{Fill} of correlated clusters.
@@
 Rather, neurons should process relative spike patterns in a factor-dependent manner.
+
+% , which could be determined by some underlying factor or non-stationarity.
+
+% Characterisation techniques for sparse (low trial count) datasets are often used, which make assumptions about distribution structure, risking fundamental distribution properties and encoding forms being overlooked.
+
+% Projection of the stimulus-response distribution to a first spike time vs difference space would not decrease the area of the distribution, and thus would not improve discriminability.
+
+% The standard deviation of spike times and spike time differences would not capture clustering structure, thus underestimating discriminability.
+If trial-wise cluster was determined by some state, then discriminability given a state, would equal the corresponding cluster's discriminability.
+
+\section{Results}
 
@@
 \ExecuteMetaData[AnalysisOutputLatex.tex]{ClusterCorrelationSummary2}
+
+% The spike time of either neuron would be correlated with the difference \todo[color=green!40]{Figure?}(Fig.
+X) but due to independence there would be no process underlying this correlation, and thus no means by which downstream neurons could interpret the difference in a state-dependent manner (other than an event signal).
+Moreover, the discriminability of one neuron's spike time and the difference would be the same as that of the two neuron's spike times \todo[color=green!40]{Same figure}(Fig.
+X).
 
@@
 [TO DO]
+
+Trial-averaged firing rate population codes (often calculated over windows $>100ms$) allow decoding of object \cite{hong2016explicit} and face \cite{chang2017code} properties from macaque higher visual areas.
+% Such long integration windows are discordant with behavioural speed \cite{thorpe1996speed} and the fast propagation of activity through cortical layers and areas \cite{perrett1982visual}. 
+Object category can be decoded population trial averaged multi-unit activity (MUA) over short 10ms time windows \cite{kar2019evidence}.
+Decoders do not generalise to later windows \cite{kar2019evidence}, excluding the notion of sustained firing rate representations.
+Higher encoding capacity is conferred by high spike counts of unsorted MUA \cite{hong2016explicit, kar2019evidence} or for persisting visual stimuli \cite{hong2016explicit, chang2017code, kar2019evidence}.
+Dimensionality reduction is commonly used to find informative dynamical population firing rate patterns \cite{ahrens2012brain, mante2013context, cunningham2014dimensionality}.
+Spike count and firing rates do not superimpose (Fig 1d\todo[color=red!40]{To do}), although competition schemes have been presented for simultaneously active overlapping representations \cite{bao2018representation}.
+% Rate and timing codes may be aligned to internal oscillations \cite{fries2007gamma, havenith2011synchrony, vinck2010gamma, vinck2013attentional}.
+
 % The \textit{presence} of a sensory feature (i.e.

### 2020-07-23 manuscript.tex
Gut

@@
 % Temporal codes provide the possibility of a fast and efficient code. 
-Whilst precise temporal patterns have been observed \cite{uzzell2004precision, johansson2004first, gollisch2008rapid, storchi2012comparison, reinagel2002precise, reyes2014laminar}, a large proportion of trial-to-trial variability remains unexplained.
+Whilst precise temporal patterns have been observed \cite{uzzell2004precision, johansson2004first, gollisch2008rapid, storchi2012comparison, reinagel2002precise, reyes2014laminar}, a large proportion of neural variability remains unexplained.
 Such variability leads to the common assumption of average firing rate codes, under the premise that variability cancels out across a population \cite{abbott1999effect, moreno2014information, stringer2019high}.
-Continuous firing rates, averaged over time, trial or population, have capacity beyond the single trial discrete spike count which they abstract, leaving a full understanding of neural coding and variability undetermined.
+Continuous firing rates (averaged over time, trial or population \cite{gerstner1997neural}) have capacity beyond the single trial discrete spike count from which they are abstracted, leaving a full understanding of neural coding and variability undetermined.
 It is important to better characterise neural codes in order to understand a rich variety of dynamic, structural, homeostatic and cognitive processes, as well as properties of action potential patterns directly involved in behaviour.
@@
 
-If two stimuli have a behaviourally relevant difference, downstream neurons must be able to respond differently to samples from the underlying response distribution of either stimulus.
+If two stimuli have a behaviourally relevant difference, downstream neurons must be able to respond differently to samples from the response distribution of either stimulus.
 This is possible if the response distributions are highly non-overlapping in dimensions that downstream neurons are sensitive to, such as relative spike time differences (\cite{gasparini2006state, branco2010dendritic, branco2011synaptic}; Fig.
@@
 
-Firing correlations between spatially distanced neurons (see \cite{reid2012functional, feldmeyer2013barrel}) suggest the possible presence of a \textit{continuous} underlying state covarying neural responses.
+Firstly, correlative structure between the responses of spatially distanced neurons (see \cite{reid2012functional, feldmeyer2013barrel}) in a single stimulus response distribution, opens the possibility of a \textit{continuous} underlying state, which covaries neural responses from trial-to-trial.
 Response samples on a single trial, when the continuous state has a certain value, would then come from a state-conditioned stimulus-response distribution.
@@
 
-Characterisation of response distribution structure is also made complicated by neural adaptation, which occurs across species and modalities \cite{dragoi2002dynamics, ulanovsky2003processing, sharpee2006adaptive}, and in barrel cortex to repetitive whisker stimulation for spike count \cite{ahissar2000transformation, ahissar2001temporal, kheradpezhouh2017response, khatri2009stimulus, martin2014tactile, barros2019response} and spike latency \cite{ahissar2000transformation, ahissar2001temporal, kheradpezhouh2017response}.
-That is, the response of some neurons are \textit{non-stationary} and change as a function of trial index.
-It is therefore important to control for neural adaptation when searching for correlative structure indicative of a possible underlying state.
+Characterisation of response distribution structure is also complicated by neural adaptation, which occurs across species and modalities \cite{dragoi2002dynamics, ulanovsky2003processing, sharpee2006adaptive}, and in barrel cortex to repetitive whisker stimulation for spike count \cite{ahissar2000transformation, ahissar2001temporal, kheradpezhouh2017response, khatri2009stimulus, martin2014tactile, barros2019response} and spike latency \cite{ahissar2000transformation, ahissar2001temporal, kheradpezhouh2017response}.
+That is, the response distributions are \textit{non-stationary} and change as a function of trial index for some neurons.
+Adaptation could therefore lead to correlative structure in single stimulus response distributions.
+Adaptation must therefore be controlled for when searching for correlative structure indicative of a possible underlying state.
 
@@
 The majority of barrel cortex neurons respond 0 or 1 times to single whisker deflections \cite{reyes2014laminar}.
-First spike patterns allow direct study of the relationship between known entities (e.g.
-stimulus location, intensity, frequency) and neuronal activity.
+Such first spike patterns allow direct study of the relationship between known stimuli and neuronal activity.
 A recent optogenetics study showed that early visual cortex patterns of single spikes are sufficient to enable visual discrimination \cite{resulaj2018first}, implicating single spike patterns as a candidate for atomic representation of information.
 
-First spike times can have millisecond order precision relative to stimulus onset in sensory organs \cite{johansson2004first, uzzell2004precision, gollisch2008rapid}, thalamic areas \cite{reinagel2002precise, storchi2012comparison} and early sensory cortices \cite{reyes2014laminar}, with some evidence of continuum-like encoding relative to onset \cite{gollisch2008rapid, storchi2012comparison} and oscillations \cite{havenith2011synchrony}.
+First spike times with millisecond order precision have been observed relative to stimulus onset in sensory organs \cite{johansson2004first, uzzell2004precision, gollisch2008rapid}, thalamic areas \cite{reinagel2002precise, storchi2012comparison} and early sensory cortices \cite{reyes2014laminar}, as well as relative to oscillations \cite{havenith2011synchrony}.
 Relative differences \cite{izhikevich2006polychronization} observed experimentally \cite{johansson2004first, gollisch2008rapid, reyes2014laminar} have been shown to encode information when trial averaged \cite{gollisch2008rapid} and do not require reference to an onset signal.
-Response distributions may form multiple clusters, implying stochasticity in the cluster from which representations come (Fig. 1d; \cite{reyes2014laminar}).
+Response distributions may form multiple clusters, implying stochasticity in the cluster from which representations come on a single trial (Fig. 1d; \cite{reyes2014laminar}).
 % as neurons are sensitive to the spatial and temporal pattern of synaptic activations \cite{gasparini2006state, branco2010dendritic, branco2011synaptic}.
-Action potential instantaneity suggests that relative spike codes could superimpose (Fig.
-1d).
-Response distributions may form multiple clusters, implying stochasticity in the cluster from which representations come (Fig.
-1d; \cite{reyes2014laminar}).
-
-Using a large data set of spike sorted recordings in anesthetized rat barrel cortex, pairwise first spike response distributions were first observed for repeated presentations of single stimuli.
-That is for repeated presentations of a single stimulus, the first spike times of two units were plotted against each other for trials on which they both spiked.
-The reader is strongly recommended to view these plots in Supplementary Vid.
+
+Using a large data set of spike sorted recordings in anesthetized rat barrel cortex, first spike response distributions were first observed for single stimuli and pairs of units.
+That is, the first spike times of two units were plotted against each other for trials on which they both spiked in response to a particular stimulus.
+The reader is strongly recommended to view these plots in the top left of Supplementary Vid.
 1.
-Lines, colours and ellipses can be ignored at this stage.
-Simple observation of these plots shows that clusters of first spike pair responses appear correlated, suggesting that first spike pairs are correlated for pairs of neurons on a subset of trials of a stimulus.
-
+Simple observation of these plots (Ignoring lines, colours and ellipses at this stage) shows that clusters of first spike pair responses appear correlated, suggesting that first spike pairs may correlated for pairs of neurons on a subset of trials of a stimulus.
 Moreover, the correlation angle of some of the correlated clusters appears to be different from $45^{\circ}$ (i.e.
 has a gradient different from 1).
-Such correlations with `non-unitary gradient' may be indicative that correlations are due to an internal process (i.e.
-state) rather than variation in the precise timing of stimulus onset.
-However, this must be considered after checking and controlling for neural adaptive trends.
-
-A novel algorithm extracted the observed correlated clusters and found that they appeared above chance levels.
+
+% That is, each axis corresponds to the first spike time of a different recorded unit.
+The scatter points therefore represent the first spike times of both units for the trials (of a single stimulus) on which both units spiked.
+
+A simple shift in the precise timing of stimulus onset or the onset of cortical activity, would produce correlations with a unitary gradient (i.e.
+a $45^{\circ}$ angle).
+The observed correlations with `non-unitary gradient' may suggest that correlations are due to an internal process, for example, a shared cortical state underlying state-dependent processing.
+Such correlations with `non-unitary gradient' must be considered after checking and controlling for neural adaptive trends.
+
+A novel algorithm extracted the observed clusters and found that they were correlated appeared above chance levels.
 To control for adaptive trends, the cluster spike times of each neuron in the pair were tested for stationarity.
 The cluster spike times of both units were stationary for X\% of correlated clusters.
-For the remaining clusters, which had at least one unit with non-stationary cluster spike times, time series analysis was applied to the cluster spike times of the unit to remove the non stationarity (i.e.
+For the remaining clusters, which had at least one unit with non-stationary cluster spike times, time series analysis was applied to the cluster spike times of the non-stationary unit/units to remove the non-stationarities (i.e.
 neural adaptive trend).
-
-% That is, each axis corresponds to the first spike time of a different recorded unit.
-The scatter points therefore represent the first spike times of both units for the trials (of a single stimulus) on which both units spiked.
-
-when the is spike times are correlate. Such first spike response distribution structure can be seen in the top left plots of Supplementary Vid. 1, The the reader is strongly recommended to view.
-
-Simple observation .  It is clear
-
-It is important for neurophysiology to determine whether correlations between spatially distanced neurons are evidence of a shared continuous state, and whether this shared state also modulates downstream decoding neurons.
+X\% of the original clusters
+
+The characterisation offers a prediction of the encoding resolution of the possible state-dependent coding form, which would surpass previous unconditioned coding forms and controls for neural adaptive trends.
+It also provides a precise understanding of stimulus responses and correlative variability, which can be interpreted in the classic sense [Vicente Refs].
+
+The precise characterisation of correlative structure in this paper is a necessary component step towards either rejecting state-dependent coding, or confirming its precise nature.
+It is important for neurophysiology to determine whether correlative structure between spatially distanced neurons is caused by continuous state, which also modulates downstream decoding neurons.
+Testing the state-dependent decoding performance of the uncovered response properties would require explicit knowledge of the underlying state so is left for future studies.
 Subthreshold membrane potentials are a candidate for such state, as they affect action potential initiation \cite{petersen2003interaction} and are correlated across areas \cite{poulet2008internal}.
-
-From observation of first spi
-
-This paper presents a novel algorithm, which characterises the occurrence and nature of correlations between the first spike times of pairs of neurons on subsets of trials.
-This offers (1) a precise understanding of stimulus responses and correlative variability [Vicente Refs], and (2) as a necessary step towards either rejecting state-dependent coding, or confirming its precise nature.
 
@@
 
-The characterisation offers a prediction of the encoding resolution of this state-dependent coding form, which would surpass previous unconditioned coding forms.
-Testing the decoding performance of the uncovered response properties would require explicit knowledge of the underlying state so is left for future studies.
-
 % Using a large data set of spike sorted recordings in anesthetized rat barrel cortex, the paper aims to address the following questions:
@@
 \textit{1.} To repeated repeated trials of single stimuli (i.e. single whisker deflections at a particular velocity) do the first spike times of pairs of units form clear clusters?
-
-\textit{2.}
-
-first spike responses of two units
 
@@
 \textit{4.} For pairs of units in which
-
-analyses first-spike stimulus-response distributions for pairs of barrel cortex units in response to repeated single whisker deflections. The
 
@@
 
+Action potential instantaneity suggests that relative spike codes could superimpose (Fig. 1d).
+
+some evidence of continuum-like encoding relative to onset \cite{gollisch2008rapid, storchi2012comparison} and oscillations \cite{havenith2011synchrony}.
+
 Trial-averaged firing rate population codes (often calculated over windows $>100ms$) allow decoding of object \cite{hong2016explicit} and face \cite{chang2017code} properties from macaque higher visual areas.

### 2020-07-23 manuscript.tex
Gut

@@
 (b) Illustration of trial dependence of a sampled response distribution (dots). Circle illustrates the boundary of a possible trial-dependent response distribution on a single trial. 
-(c) Illustration of boundary of factor-dependent first spike response distribution for a single factor value.
-The green line shows the mean points that would be predicted for a continuous set of factor values.
+(c) Illustration of boundary of state-dependent first spike response distribution for a single state value.
+The green line shows the mean points that would be predicted for a continuous set of state values.
 (d) Illustration of superposition of two single spike pattern feature representations.
@@
 The orange ellipses illustrate the boundary of an underlying response distribution under the assumption of independence and normality for the sample (solid) and another stimulus (dashed).
-Centre: Illustration of spike time differences for values of the factor.
-The green circle illustrates the boundary of a factor-dependent distribution of spike time differences.
+Centre: Illustration of spike time differences for values of the state.
+The green circle illustrates the boundary of a state-dependent distribution of spike time differences.
 Right: Conceptual illustration of time warp relative differences.}
@@
 The topographic representation of each single whisker on the snout of a rodent in the somatosensory cortex, the so-called barrel cortex, represents an excellent model to study neuronal processing of defined sensory stimulation \cite{feldmeyer2013barrel}.
-The majority of barrel cortex neurons respond 0 or 1 times to single whisker deflections \cite{reyes2014laminar}.
-Such first spike patterns allow direct study of the relationship between known stimuli and neuronal activity.
-A recent optogenetics study showed that early visual cortex patterns of single spikes are sufficient to enable visual discrimination \cite{resulaj2018first}, implicating single spike patterns as a candidate for atomic representation of information.
+The shortness of whisker deflections reduces ambiguity in what temporal aspect of a stimulus neural activity represents, offering controlled study of the neural representation of atomic entities.
+The majority of barrel cortex neurons respond 0 or 1 times to single whisker deflections \cite{reyes2014laminar}, implicating single spike patterns as a candidate for atomic representation of information.
+Representation by single spike patterns is supported in other sensory modalities, such as vision, where a recent optogenetics study showed that early visual cortex patterns of single spikes are sufficient to enable visual discrimination \cite{resulaj2018first}.
 
@@
 Using a large data set of spike sorted recordings in anesthetized rat barrel cortex, first spike response distributions were first observed for single stimuli and pairs of units.
-That is, the first spike times of two units were plotted against each other for trials on which they both spiked in response to a particular stimulus.
+That is, for repeated trials of a single stimulus, the first spike times of two units were plotted against each other for trials on which they both spiked.
 The reader is strongly recommended to view these plots in the top left of Supplementary Vid.
 1.
-Simple observation of these plots (Ignoring lines, colours and ellipses at this stage) shows that clusters of first spike pair responses appear correlated, suggesting that first spike pairs may correlated for pairs of neurons on a subset of trials of a stimulus.
-Moreover, the correlation angle of some of the correlated clusters appears to be different from $45^{\circ}$ (i.e.
-has a gradient different from 1).
+Simple observation of these plots (ignoring lines, colours and ellipses at this stage) shows that clusters of first spike pair responses appear correlated, suggesting that first spike pairs are correlated for pairs of neurons on a subset of trials of a stimulus.
 
@@
 
+Moreover, the correlation angle of some of the correlated clusters appears to be different from $45^{\circ}$ (i.e. has a gradient different from 1).
 A simple shift in the precise timing of stimulus onset or the onset of cortical activity, would produce correlations with a unitary gradient (i.e.
@@
 
-A novel algorithm extracted the observed clusters and found that they were correlated appeared above chance levels.
+A novel algorithm extracted the observed clusters and found that they were correlated above chance levels.
 To control for adaptive trends, the cluster spike times of each neuron in the pair were tested for stationarity.
 The cluster spike times of both units were stationary for X\% of correlated clusters.
-For the remaining clusters, which had at least one unit with non-stationary cluster spike times, time series analysis was applied to the cluster spike times of the non-stationary unit/units to remove the non-stationarities (i.e.
-neural adaptive trend).
-X\% of the original clusters
-
-The characterisation offers a prediction of the encoding resolution of the possible state-dependent coding form, which would surpass previous unconditioned coding forms and controls for neural adaptive trends.
-It also provides a precise understanding of stimulus responses and correlative variability, which can be interpreted in the classic sense [Vicente Refs].
+X\% of these stationary correlated clusters had non-unitary (different from $45^{\circ}$) correlation angles, suggestive of an internal underlying state co-varying first spike times and thus \textit{warping} the distribution of (decodable) relative spike time differences from trial-to-trial (Fig.
+1d).
+
+For the remaining clusters, with at least one unit with non-stationary cluster spike times, time series analysis was applied to remove the neural adaptive trends of the non-stationary unit/units.
+X\% of the detrended clusters were correlated with non-unitary gradient, suggestive of an underlying state.
+
+Application of state analysis to the stationary and detrended clusters provided an estimate of possible state-dependent response distributions.
+This gave a prediction of the resolution of possible state-dependent encoding, which would surpass previously proposed coding forms and controls for neural adaptive trends.
+It also provides a precise understanding of stimulus responses and correlative variability, which could be interpreted classically (i.e.
+independently of state [Vicente Refs]).
 
@@
 It is important for neurophysiology to determine whether correlative structure between spatially distanced neurons is caused by continuous state, which also modulates downstream decoding neurons.
-Testing the state-dependent decoding performance of the uncovered response properties would require explicit knowledge of the underlying state so is left for future studies.
+Testing the state-dependent decoding performance of the response properties uncovered in this paper would require explicit knowledge (or an estimate) of the underlying trial-to-trial state so is left for future studies.
 Subthreshold membrane potentials are a candidate for such state, as they affect action potential initiation \cite{petersen2003interaction} and are correlated across areas \cite{poulet2008internal}.
+
+The characterised response distribution structure is referred to as `stationary and non-stationary clustered time-warped spike timing'.
+This differs from previous `time-warp' studies, which aim to align spike trains over trials and do not consider discriminable encoding, clustering, non-stationarity and different warping functions for different neurons \cite{williams2020discovering}.
+One previous study has modelled time-warp-invariant integration \cite{gutig2009time}, although neurons need not respond identically to warps of the same pattern.
+Rather, neurons should process relative spike patterns in a state-dependent manner.
+
+The results suggest that consideration of clustering, state-dependence and non-stationarity, together, are necessary to explain a large proportion of neural variability, and could explain the failure to find a fast and efficient ubiquitous spike code that utilises the high sensitivity of neural architecture and functional dynamics.
+In this study, such consideration gave a novel prediction of the resolution of neural stimulus representation.
+
+% To our knowledge this is the first report demonstrating that
+
+% Algorithms for detecting correlated clusters and estimating correlation angles were developed for this study and verified on surrogate control datasets.
 
@@
 
-The paper uncovers interesting correlative properties, which offer a novel description of neural representations and are suggestive (but not confirmatory) of an internal continuous state varying responses from trial-to-trial.
+% The paper uncovers interesting correlative properties, which offer a novel description of neural representations and are suggestive (but not confirmatory) of an internal continuous state varying responses from trial-to-trial.
 
@@
 
-\textit{1.} To repeated repeated trials of single stimuli (i.e. single whisker deflections at a particular velocity) do the first spike times of pairs of units form clear clusters?
-
-Firstly, for repeated trials of single stimuli (i.e.
+% \textit{1.} To repeated repeated trials of single stimuli (i.e. single whisker deflections at a particular velocity) do the first spike times of pairs of units form clear clusters?
+
+% Firstly, for repeated trials of single stimuli (i.e.
 single whisker deflections at a particular velocity), are the first spike times of pairs of units correlated?
@@
 
-In first-spike stimulus-response distributions for pairs of units, correlated clusters were found above chance-levels. Clustering was necessary for the detection of a large proportion of correlations.
-
-\textit{2.} For sets of correlated spike pairs, are the angles of correlation significantly different from $45^{\circ}$ (i.e.
+% In first-spike stimulus-response distributions for pairs of units, correlated clusters were found above chance-levels.
+Clustering was necessary for the detection of a large proportion of correlations.
+
+% \textit{2.} For sets of correlated spike pairs, are the angles of correlation significantly different from $45^{\circ}$ (i.e.
 is the gradient of correlation different from 1)?
@@
 
-\textit{3.} Does this offer a potential improvement in encoding capacity
-
-\textit{4.} For pairs of units in which
-
-we addressed the following questions / hypotheses. (1) …. (2) …. (3)…
-
-To our knowledge this is the first report demonstrating that
-
-Cluster spike times of both units were stationary for X\%\todo[color=red!40]{Fill} of correlated clusters.
-X\% of these stationary correlated clusters had non-unitary (different from $45^{\circ}$) correlation angles, suggestive of an internal underlying factor co-varying first spike times and thus \textit{warping} the distribution of (decodable) relative spike time differences from trial-to-trial (Fig.
-1d).
-Estimated factor-conditioned response distributions for these stationary clusters showed a high potential encoding capacity.
-% Algorithms for detecting correlated clusters and estimating correlation angles were developed for this study and verified on surrogate control datasets.
-
-Time series analysis was applied to the non-stationary units of the remaining correlated clusters, to remove neural adaptive trends.
-A large proportion of detrended clusters were correlated with non-unitary gradient, suggestive of an underlying factor with an improved encoding potential.
-The results suggest that consideration of clustering, factor-dependence and non-stationarity, together, are necessary to explain a large proportion of neural variability, and could explain the failure to find a fast and efficient ubiquitous spike code that utilises the high sensitivity of neural architecture and plasticity.
-In this study, such consideration gave an estimate of the potential resolution of neural stimulus representation.
+% \textit{3.} Does this offer a potential improvement in encoding capacity
+
+% \textit{4.} For pairs of units in which
+
+% we addressed the following questions / hypotheses. (1) …. (2) …. (3)…
+
 % Although the study analyses responses to single stimuli in isolation, it also considers the proposed and previously suggested neural codes with respect to feature-binding and continua encoding.
-
-The characterised response distribution structure is referred to as \textbf{stationary and non stationary clustered time-warped spike timing}.
-This differs from previous `time-warp' studies, which aim to align spike trains over trials and do not consider discriminable encoding, clustering, non-stationarity and different warping functions for different neurons \cite{williams2020discovering}.
-One previous study has modelled time-warp-invariant integration \cite{gutig2009time}, although neurons needn't respond identically to warps of the same pattern.
-Rather, neurons should process relative spike patterns in a factor-dependent manner.
 
@@
 
-The dataset used for this paper is reused from an earlier barrel cortex study, of which analyses and data collection methods have previously been reported \cite{reyes2014laminar}.
+The dataset used for this paper is reused from an earlier study, of which analyses and data collection methods have previously been reported \cite{reyes2014laminar}.
 In summary, extracellular recordings were made in 15 anaesthetised Wistar rats using 16 or 128 electrode silicon probes, inserted perpendicularly into the barrel cortex.
@@
 Experimental sessions varied in terms of electrode placement, stimulated whiskers, stimulation frequencies and number of trials per unique stimulus.
-7 experimental sessions used single shanks with 16 electrodes, whilst 8 experimental sessions used 8 neighbouring shanks (16 electrodes per shank) covering 2-3 neighbouring barrels.
-Spike sorting classified units by E/I type and cortical layers L2/3, L4, L5A, L5B (see Methods \ref{methods:spike_sorting}).
+7 experimental sessions used single shanks with 16 electrodes, whilst 8 experimental sessions used 8 neighbouring shanks (16 electrodes per shank) covering 2-4 barrel-related columns (median 4).
+Units were classified as excitatory or inhibitory and assigned to a cortical layers L2/3, L4, L5A and L5B/6 (see Methods \ref{methods:spike_sorting}).
+Spiking activity is analysed following cortical activation onset, determined as 5.2ms following stimulus onset (Supplementary Fig.
+1).
 
@@
 
-The shortness of whisker deflections reduces ambiguity in what temporal aspect of a stimulus neural activity represents, offering relatively controlled study of the neural representation of atomic entities.
-Excluding non-spiking responses over 25 and 100ms following cortical onset (5.2ms following stimulus onset; Supplementary Fig.
-1), responses to single whisker deflections in barrel cortex are largely single spike responses (Fig.
+This section justifies the analysis of first spike patterns as a cortical representation of single whisker deflections.
+Firstly, this section demonstrates that the majority of neurons respond to single whisker deflections with either 0 or 1 spike (Fig.
 2a).
@@
 2b).
-It is possible that double spike responses are examples of superpositions of single spike representations.
 The responses of each unit-type and layer group (averaged over stimuli) demonstrate a clear initial response, followed by a tailing-off (Fig.
@@
 \subsection*{First spike non-stationarity}
+
+This section demonstrates that the responses of some neurons exhibit depressive latency adaption, which should be controlled for when testing for correlative structure caused by a non-adaptive state.
 
@@
 
+It is possible that double spike responses are examples of superpositions of single spike representations, which would increase the encoding capacity further by allowing co-activity of overlapping cell assemblies.
+
 Action potential instantaneity suggests that relative spike codes could superimpose (Fig. 1d).

### 2020-07-24 manuscript.tex
Smashing

@@
 
-\keywords{Neural coding, precise spike timing, time-warp, feature-binding, barrel cortex, noise correlations, neural adaptation, in vivo recording}
+\keywords{Neural coding, precise spike timing, time-warp, state-dependent processing, barrel cortex, noise correlations, neural adaptation, in vivo recording, feature-binding}
 
@@
 
+Information theoretic approaches aim to understand 
 If two stimuli have a behaviourally relevant difference, downstream neurons must be able to respond differently to samples from the response distribution of either stimulus.
@@
 1a).
-Characterising the distribution from which a stimulus response is `drawn' from on a single trial is complicated by state dependence, neural adaptation and heterogeneity of sample response distribution structure.
+Characterising the distribution from which a stimulus response is `drawn' from on a single trial is complicated by correlative structure, neural adaptation and heterogeneity of sample response distribution structure.
+
+Correlative structure between the responses of different neurons in single stimulus response distributions indicates the presence of noise shared between the responses of neurons.
+These are referred to as `noise correlations' and are the focus of this manuscript, rather than `signal correlations' which refer to tuning similarity between neurons over multiple stimuli.
+
+Information theoretic approaches classically consider neural encoding with respect to fixed-state encoding and decoding.
+These approaches often do not consider that correlative structure (i.e.
+covariance of multi-neuron responses over trials of a single stimulus) could be caused by a \textit{continuous} underlying state, which covaries neural responses from trial-to-trial.
 
@@
 
-\subsection*{Low spike counts}
+\subsection*{Single spike patterns}
+
+\todo{Add PSTH of single stimulus. i.e. 1Hz over 10 and all stimuli maybe}
+\todo{Do conclusion}
 
@@
 
-\subsection*{First spike non-stationarity}
+\subsection*{Spike latency adaptation}
+
+\todo{Look at adaption by neuron type and layer}
+\todo{Select final adaption examples, with layers and stats.}
+\todo{Do adaption conclusion}
+\todo{Make adaption stats overview paragraph simpler, with nice single bar chart}
 
 This section demonstrates that the responses of some neurons exhibit depressive latency adaption, which should be controlled for when testing for correlative structure caused by a non-adaptive state.
-
-Fig 2d.
-shows the first spike times of 5 single units in response to repeated presentations of single stimuli.
-The unit of Fig.
-2d(i) demonstrates depressive latency adaptation.
-The units of Fig.
-2d(ii, iii) demonstrate no obvious latency adaptation.
-The units of Fig.
+Figure 2d shows the first spike times of 5 single units in response to repeated stimulations.
+The L23E unit of Fig.
+2d(i) demonstrates an irregular response pattern with adaptation.
+Whereas the L4I unit of Fig.
+2d(ii) shows a short and consistent response, the L… unit of Fig.
+2d(iii) reveals a sparse and variable pattern and no obvious adaptation.
+The L4I and L4I units of Fig.
 2d(iv, v) demonstrate more complex response-distribution structure and possible non-linear non-stationarity.
-For repeated presentations single stimuli, single unit first spike times are linearly correlated with trial number above chance
-(Fisher's method: $\chi^{2} (1786, N = 893) = 3672.69, p<.001$; Fig. 2e).
+
+Single unit first spike times are linearly correlated with trial number above chance (Fisher's method: $\chi^{2} (1786, N = 893) = 3672.69, p<.001$; Fig. 2e).
 % (Fisher's method: $\chi^{2} (1786, N = 893) = 3672.69, p = 1.027\textrm{e-132}$; Fig. 2e). 
@@
 2f), supporting depressive latency adaptation \cite{ahissar2000transformation, ahissar2001temporal, kheradpezhouh2017response}.
-The KPSS null hypothesis of stationarity was rejected for p-value < 0.05 for 19\% (168/893) single stimulus single unit first spike response distributions (Fig.
+The KPSS null hypothesis of stationarity was rejected for p-value < 0.05 for 19\% (168/893) first spike response distributions (Fig.
 2g).
-
 Autocorrelations of single-unit single-stimulus first spike times were above chance for lags 1-5 (Supplementary Fig.
-2), supportive of general adaptive trends.
+2), furthering support of general adaptive trends.
 Only the lag-2 partial autocorrelation was slightly above chance (Supplementary Fig.
-2), suggestive of some additional structure in the trial-to-trial sequence of first spike times for some units.
+2), suggestive of additional structure in the trial-to-trial sequence of first spike times for a small number of units, beyond adaptive trends.
 % after accounting for general adaptive trends.
 
-\subsection*{First spike pair response distribution example}
+\subsection*{Example first spike pair response distribution}
 
@@
 \caption{Clustered time-warp example and correlated clusters.
-(a) Sampled first spike response distribution for two units (L5A Inh, L4 Inh) over a 50ms window following cortical onset (1 outlier out of view).
+(a) Sampled first spike response distribution for two units (L5AI, L4I) over a 50ms window following cortical onset (1 outlier out of view).
 Point colours illustrate DBSCAN cluster (EPS = 1.35).
@@
 
+\todo{Do introduction (objectives) and summary}
+\todo{Clear up final paragraph}
+\todo{Differences standard deviation}
+
+This section presents an example first spike pair response distribution for the purposes of understanding.
 Fig.
-3a shows the distribution of first spike pairs in response to a 4Hz stimulus for two units (L5A Inh, L4 Inh) over a 50ms window following cortical onset.
-Only responses on `conjunctive trials' (trials on which both units spiked) are shown.
-The orange Gaussian is centered at the mean and has widths equal to 4 times the first spike time standard deviations of each unit (on conjunctive trial), illustrating the region from which 99.9\% of first spike pairs would be predicted to fall under the assumption of independence and normality.
-The assumptions fail to capture the distribution structure of many spike pair distributions (Supplementary Vid.
-1) and would allow very few stimuli to be encoded with high-discriminability.
-The spike time differences have a standard deviation of X (Fig.
-3b), which is also poor for discriminable encoding.
+3a shows the distribution of first spike pairs in response to a 4Hz stimulation of a single whisker for two units in L5AI and L4I of the corresponding principle column (over a 50ms window following cortical activation onset).
+Only responses on `conjunctive trials' (trials in which both units spiked for the single stimulus) are shown.
+
+The orange Gaussian is centered at the mean of the first spike pairs and has widths equal to 4 times the first spike time standard deviations of each unit (on conjunctive trial), illustrating the region from which 99.9\% of first spike pairs would be predicted to fall under the assumption of independence and normality (i.e.
+spike time jitter).
+This assumption fails to capture the distribution structure of this and many other spike pair distributions (Supplementary Vid.
+1) and very few different stimuli to be encoded with high-discriminability.
+
+The spike time differences have a standard deviation of X (Fig. 3b), which is also poor for discriminable encoding.
 
 First spike pairs in Fig.
-3a appear to form two clusters (blue and green dots).
+3a appear to be drawn from one of two clusters (blue and green dots) in response to the same stimulus.
 The green cluster appears correlated with non-unitary gradient, suggestive of an underlying factor, which covaries first spike times, such that distributions of differences would be smaller when factor-conditioned, offering a higher power for discriminable encoding.
@@
 Fig.
-4g shows the amount of variance explained by the stationary angled cluster correlations.
+4g shows the histogram of variance explained by the stationary angled cluster correlations.
 Autocorrelations, cross-correlations, partial autocorrelations and partial cross-correlations were at chance levels for the stationary clusters (Supplementary Fig.
@@
 % and $s_1 = \lambda_1 \eta + \sigma_1$, 
-where $\eta$ is some underlying factor, 
-% $\lambda_i$ is the gradient at which the factor affects the spike time of unit $i$, 
+where $\eta$ represents the value of the underlying state, 
+$\lambda_i$ is the gradient at which the spike time of unit $i$ varies w.r.t. the state $\eta$, 
 and $\sigma_i$ is the predicted standard deviation of $s_i$ given $\eta$. Bootstrapping (1000 iterations) was used to calculate means of these values, which were used in subsequent analyses.

### 2020-07-29 manuscript.tex
Jolly goood

@@
 
-\title{evidence of stationary and non-stationary clustered time-warped spike timing in cortical networks}
+\title{evidence of adaption-independent time-warp spike timing in cortical networks}
 
@@
 
-Responses following sensory stimulation can be analysed to explore principles of neural coding.
-Due to response variability, characterising the neural representation of a single stimulus is a matter of characterising the \textit{distribution} of possible neural responses for the stimulus.
-A multi-neuron response can be expressed as a point in multi-dimensional space, where each dimension represents a spike train metric.
-Responses from multiple presentations of a stimulus form a sample stimulus-response distribution in this space (Fig.
+Responses following sensory stimulation can be analysed to explore such principles of neural coding. 
+% Due to response variability, characterising the neural representation of a single stimulus is a matter of characterising the \textit{distribution} of possible neural responses for the stimulus. 
+A multi-neuron response can be expressed as a point in multi-dimensional space, where each dimension represents the value of a spike train metric.
+Responses from multiple presentations of a stimulus form a sample stimulus-response distribution in this space, which can be used to estimate an underlying response distribution (Fig.
 1a).
+Estimating the distribution from which a stimulus response is `drawn' from on a single trial is complicated by correlative structure, neural adaptation and heterogeneity of sample response distribution structure.
 % which approximates an underlying stimulus-response distribution.
@@
 
-Information theoretic approaches aim to understand 
+% Information theoretic approaches aim to understand 
 If two stimuli have a behaviourally relevant difference, downstream neurons must be able to respond differently to samples from the response distribution of either stimulus.
@@
 1a).
-Characterising the distribution from which a stimulus response is `drawn' from on a single trial is complicated by correlative structure, neural adaptation and heterogeneity of sample response distribution structure.
-
-Correlative structure between the responses of different neurons in single stimulus response distributions indicates the presence of noise shared between the responses of neurons.
-These are referred to as `noise correlations' and are the focus of this manuscript, rather than `signal correlations' which refer to tuning similarity between neurons over multiple stimuli.
-
-Information theoretic approaches classically consider neural encoding with respect to fixed-state encoding and decoding.
-These approaches often do not consider that correlative structure (i.e.
-covariance of multi-neuron responses over trials of a single stimulus) could be caused by a \textit{continuous} underlying state, which covaries neural responses from trial-to-trial.
-
-Firstly, correlative structure between the responses of spatially distanced neurons (see \cite{reid2012functional, feldmeyer2013barrel}) in a single stimulus response distribution, opens the possibility of a \textit{continuous} underlying state, which covaries neural responses from trial-to-trial.
-Response samples on a single trial, when the continuous state has a certain value, would then come from a state-conditioned stimulus-response distribution.
-This would be a smaller region of stimulus-response space than the region from which all samples come (Fig.
+Neural coding is classically interested in determining the dimensions in which response distributions do not overlap \cite{averbeck2006neural} using approaches from information theory.
+
+Under such approaches, `noise correlations' refer to shared trial-to-trial variability between the responses of multiple neurons for repeated presentations of a single stimulus \cite{averbeck2006neural}.
+This paper considers the precise nature of noise correlations for single stimuli.
+Noise correlations differ from `signal correlations' which concern tuning similarity between neurons over multiple stimuli (see \cite{averbeck2006neural}).
+Information theory approaches classically consider neural coding and noise correlations with respect to fixed-state decoders \cite{averbeck2006neural, moreno2014information, stringer2019high}.
+% (and so the stimulus-dependency of noise correlations is left to future studies).
+
+% Correlative structure between the responses of different neurons in single stimulus response distributions indicates the presence of trial-to-trial noise shared between the responses of multiple neurons.
+These are classically referred to as `noise correlations' \cite{averbeck2006neural}, rather than
+It is important, however, to consider that noise correlations between spatially distanced neurons (see \cite{reid2012functional, feldmeyer2013barrel}) could be caused by an underlying \textit{continuous} single or low-dimensional latent state covarying neural responses from trial-to-trial.
+A response on a single trial, when the continuous latent state has a certain value, would then come from a \textit{state-conditioned} stimulus-response distribution.
+This would be a smaller region of stimulus-response space than the region from which all samples are drawn from over multiple presentations of a single stimulus (Fig.
 1c).
-Therefore, if downstream neurons could respond in a state-dependent manner (contrary to common neural coding assumptions \cite{moreno2014information, stringer2019high}), state-dependent encoding would increase the capacity for discriminable stimulus encoding.
-
-Characterisation of response distribution structure is also complicated by neural adaptation, which occurs across species and modalities \cite{dragoi2002dynamics, ulanovsky2003processing, sharpee2006adaptive}, and in barrel cortex to repetitive whisker stimulation for spike count \cite{ahissar2000transformation, ahissar2001temporal, kheradpezhouh2017response, khatri2009stimulus, martin2014tactile, barros2019response} and spike latency \cite{ahissar2000transformation, ahissar2001temporal, kheradpezhouh2017response}.
-That is, the response distributions are \textit{non-stationary} and change as a function of trial index for some neurons.
-Adaptation could therefore lead to correlative structure in single stimulus response distributions.
-Adaptation must therefore be controlled for when searching for correlative structure indicative of a possible underlying state.
+Therefore, if downstream neurons could respond in a state-dependent manner (contrary to classical neural coding assumptions \cite{averbeck2006neural, moreno2014information, stringer2019high}), state-dependent encoding would increase the capacity for discriminable stimulus encoding.
+
+Characterising response correlations and stimulus-response distributions at a point in time is complicated by neural adaptation, which occurs across species and modalities \cite{dragoi2002dynamics, ulanovsky2003processing, sharpee2006adaptive}.
+That is, the response distributions are \textit{non-stationary} and change monotonically as a function of trial index for some neurons.
+Processes of adaptation, which we define as different to possible \textit{non-adaptive} underlying latent states, could also lead to observed correlative structure between neuronal responses (Fig.
+X).
+Adaptation must therefore be controlled for when characterising correlative structure, which could be indicative of \textit{non-adaptive} underlying latent states.
 
 The topographic representation of each single whisker on the snout of a rodent in the somatosensory cortex, the so-called barrel cortex, represents an excellent model to study neuronal processing of defined sensory stimulation \cite{feldmeyer2013barrel}.
-The shortness of whisker deflections reduces ambiguity in what temporal aspect of a stimulus neural activity represents, offering controlled study of the neural representation of atomic entities.
-The majority of barrel cortex neurons respond 0 or 1 times to single whisker deflections \cite{reyes2014laminar}, implicating single spike patterns as a candidate for atomic representation of information.
+The shortness of whisker deflections reduces ambiguity in what temporal aspect of a stimulus neural activity represents.
+The majority of barrel cortex neurons respond 0 or 1 times to single whisker deflections \cite{reyes2014laminar}, implicating single spike patterns as a candidate for the representation of atomic stimuli.
 Representation by single spike patterns is supported in other sensory modalities, such as vision, where a recent optogenetics study showed that early visual cortex patterns of single spikes are sufficient to enable visual discrimination \cite{resulaj2018first}.
+Adaptation occurs in the barrel cortex to repetitive whisker stimulation for both spike count \cite{ahissar2000transformation, ahissar2001temporal, kheradpezhouh2017response, khatri2009stimulus, martin2014tactile, barros2019response} and spike latency \cite{ahissar2000transformation, ahissar2001temporal, kheradpezhouh2017response}.
+
+[Up to here]
 
 First spike times with millisecond order precision have been observed relative to stimulus onset in sensory organs \cite{johansson2004first, uzzell2004precision, gollisch2008rapid}, thalamic areas \cite{reinagel2002precise, storchi2012comparison} and early sensory cortices \cite{reyes2014laminar}, as well as relative to oscillations \cite{havenith2011synchrony}.
-Relative differences \cite{izhikevich2006polychronization} observed experimentally \cite{johansson2004first, gollisch2008rapid, reyes2014laminar} have been shown to encode information when trial averaged \cite{gollisch2008rapid} and do not require reference to an onset signal.
-Response distributions may form multiple clusters, implying stochasticity in the cluster from which representations come on a single trial (Fig. 1d; \cite{reyes2014laminar}).
+Relative spike time differences \cite{izhikevich2006polychronization} observed experimentally \cite{johansson2004first, gollisch2008rapid, reyes2014laminar} do not require reference to an onset signal and have been shown to encode information when trial averaged \cite{gollisch2008rapid}.
+Response distributions may form multiple clusters, implying stochasticity in the cluster from which representations of a single stimulus come on a single trial (Fig. 1d; \cite{reyes2014laminar}).
 % as neurons are sensitive to the spatial and temporal pattern of synaptic activations \cite{gasparini2006state, branco2010dendritic, branco2011synaptic}.
@@
 1.
-Simple observation of these plots (ignoring lines, colours and ellipses at this stage) shows that clusters of first spike pair responses appear correlated, suggesting that first spike pairs are correlated for pairs of neurons on a subset of trials of a stimulus.
+Simple observation (ignoring lines, colours and ellipses at this stage) of these single stimulus first spike pair response distributions shows the presence of isolated clusters, which often appear positively correlated.
+A novel algorithm extracted the observed clusters and found that they were positively correlated above chance levels, showing that first spike pairs within isolated regions of response distribution space are positively correlated above chance-levels.
 
@@
 
-Moreover, the correlation angle of some of the correlated clusters appears to be different from $45^{\circ}$ (i.e. has a gradient different from 1).
+Further observation of the response distributions, shows that the correlation angle of some of the correlated clusters appears to be different from $45^{\circ}$ (i.e.
+has a gradient different from 1).
 A simple shift in the precise timing of stimulus onset or the onset of cortical activity, would produce correlations with a unitary gradient (i.e.
 a $45^{\circ}$ angle).
-The observed correlations with `non-unitary gradient' may suggest that correlations are due to an internal process, for example, a shared cortical state underlying state-dependent processing.
+The observed correlations, instead have `non-unitary gradients' (different from $45^{\circ}$), which may be indicative that correlations are due to an internal process, for example, a shared latent state underlying state-dependent processing.
+Moreover, non-unitary correlation angles would suggest that the distribution of decodable relative spike time differences for the pair of neurons also varies with respect to the underlying latent state.
 Such correlations with `non-unitary gradient' must be considered after checking and controlling for neural adaptive trends.
 
-A novel algorithm extracted the observed clusters and found that they were correlated above chance levels.
 To control for adaptive trends, the cluster spike times of each neuron in the pair were tested for stationarity.
 The cluster spike times of both units were stationary for X\% of correlated clusters.
-X\% of these stationary correlated clusters had non-unitary (different from $45^{\circ}$) correlation angles, suggestive of an internal underlying state co-varying first spike times and thus \textit{warping} the distribution of (decodable) relative spike time differences from trial-to-trial (Fig.
+X\% of these stationary correlated clusters had angles of correlation which were significantly different from $45^{\circ}$ (non-unitary), suggestive of an internal underlying latent state co-varying first spike times (and relative differences) from trial-to-trial (Fig.
 1d).
-
 For the remaining clusters, with at least one unit with non-stationary cluster spike times, time series analysis was applied to remove the neural adaptive trends of the non-stationary unit/units.
 X\% of the detrended clusters were correlated with non-unitary gradient, suggestive of an underlying state.
-
-Application of state analysis to the stationary and detrended clusters provided an estimate of possible state-dependent response distributions.
-This gave a prediction of the resolution of possible state-dependent encoding, which would surpass previously proposed coding forms and controls for neural adaptive trends.
-It also provides a precise understanding of stimulus responses and correlative variability, which could be interpreted classically (i.e.
-independently of state [Vicente Refs]).
+The analyses also provides a useful characterisation of spike-timing correlative variability for single stimuli, which could be useful to understanding how different stimuli are encoded under the classical neural coding interpretation (i.e.
+independently of a shared state).
+
+Application of factor analysis to the stationary and detrended clusters provided an estimate of state-dependent response distributions of first spike pairs (and relative first spike differences).
+This gave a prediction of the resolution of possible state-dependent encoding, which would surpass previously proposed coding forms.
+Correlated clusters appeared for pairs of neurons which were spatially separated over multiple barrels and cortical layers.
+The non-unitary positive correlations between spatially distanced neurons is tantalisingly suggestive of a single shared lantent state, that could also modulate downstream neuron and allow them to respond in a state-dependent manner, with the benefits of high-capacity state-dependent encoding.
+The number of low-dimensional continuous latent states underlying the correlative structure, however, must be confirmed.
+
+Estimating whether cluster correlations observed for different pairs of neurons is modulated by a single or low-dimensional shared state was made difficult by variability between clusters in the trials from which cluster samples came.
+[I have begun developing several techniques to address this but I think it may prove difficult to show anything strongly conclusive either way.
+I can give more details of this once we are agreed on the overall objectives.
+We may be able to address this issue, but it may turnout to require future experiments designed to get a higher cluster count.
+A bit of help attacking this problem and working on factor analysis in 3+ dimensions would also be helpful.]
 
 The precise characterisation of correlative structure in this paper is a necessary component step towards either rejecting state-dependent coding, or confirming its precise nature. 
-It is important for neurophysiology to determine whether correlative structure between spatially distanced neurons is caused by continuous state, which also modulates downstream decoding neurons.
-Testing the state-dependent decoding performance of the response properties uncovered in this paper would require explicit knowledge (or an estimate) of the underlying trial-to-trial state so is left for future studies.
-Subthreshold membrane potentials are a candidate for such state, as they affect action potential initiation \cite{petersen2003interaction} and are correlated across areas \cite{poulet2008internal}.
-
-The characterised response distribution structure is referred to as `stationary and non-stationary clustered time-warped spike timing'.
+% It is important for neurophysiology to determine whether correlative structure between spatially distanced neurons is caused by continuous state, which also modulates downstream decoding neurons. 
+Testing the state-dependent decoding performance of the response properties uncovered in this paper is left for future studies, as it would require explicit knowledge of the underlying shared state or an estimate from the spike times of other neurons (which is difficult for the current extracted cluster counts).
+Subthreshold membrane potentials would be a candidate for such state, as they affect action potential initiation \cite{petersen2003interaction} and are correlated across areas \cite{poulet2008internal}.
+The LFP may offer a useful proxy for the subthreshold membrane potential which could predict the correlations uncovered in this paper but it may prove useful to determine the dimensionality of the states underlying the observed correlations first.
+
+The characterised response distribution structure is referred to as `adaption-independent clustered time-warped spike timing' and describes that first spike time clusters for pairs of neurons are correlated with non-unitary gradient, such that the relative spike timing encoding is `warped' from trial-to-trial by some `non-adaptive' continuous latent state.
 This differs from previous `time-warp' studies, which aim to align spike trains over trials and do not consider discriminable encoding, clustering, non-stationarity and different warping functions for different neurons \cite{williams2020discovering}.
@@
 
-The results suggest that consideration of clustering, state-dependence and non-stationarity, together, are necessary to explain a large proportion of neural variability, and could explain the failure to find a fast and efficient ubiquitous spike code that utilises the high sensitivity of neural architecture and functional dynamics.
+The results suggest that consideration of clustering, correlative structure and adaption, together, are necessary to explain a large proportion of neural variability, and could explain the failure to find a fast and efficient ubiquitous spike code that utilises the high sensitivity of neural architecture and functional dynamics.
 In this study, such consideration gave a novel prediction of the resolution of neural stimulus representation.

### 2020-07-30 manuscript.tex
V2.0 - Introduction

--- 2020-07-24 manuscript.tex
@@
 
-\title{evidence of stationary and non-stationary clustered time-warped spike timing in cortical networks}
+\title{evidence of adaption-independent time-warped spike timing noise correlations in cortical networks}
 
@@
 \section{Introduction}
-The neural coding problem asks how information is encoded through patterns of action potentials in the nervous system \cite{gerstner1997neural, johnson2000neural, stein2005neuronal, gerstner2014neuronal}, with a dichotomy between temporal \cite{uzzell2004precision, johansson2004first, gollisch2008rapid, storchi2012comparison, reinagel2002precise, reyes2014laminar} and rate codes \cite{hong2016explicit, chang2017code, kar2019evidence, stringer2019high}.
-% Temporal codes provide the possibility of a fast and efficient code. 
-Whilst precise temporal patterns have been observed \cite{uzzell2004precision, johansson2004first, gollisch2008rapid, storchi2012comparison, reinagel2002precise, reyes2014laminar}, a large proportion of neural variability remains unexplained.
-Such variability leads to the common assumption of average firing rate codes, under the premise that variability cancels out across a population \cite{abbott1999effect, moreno2014information, stringer2019high}.
-Continuous firing rates (averaged over time, trial or population \cite{gerstner1997neural}) have capacity beyond the single trial discrete spike count from which they are abstracted, leaving a full understanding of neural coding and variability undetermined.
-It is important to better characterise neural codes in order to understand a rich variety of dynamic, structural, homeostatic and cognitive processes, as well as properties of action potential patterns directly involved in behaviour.
+The neural coding problem asks how information is encoded through patterns of action potentials in the nervous system \cite{gerstner1997neural, johnson2000neural, stein2005neuronal, gerstner2014neuronal}.
+Whilst precise temporal patterns have been observed \cite{uzzell2004precision, johansson2004first, gollisch2008rapid, storchi2012comparison, reinagel2002precise, reyes2014laminar}, a large proportion of spike time variability remains unexplained.
+Such variability leads to the common assumption of population codes based on spike counts \cite{reyes2015high, lin2015nature} or trial-averaged firing rates \cite{hong2016explicit, chang2017code, kar2019evidence, stringer2019high}.
+% \cite{gerstner1997neural}, which have capacity beyond the original spike count \cite{}. 
+% Continuous firing rates (averaged over time, trial or population ) 
+% These population codes assume that trial-to-trial variability is either independent or correlated across the population \cite{averbeck2006neural, lin2015nature}. 
+Trial-to-trial spike count variability is in fact correlated across the population in early sensory cortices \cite{reyes2015high, lin2015nature} and was shown to be modulated by a two-dimensional continuous state \cite{lin2015nature}.
+It is possible that spike time variability is modulated across a population by the same or similar low-dimensional underlying state.
+This study aims to characterise the nature of trial-to-trial spike time correlations as a step towards confirming or rejecting the modulation of spike times by an underlying low-dimensional latent state.
+This offers a novel prediction of the encoding capacity of state-dependent spike time encoding.
+Characterising spike timing codes is important to understanding a variety of dynamic, structural, homeostatic and cognitive processes, as well as properties of action potential patterns directly involved in behaviour.
+
+% Continuous firing rates (averaged over time, trial or population \cite{gerstner1997neural}) have capacity beyond the single trial discrete spike count from which they are calculated, leaving a full understanding of neural coding and variability undetermined.
+
+% , under the premise that variability cancels out across a population \cite{abbott1999effect, moreno2014information, stringer2019high} or that variability is shared across the population and modulated from trial-to-trial by some underlying low-dimensional latent state \cite{lin2015nature}.
+
 % Characterising responses may also uncover response properties involved in behaviour.
 
-Responses following sensory stimulation can be analysed to explore principles of neural coding.
-Due to response variability, characterising the neural representation of a single stimulus is a matter of characterising the \textit{distribution} of possible neural responses for the stimulus.
-A multi-neuron response can be expressed as a point in multi-dimensional space, where each dimension represents a spike train metric.
-Responses from multiple presentations of a stimulus form a sample stimulus-response distribution in this space (Fig.
+Responses following sensory stimulation can be analysed to explore such principles of neural coding. 
+% Due to response variability, characterising the neural representation of a single stimulus is a matter of characterising the \textit{distribution} of possible neural responses for the stimulus. 
+Particularly, a multi-neuron response can be expressed as a point in multi-dimensional space, where each dimension represents the value of a spike train metric.
+Responses from multiple presentations of a stimulus form a sample stimulus-response distribution in this space, which can be used to estimate an underlying response distribution from which single-trial responses are `drawn' (Fig.
 1a).
+% Estimating the underlying response distribution (including correlative structure) for single stimuli is complicated by neural adaptation and heterogeneity of sample response distribution structure.
 % which approximates an underlying stimulus-response distribution.
@@
 1a).
-Characterising the distribution from which a stimulus response is `drawn' from on a single trial is complicated by correlative structure, neural adaptation and heterogeneity of sample response distribution structure.
-
-Correlative structure between the responses of different neurons in single stimulus response distributions indicates the presence of noise shared between the responses of neurons.
-These are referred to as `noise correlations' and are the focus of this manuscript, rather than `signal correlations' which refer to tuning similarity between neurons over multiple stimuli.
-
-Information theoretic approaches classically consider neural encoding with respect to fixed-state encoding and decoding.
-These approaches often do not consider that correlative structure (i.e.
-covariance of multi-neuron responses over trials of a single stimulus) could be caused by a \textit{continuous} underlying state, which covaries neural responses from trial-to-trial.
-
-Firstly, correlative structure between the responses of spatially distanced neurons (see \cite{reid2012functional, feldmeyer2013barrel}) in a single stimulus response distribution, opens the possibility of a \textit{continuous} underlying state, which covaries neural responses from trial-to-trial.
-Response samples on a single trial, when the continuous state has a certain value, would then come from a state-conditioned stimulus-response distribution.
-This would be a smaller region of stimulus-response space than the region from which all samples come (Fig.
+Neural coding is classically interested in determining the dimensions in which response distributions do not overlap \cite{averbeck2006neural}.
+
+Under such approaches, `noise correlations' refer to shared trial-to-trial variability between the responses of multiple neurons over repeated presentations of single stimulus \cite{averbeck2006neural}.
+`Signal correlations' concern tuning similarity between neurons over multiple stimuli (see \cite{averbeck2006neural}).
+As discussed, previous studies have demonstrated that spike count noise correlations are modulated across a population  by low-dimensional continuous latent states \cite{lin2015nature}.
+% For example, a 2-dimensional continuous latent state modulates responses in V1 \cite{lin2015nature}.
+
+Neural coding and noise correlations are classically considered with respect to fixed-state decoders \cite{averbeck2006neural, moreno2014information, stringer2019high}.
+It is important to consider, that if downstream neurons are also modulated by the same population wide continuous latent state, then downstream `decoder' neurons could respond in a state-dependent manner.
+When the continuous latent state has a certain value on a single trial, a response comes from a \textit{state-conditioned} stimulus-response distribution, which would occupy a smaller region of response distribution space than the region sampled over multiple trials (Fig.
 1c).
-Therefore, if downstream neurons could respond in a state-dependent manner (contrary to common neural coding assumptions \cite{moreno2014information, stringer2019high}), state-dependent encoding would increase the capacity for discriminable stimulus encoding.
-
-Characterisation of response distribution structure is also complicated by neural adaptation, which occurs across species and modalities \cite{dragoi2002dynamics, ulanovsky2003processing, sharpee2006adaptive}, and in barrel cortex to repetitive whisker stimulation for spike count \cite{ahissar2000transformation, ahissar2001temporal, kheradpezhouh2017response, khatri2009stimulus, martin2014tactile, barros2019response} and spike latency \cite{ahissar2000transformation, ahissar2001temporal, kheradpezhouh2017response}.
-That is, the response distributions are \textit{non-stationary} and change as a function of trial index for some neurons.
-Adaptation could therefore lead to correlative structure in single stimulus response distributions.
-Adaptation must therefore be controlled for when searching for correlative structure indicative of a possible underlying state.
-
-The topographic representation of each single whisker on the snout of a rodent in the somatosensory cortex, the so-called barrel cortex, represents an excellent model to study neuronal processing of defined sensory stimulation \cite{feldmeyer2013barrel}.
-The shortness of whisker deflections reduces ambiguity in what temporal aspect of a stimulus neural activity represents, offering controlled study of the neural representation of atomic entities.
-The majority of barrel cortex neurons respond 0 or 1 times to single whisker deflections \cite{reyes2014laminar}, implicating single spike patterns as a candidate for atomic representation of information.
-Representation by single spike patterns is supported in other sensory modalities, such as vision, where a recent optogenetics study showed that early visual cortex patterns of single spikes are sufficient to enable visual discrimination \cite{resulaj2018first}.
-
-First spike times with millisecond order precision have been observed relative to stimulus onset in sensory organs \cite{johansson2004first, uzzell2004precision, gollisch2008rapid}, thalamic areas \cite{reinagel2002precise, storchi2012comparison} and early sensory cortices \cite{reyes2014laminar}, as well as relative to oscillations \cite{havenith2011synchrony}.
-Relative differences \cite{izhikevich2006polychronization} observed experimentally \cite{johansson2004first, gollisch2008rapid, reyes2014laminar} have been shown to encode information when trial averaged \cite{gollisch2008rapid} and do not require reference to an onset signal.
-Response distributions may form multiple clusters, implying stochasticity in the cluster from which representations come on a single trial (Fig. 1d; \cite{reyes2014laminar}).
+State-dependent decoding would therefore offer an improved encoding capacity, compared to the classical fixed-state decoder interpretation.
+
+% Correlative structure between the responses of different neurons in single stimulus response distributions indicates the presence of trial-to-trial noise shared between the responses of multiple neurons.
+These are classically referred to as `noise correlations' \cite{averbeck2006neural}, rather than
+
+% reid2012functional
+% It is important, however, to consider that trial-to-trial noise correlations between spatially distanced neurons could be caused by an underlying \textit{continuous} single or low-dimensional latent state that varies from trial-to-trial.
+Such state-dependent coding would depend on the ability of downstream neurons to respond in a state-dependent manner.
+This could be possible if downstream neurons were also modulated by the same underlying continuous latent states.
+% Such state-dependent coding differs from the classical neural coding assumption of fixed-state decoders \cite{averbeck2006neural, moreno2014information, stringer2019high}).
+
+% This paper aims to characterise the precise nature of spike timing noise correlations for single stimuli to uncover the potential encoding capacity of a state-dependent spike time code. 
+% The stimulus-dependency of noise correlations is left to future studies.  
+Characterising noise correlations within stimulus-response distributions is complicated by the heterogeneity of sample response distribution structure and neural adaptation, which occurs across species and modalities \cite{dragoi2002dynamics, ulanovsky2003processing, sharpee2006adaptive}.
+% That is, the response distributions are \textit{non-stationary} and change monotonically as a function of trial index for some neurons. 
+Processes of adaptation, which we define as different to latent states underlying trial-to-trial noise correlations, could also lead to correlative structure in the stimulus-response distribution (Fig.
+X).
+Adaptation must therefore be controlled for when characterising adaption-independent noise correlations.
+
+The topographic representation of each single whisker on the snout of a rodent in the somatosensory cortex, the so-called barrel cortex, represents an excellent model to study noise correlations for controlled sensory stimulation \cite{feldmeyer2013barrel}.
+% The shortness of whisker deflections reduces ambiguity in what temporal aspect of a stimulus neural activity represents. 
+The majority of barrel cortex neurons respond 0 or 1 times to single whisker deflections \cite{reyes2014laminar}, implicating single spike patterns as a candidate for encoding.
+Single spike pattern encoding is supported in other sensory modalities, such as vision, where patterns of single spikes are sufficient to enable visual discrimination \cite{resulaj2018first}.
+
+The nature of first spike patterns has been explored in previous research.
+First spike times have been observed with millisecond order precision relative to stimulus onset across species and modalities in sensory organs \cite{johansson2004first, uzzell2004precision, gollisch2008rapid}, thalamic areas \cite{reinagel2002precise, storchi2012comparison} and early sensory cortices \cite{reyes2014laminar}, as well as relative to oscillations \cite{havenith2011synchrony}.
+First spike times in the rat barrel cortex have been shown to improve stimulus encoding by 44\% over spike count \cite{panzeri2001role}.
+Precise relative differences between the spike times of multiple neurons \cite{izhikevich2006polychronization} do not require reference to an onset signal and have been observed experimentally \cite{johansson2004first, gollisch2008rapid, reyes2014laminar} and shown to encode stimulus information when averaged over trials \cite{gollisch2008rapid}.
+
+Using a large data set of spike sorted recordings in anesthetized rat barrel cortex, this paper aimed to precisely characterise the nature of spike time noise correlations for first-spike patterns, towards estimating the potential encoding capacity of a state-dependent spike time encoding and determining the dimensionality of such an underlying latent state.
+The classical `stimulus-dependency' of noise correlations is left to future studies.
+Previous studies analysing the same dataset found precise multi-neuron intralaminar spike patterns \cite{reyes2014laminar}, spike count noise correlations and could decode stimulus properties (stimulation frequency and locaction) from spike counts and spike patterns \cite{reyes2015high}.
+The new analysis uses time series analysis to control for neural adaption over all stimulus trials.
+Adaption occurs in the barrel cortex to repetitive whisker stimulation for both spike count \cite{ahissar2000transformation, ahissar2001temporal, kheradpezhouh2017response, khatri2009stimulus, martin2014tactile, barros2019response} and spike latency \cite{ahissar2000transformation, ahissar2001temporal, kheradpezhouh2017response}, and was confirmed to occur for 7 out of 8 neuron groups in this study.
+
 % as neurons are sensitive to the spatial and temporal pattern of synaptic activations \cite{gasparini2006state, branco2010dendritic, branco2011synaptic}.
 
-Using a large data set of spike sorted recordings in anesthetized rat barrel cortex, first spike response distributions were first observed for single stimuli and pairs of units.
-That is, for repeated trials of a single stimulus, the first spike times of two units were plotted against each other for trials on which they both spiked.
-The reader is strongly recommended to view these plots in the top left of Supplementary Vid.
+For repeated trials of a single stimulus, the first spike times of two units were plotted against each other for trials on which they both spiked.
+The reader is encouraged to view these plots in the top left of Supplementary Vid.
 1.
-Simple observation of these plots (ignoring lines, colours and ellipses at this stage) shows that clusters of first spike pair responses appear correlated, suggesting that first spike pairs are correlated for pairs of neurons on a subset of trials of a stimulus.
-
-% That is, each axis corresponds to the first spike time of a different recorded unit.
-The scatter points therefore represent the first spike times of both units for the trials (of a single stimulus) on which both units spiked.
-
-Moreover, the correlation angle of some of the correlated clusters appears to be different from $45^{\circ}$ (i.e. has a gradient different from 1).
-A simple shift in the precise timing of stimulus onset or the onset of cortical activity, would produce correlations with a unitary gradient (i.e.
+A novel algorithm extracted isolated clusters from these single stimulus first spike pair response distributions, demonstrating that temporally precise spike patterns fire on subsets of trials, as shown previously for the data set \cite{reyes2014laminar}.
+The novel algorithm, however, also showed that first spike pairs within these isolated clusters were positively correlated above chance-levels.
+
+Moreover, the correlation angle of many of the correlated clusters was different from $45^{\circ}$.
+This cannot be explained by a trial-to-trial shift in the onset of the stimulus or cortical activity, which would produce correlations with a unitary gradient (i.e.
 a $45^{\circ}$ angle).
-The observed correlations with `non-unitary gradient' may suggest that correlations are due to an internal process, for example, a shared cortical state underlying state-dependent processing.
-Such correlations with `non-unitary gradient' must be considered after checking and controlling for neural adaptive trends.
-
-A novel algorithm extracted the observed clusters and found that they were correlated above chance levels.
-To control for adaptive trends, the cluster spike times of each neuron in the pair were tested for stationarity.
-The cluster spike times of both units were stationary for X\% of correlated clusters.
-X\% of these stationary correlated clusters had non-unitary (different from $45^{\circ}$) correlation angles, suggestive of an internal underlying state co-varying first spike times and thus \textit{warping} the distribution of (decodable) relative spike time differences from trial-to-trial (Fig.
-1d).
-
-For the remaining clusters, with at least one unit with non-stationary cluster spike times, time series analysis was applied to remove the neural adaptive trends of the non-stationary unit/units.
-X\% of the detrended clusters were correlated with non-unitary gradient, suggestive of an underlying state.
-
-Application of state analysis to the stationary and detrended clusters provided an estimate of possible state-dependent response distributions.
-This gave a prediction of the resolution of possible state-dependent encoding, which would surpass previously proposed coding forms and controls for neural adaptive trends.
-It also provides a precise understanding of stimulus responses and correlative variability, which could be interpreted classically (i.e.
-independently of state [Vicente Refs]).
-
-The precise characterisation of correlative structure in this paper is a necessary component step towards either rejecting state-dependent coding, or confirming its precise nature.
-It is important for neurophysiology to determine whether correlative structure between spatially distanced neurons is caused by continuous state, which also modulates downstream decoding neurons.
-Testing the state-dependent decoding performance of the response properties uncovered in this paper would require explicit knowledge (or an estimate) of the underlying trial-to-trial state so is left for future studies.
-Subthreshold membrane potentials are a candidate for such state, as they affect action potential initiation \cite{petersen2003interaction} and are correlated across areas \cite{poulet2008internal}.
-
-The characterised response distribution structure is referred to as `stationary and non-stationary clustered time-warped spike timing'.
-This differs from previous `time-warp' studies, which aim to align spike trains over trials and do not consider discriminable encoding, clustering, non-stationarity and different warping functions for different neurons \cite{williams2020discovering}.
+Instead the `non-unitary' correlation gradients (different from $45^{\circ}$) may be indicative that correlations are due to an internal shared latent state.
+Clusters remained correlated with non-unitary gradient after controlling for neural adaption using time series analysis, confirming the correlated clusters as evidence of precise non-adaptive noise correlations.
+A large proportion of these noise correlations would have been missed without clustering.
+
+Importantly, the observation of adaption-independent correlated clusters with non-unitary gradient shows that the distribution of decodable first spike time relative differences for pairs of neurons are `warped' from trial-to-trial by some `non-adaptive' continuous latent state.
+This phenomenon is referred to as `adaption-independent clustered time-warped spike timing'.
+This differs from previous `time-warp' studies, which aim to align spike trains over trials and do not consider discriminable encoding, clustering, adaption and different warping functions for different neurons \cite{williams2020discovering}.
 One previous study has modelled time-warp-invariant integration \cite{gutig2009time}, although neurons need not respond identically to warps of the same pattern.
-Rather, neurons should process relative spike patterns in a state-dependent manner.
-
-The results suggest that consideration of clustering, state-dependence and non-stationarity, together, are necessary to explain a large proportion of neural variability, and could explain the failure to find a fast and efficient ubiquitous spike code that utilises the high sensitivity of neural architecture and functional dynamics.
+Rather, neurons should process relative spike patterns in a state-dependent manner to utilise the improved state-dependent coding capacity.
+
+Adaption-controlled non-unitary correlated clusters appeared for pairs of neurons spatially separated over multiple barrels and cortical layers.
+This offers suggestive evidence of a shared low-dimensional continuous latent state affecting the spike times of spatially distanced neurons, which in turn could also modulate downstream neurons and allow them to respond in a state-dependent manner.
+Factor analysis provided an estimate of state-dependent response distributions of first spike pairs (and relative first spike differences) for each adaption-controlled correlated cluster.
+This gave a prediction of the resolution of possible state-dependent encoding, which surpasses previously proposed coding forms.
+
+The number of low-dimensional continuous latent states underlying the correlative structure, however, must be confirmed.
+Estimating whether cluster correlations observed for different pairs of neurons is modulated by a single or low-dimensional shared state was made difficult by variability between clusters in the trials from which cluster samples came [See notes below].
+The work of this paper in extracting and characterising adaption-independent correlated clusters provides a necessary component step towards confirming or rejecting state-dependent spike-time coding, and is also important under the classical fixed-state decoder interpretation.
+The results show that consideration of clustering, correlative structure and adaption, together, are necessary to explain a large proportion of neural variability.
+
+% It is important for neurophysiology to determine whether correlative structure between spatially distanced neurons is caused by continuous state, which also modulates downstream decoding neurons. 
+Testing the state-dependent decoding performance of the clusters is left for future studies, as it would require explicit knowledge of the underlying shared state or an estimate from the spike times of other neurons (which is difficult for the current extracted cluster counts per stimulus in single experimental sessions).
+Subthreshold membrane potentials would be a candidate for such state, as they affect action potential initiation \cite{petersen2003interaction} and are correlated across areas \cite{poulet2008internal}.
+The LFP may offer a useful proxy for the subthreshold membrane potential.
+
+% and could explain the failure to find a fast and efficient ubiquitous spike code that utilises the high sensitivity of neural architecture and functional dynamics.
 In this study, such consideration gave a novel prediction of the resolution of neural stimulus representation.
+
+\subsection{Notes}
+I have begun developing several techniques to address this but I think it may prove difficult to show anything strongly conclusive either way.
+I can give more details of this once we are agreed on the overall objectives.
+We may be able to address this issue, but it may turnout to require future experiments designed to get a higher cluster count per stimulus and experimental session.
+A bit of help attacking this problem and working on factor analysis in 3+ dimensions would also be helpful.
 

### 2020-08-08 manuscript.tex
Carry on my wayward son

@@
 
-\title{evidence of adaption-independent time-warped spike timing noise correlations in cortical networks}
+\title{evidence of adaptation-independent time-warped spike timing noise correlations in cortical networks}
 
@@
 \section{Introduction}
-The neural coding problem asks how information is encoded through patterns of action potentials in the nervous system \cite{gerstner1997neural, johnson2000neural, stein2005neuronal, gerstner2014neuronal}.
-Whilst precise temporal patterns have been observed \cite{uzzell2004precision, johansson2004first, gollisch2008rapid, storchi2012comparison, reinagel2002precise, reyes2014laminar}, a large proportion of spike time variability remains unexplained.
+The neural coding problem addresses the question of how information is encoded by patterns of action potentials in the nervous system \cite{gerstner1997neural, johnson2000neural, stein2005neuronal, gerstner2014neuronal}.
+Whilst precise temporal spike patterns have been observed \cite{uzzell2004precision, johansson2004first, gollisch2008rapid, storchi2012comparison, reinagel2002precise, reyes2014laminar}, a large proportion of spike time variability remains unexplained.
 Such variability leads to the common assumption of population codes based on spike counts \cite{reyes2015high, lin2015nature} or trial-averaged firing rates \cite{hong2016explicit, chang2017code, kar2019evidence, stringer2019high}.
@@
 This offers a novel prediction of the encoding capacity of state-dependent spike time encoding.
-Characterising spike timing codes is important to understanding a variety of dynamic, structural, homeostatic and cognitive processes, as well as properties of action potential patterns directly involved in behaviour.
+Characterising spike timing codes is important to understanding a variety of dynamic, homeostatic and plastic processes, as well as properties of action potential patterns directly involved in cognition.
 
@@
 % Information theoretic approaches aim to understand 
-If two stimuli have a behaviourally relevant difference, downstream neurons must be able to respond differently to samples from the response distribution of either stimulus.
+If two sensory stimuli have a cognitively relevant difference, downstream neurons must be able to respond differently to samples from the response distribution of either stimulus.
 This is possible if the response distributions are highly non-overlapping in dimensions that downstream neurons are sensitive to, such as relative spike time differences (\cite{gasparini2006state, branco2010dendritic, branco2011synaptic}; Fig.
@@
 X).
-Adaptation must therefore be controlled for when characterising adaption-independent noise correlations.
+Adaptation must therefore be controlled for when characterising adaptation-independent noise correlations.
 
@@
 The classical `stimulus-dependency' of noise correlations is left to future studies.
-Previous studies analysing the same dataset found precise multi-neuron intralaminar spike patterns \cite{reyes2014laminar}, spike count noise correlations and could decode stimulus properties (stimulation frequency and locaction) from spike counts and spike patterns \cite{reyes2015high}.
-The new analysis uses time series analysis to control for neural adaption over all stimulus trials.
-Adaption occurs in the barrel cortex to repetitive whisker stimulation for both spike count \cite{ahissar2000transformation, ahissar2001temporal, kheradpezhouh2017response, khatri2009stimulus, martin2014tactile, barros2019response} and spike latency \cite{ahissar2000transformation, ahissar2001temporal, kheradpezhouh2017response}, and was confirmed to occur for 7 out of 8 neuron groups in this study.
+A previous study analysing the same dataset found precise multi-neuron spike patterns that were intra- and inter-laminar, as well intra- and intra-columnar \cite{reyes2014laminar}.
+Another study analysing the data set found spike count noise correlations and could decode stimulus properties (stimulation frequency and location) from spike counts and spike patterns \cite{reyes2015high}.
+The new analysis uses time series analysis to control for neural adaptation over all stimulus trials.
+adaptation occurs in the barrel cortex to repetitive whisker stimulation for both spike count \cite{ahissar2000transformation, ahissar2001temporal, kheradpezhouh2017response, khatri2009stimulus, martin2014tactile, barros2019response} and spike latency \cite{ahissar2000transformation, ahissar2001temporal, kheradpezhouh2017response}, and was confirmed to occur for 7 out of 8 neuron groups in this study.
 
@@
 
-For repeated trials of a single stimulus, the first spike times of two units were plotted against each other for trials on which they both spiked.
-The reader is encouraged to view these plots in the top left of Supplementary Vid.
-1.
+For repeated trials of a single stimulus, the first spike times of two units were plotted against each other for trials on which they both spiked (see plots in the top left of Supplementary Vid.
+1).
 A novel algorithm extracted isolated clusters from these single stimulus first spike pair response distributions, demonstrating that temporally precise spike patterns fire on subsets of trials, as shown previously for the data set \cite{reyes2014laminar}.
@@
 Instead the `non-unitary' correlation gradients (different from $45^{\circ}$) may be indicative that correlations are due to an internal shared latent state.
-Clusters remained correlated with non-unitary gradient after controlling for neural adaption using time series analysis, confirming the correlated clusters as evidence of precise non-adaptive noise correlations.
+Clusters remained correlated with non-unitary gradient after controlling for neural adaptation using time series analysis, confirming the correlated clusters as evidence of precise non-adaptive noise correlations.
 A large proportion of these noise correlations would have been missed without clustering.
 
-Importantly, the observation of adaption-independent correlated clusters with non-unitary gradient shows that the distribution of decodable first spike time relative differences for pairs of neurons are `warped' from trial-to-trial by some `non-adaptive' continuous latent state.
-This phenomenon is referred to as `adaption-independent clustered time-warped spike timing'.
-This differs from previous `time-warp' studies, which aim to align spike trains over trials and do not consider discriminable encoding, clustering, adaption and different warping functions for different neurons \cite{williams2020discovering}.
+Importantly, the observation of adaptation-independent correlated clusters with non-unitary gradient shows that the distribution of decodable first spike time relative differences for pairs of neurons are `warped' from trial-to-trial by some `non-adaptive' continuous latent state.
+This phenomenon is referred to as `adaptation-independent clustered time-warped spike timing'.
+This differs from previous `time-warp' studies, which aim to align spike trains over trials and do not consider discriminable encoding, clustering, adaptation and different warping functions for different neurons \cite{williams2020discovering}.
 One previous study has modelled time-warp-invariant integration \cite{gutig2009time}, although neurons need not respond identically to warps of the same pattern.
@@
 
-Adaption-controlled non-unitary correlated clusters appeared for pairs of neurons spatially separated over multiple barrels and cortical layers.
+adaptation-controlled non-unitary correlated clusters appeared for pairs of neurons spatially separated over multiple barrels and cortical layers.
 This offers suggestive evidence of a shared low-dimensional continuous latent state affecting the spike times of spatially distanced neurons, which in turn could also modulate downstream neurons and allow them to respond in a state-dependent manner.
-Factor analysis provided an estimate of state-dependent response distributions of first spike pairs (and relative first spike differences) for each adaption-controlled correlated cluster.
+Factor analysis provided an estimate of state-dependent response distributions of first spike pairs (and relative first spike differences) for each adaptation-controlled correlated cluster.
 This gave a prediction of the resolution of possible state-dependent encoding, which surpasses previously proposed coding forms.
@@
 Estimating whether cluster correlations observed for different pairs of neurons is modulated by a single or low-dimensional shared state was made difficult by variability between clusters in the trials from which cluster samples came [See notes below].
-The work of this paper in extracting and characterising adaption-independent correlated clusters provides a necessary component step towards confirming or rejecting state-dependent spike-time coding, and is also important under the classical fixed-state decoder interpretation.
-The results show that consideration of clustering, correlative structure and adaption, together, are necessary to explain a large proportion of neural variability.
-
-% It is important for neurophysiology to determine whether correlative structure between spatially distanced neurons is caused by continuous state, which also modulates downstream decoding neurons. 
-Testing the state-dependent decoding performance of the clusters is left for future studies, as it would require explicit knowledge of the underlying shared state or an estimate from the spike times of other neurons (which is difficult for the current extracted cluster counts per stimulus in single experimental sessions).
-Subthreshold membrane potentials would be a candidate for such state, as they affect action potential initiation \cite{petersen2003interaction} and are correlated across areas \cite{poulet2008internal}.
-The LFP may offer a useful proxy for the subthreshold membrane potential.
+The work of this paper in extracting and characterising adaptation-independent correlated clusters provides a necessary component step towards confirming or rejecting state-dependent spike-time coding, and is also important under the classical fixed-state decoder interpretation.
+The results show that consideration of clustering, correlative structure and adaptation, together, are necessary to explain a large proportion of neural variability.
 
@@
 
-\todo{Look at adaption by neuron type and layer}
-\todo{Select final adaption examples, with layers and stats.}
-\todo{Do adaption conclusion}
-\todo{Make adaption stats overview paragraph simpler, with nice single bar chart}
-
-This section demonstrates that the responses of some neurons exhibit depressive latency adaption, which should be controlled for when testing for correlative structure caused by a non-adaptive state.
+\todo{Look at adaptation by neuron type and layer}
+\todo{Select final adaptation examples, with layers and stats.}
+\todo{Do adaptation conclusion}
+\todo{Make adaptation stats overview paragraph simpler, with nice single bar chart}
+
+This section demonstrates that the responses of some neurons exhibit depressive latency adaptation, which should be controlled for when testing for correlative structure caused by a non-adaptive state.
 Figure 2d shows the first spike times of 5 single units in response to repeated stimulations.
@@
 
+\subsection*{Physiological Relevance}
+% It is important for neurophysiology to determine whether correlative structure between spatially distanced neurons is caused by continuous state, which also modulates downstream decoding neurons. 
+Testing the state-dependent decoding performance of the clusters is left for future studies, as it would require explicit knowledge of the underlying shared state or an estimate from the spike times of other neurons (which is difficult for the current extracted cluster counts per stimulus in single experimental sessions).
+Neuronal subthreshold membrane potentials would be a candidate for such state, as they are correlated between neurons across areas \cite{poulet2008internal} and affect action potential initiation \cite{petersen2003interaction}.
+The LFP may offer a useful proxy for the subthreshold membrane potential.
+
 \subsection*{Underlying state}

### 2020-08-10 AnalysisOutputLatex.tex
Lunch!

@@
+%<*ClusterCorrelationSummary0>
+Spike pairs within Stage 1 cluster ellipses were linearly correlated above chance levels (Fisher's method: $\chi^{2}(7812, N = 3906) = 15150.9, $ $p < .001$; Fig. 3e). 
+%</ClusterCorrelationSummary0>
+
+%<*ClusterCorrelationSummary0p1>
+Moreover, for Stage 2 clusters, spike pair differences were correlated with the spike time of the first unit in the pair (Fisher's method: $\chi^{2}(324, N = 162) = 7832.7, $ $p < .001$; Fig.
+3g), showing that correlations were not explained by a model of the form $s_1 = s_0 + d + independent\_noise$ where $d$ is a fixed difference.
+%</ClusterCorrelationSummary0p1>
+
 %<*ClusterCorrelationSummary1>
-Flat clusters were linearly correlated above chance levels (Fisher's method: $\chi^{2}(7844, N = 3922) = 15380.2, $ $p < .001$; Fig.
-3e).
-Of the final angled clusters, 97.6\% (160/164) were positively correlated (Fig.
-3f), suggesting the presence of a factor affecting response latency monotonically across the population.
-For the corresponding flat cluster of the criteria fulfilling angled clusters, spike pair differences were correlated with the spike time of the first unit in the pair (Fisher's method: $\chi^{2}(328, N = 164) = 8366.9, $ $p < .001$; Fig.
-3g), showing that correlations were not explained by a model of the form $s_1 = s_0 + d + independent\_noise$ where $d$ is a fixed difference.
+97.5\% (158/162) of Stage 2 clusters were positively correlated (Fig. 3f). 
 %</ClusterCorrelationSummary1>
@@
 %<*ClusterCorrelationSummary2>
-For the final angled clusters, 81.7\% (134/164) of the corresponding flat clusters had linear regression p-values less than the p-values of linear regression applied to the unclustered first spike pairs (Fig.
+80.9\% (131/162) of the Stage 2 clusters had correlations of higher significance than correlations calculated for all unclustered first spike pairs in the original response distribution (Fig.
 3h).
-This implies that without clustering, linear regression would miss a large proportion of first spike correlations.
+Moreover, 72.8\% (118/162) of the original response distributions from which Stage 2 clusters were extracted were not correlated significantly (p>0.005) (Fig.
+3h).
 %</ClusterCorrelationSummary2>
@@
 %<*AngledClustersUniquePairsSummary>
-Within this time window, the algorithm extracted a total of 164 angled clusters from 164 unique response distributions, such that there were no repeated or similar clusters.
+A total of 162 unique Stage 2 clusters were extracted from 162 unique response distributions.
 %</AngledClustersUniquePairsSummary>
@@
 %<*OriginalAnglesSummary>
-For the positively correlated angled clusters, Fig.
-4a shows the distribution of the mean $\theta_{45}$ angles with 95\% confidence intervals.
-66.5\% (109/164) of angles were significantly greater than $0^{\circ}$ (p-value < 0.025) and less than $45^{\circ}$ (p-value < 0.025).
+67.9\% (110/162) of $	heta_{45}$ angles were both significantly less than $45^{\circ}$ (p<0.025) and greater than $0^{\circ}$ (<0.025). 
 %</OriginalAnglesSummary>
@@
 %<*OriginalTestsPassedSummary>
-Of the 164 angled clusters, 44 were determined to be stationary and sufficiently correlated by fulfilling the criteria described in the Methods \ref{methods:angle_and_factor_analysis_criteria}, such that an underlying factor could underlie the covariation.
+Of the 162 angled clusters, 46 were determined to be stationary and sufficiently correlated by fulfilling the criteria described in the Methods \ref{methods:angle_and_factor_analysis_criteria}, such that an underlying factor could underlie the covariation.
 %</OriginalTestsPassedSummary>
@@
 %<*OriginalTestsPassedAngleSummary>
-Of the 44 criteria fulfilling angled clusters, 26 had estimated angles significantly different from both $0^{\circ}$ (p-value < 0.025; Fig.
-4f) and $45^{\circ}$ (p-value < 0.025; Fig.
+Of the 46 criteria fulfilling angled clusters, 30 had estimated angles significantly different from both $0^{\circ}$ (p<0.025; Fig.
+4f) and $45^{\circ}$ (p<0.025; Fig.
 4f).
-Of the 44 criteria fulfilling angled clusters, normality was not rejected for 18 clusters (Henze-Zirkler p > 0.05), of which 11 had an estimated angle significantly different from both $0^{\circ}$ (p-value < 0.025) and $45^{\circ}$ (p-value < 0.025).
+Of the 46 criteria fulfilling angled clusters, normality was not rejected for 20 clusters (Henze-Zirkler p > 0.05), of which 14 had an estimated angle significantly different from both $0^{\circ}$ (p-value < 0.025) and $45^{\circ}$ (p-value < 0.025).
 %</OriginalTestsPassedAngleSummary>
@@
 %<*OriginalTestsPassedFactorAnalysisSummary>
-Factor analysis could be applied to the 44 criteria fulfilling angled clusters, to give an estimate of the factor-conditioned cluster response distributions for the stationary angled clusters.
+Factor analysis could be applied to the 46 criteria fulfilling angled clusters, to give an estimate of the factor-conditioned cluster response distributions for the stationary angled clusters.
 %</OriginalTestsPassedFactorAnalysisSummary>
@@
 %<*ClusterSingleUnitKPSSStationaritySummary>
-For the 164 angled clusters, 118 of the corresponding 328 ($=2*164$) single unit cluster first spike sequences were determined as KPSS non-stationary (p < 0.05; Fig. 4d).
+For the 162 angled clusters, 117 of the corresponding 324 ($=2*162$) single unit cluster first spike sequences were determined as KPSS non-stationary (p < 0.05; Fig. 4d).
 %</ClusterSingleUnitKPSSStationaritySummary>
@@
 %<*ClusterSingleUnitTrialIndexLRSummary>
-The cluster spike times of either 0, 1 or 2 of the units in a cluster pair were correlated with trial index (Fisher's method: $\chi^{2}(656, N = 328) = 2551.6, $ $p < .001$; Fig. 4b,c).
+The cluster spike times of either 0, 1 or 2 of the units in a cluster pair were correlated with trial index (Fisher's method: $\chi^{2}(648, N = 324) = 2485.1, $ $p < .001$; Fig. 4b,c).
 %</ClusterSingleUnitTrialIndexLRSummary>
@@
 %<*SelectivelyDifferencedSummary>
-In total, 87 clusters had differencing applied to atleast one unit.
-87 of these clusters fulfilled the criteria, giving a total of 87 criteria fulfilling angled clusters (including the undifferenced stationary clusters).
+In total, 83 clusters had differencing applied to atleast one unit.
+83 of these clusters fulfilled the criteria, giving a total of 83 criteria fulfilling angled clusters (including the undifferenced stationary clusters).
 %</SelectivelyDifferencedSummary>
@@
 %<*SelectivelyDifferencedCorrelatedSummary>
-In total, 59 of the 87 criteria fulfilling clusters were linearly correlated following differencing (p-value < 0.05) (Fig.
+In total, 53 of the 83 criteria fulfilling clusters were linearly correlated following differencing (p-value < 0.05) (Fig.
 6h).
-26 of the correlated criteria fulfilling clusters had estimated angles significantly greater than $0^{\circ}$ (p-value < 0.025) and less than $45^{\circ}$ (p-value < 0.025).
-Of these, normality was not rejected for 15 clusters (Henze-Zirkler p > 0.05).
+23 of the correlated criteria fulfilling clusters had estimated angles significantly greater than $0^{\circ}$ (p-value < 0.025) and less than $45^{\circ}$ (p-value < 0.025).
+Of these, normality was not rejected for 13 clusters (Henze-Zirkler p > 0.05).
 Fig.
@@
 %<*SelectivelyDifferencedTestsPassedNewSummary>
-In total, 87 clusters had differencing applied to atleast one unit.
-41 of these were correlated and fulfilled the criteria.
-26 of these clusters had estimated angles significantly greater than $0^{\circ}$ (p-value < 0.025; Fig 6b) and less than $45^{\circ}$ (p-value < 0.025; Fig 6b), of which 14 normality was not rejected for (Henze-Zirkler p < 0.05).
+In total, 83 clusters had differencing applied to atleast one unit.
+35 of these were correlated and fulfilled the criteria.
+23 of these clusters had estimated angles significantly greater than $0^{\circ}$ (p-value < 0.025; Fig 6b) and less than $45^{\circ}$ (p-value < 0.025; Fig 6b), of which 12 normality was not rejected for (Henze-Zirkler p < 0.05).
 These angles are similar to the original angles before differencing (Fig.
@@
 %<*SelectivelyDifferencedTestsPassedEigenvalueSummary>
-The factor analysis model from above can be applied to the 41 correlated criteria fulfilling differenced clusters, as the factor correlation matrices of each cluster only had one eigenvalue greater than 1.
+The factor analysis model from above can be applied to the 35 correlated criteria fulfilling differenced clusters, as the factor correlation matrices of each cluster only had one eigenvalue greater than 1.
 %</SelectivelyDifferencedTestsPassedEigenvalueSummary>
@@
 %<*SelectivelyDifferencedBoxJenkinsDifferencedSummary>
-85 of the 87 differenced clusters had an ARMA model applied to atleast one unit.
+81 of the 83 differenced clusters had an ARMA model applied to atleast one unit.
 Almost all ARMA models fit for differenced clusters were of the MA type (Fig 7a), as expected for the negative autocorrelations introduced by differencing.
@@
 %<*SelectivelyDifferencedBoxJenkinsCorrelatedSummary>
-After the described application of the Box-Jenkins method, there were 44 criteria fulfilling clusters.
+After the described application of the Box-Jenkins method, there were 42 criteria fulfilling clusters.
 Normality was not rejected for 18 of these clusters (Henze-Zirkler p > 0.05).
-26 of the correlated criteria fulfilling clusters had estimated angles significantly greater than $0^{\circ}$ (p-value < 0.025; Fig.
+27 of the correlated criteria fulfilling clusters had estimated angles significantly greater than $0^{\circ}$ (p-value < 0.025; Fig.
 7b) and less than $45^{\circ}$ (p-value < 0.025; Fig.
-7b), for which 9 normality was not rejected for (Henze-Zeckler p-value > 0.05).
+7b), for which 10 normality was not rejected for (Henze-Zeckler p-value > 0.05).
 Fig.
@@
 %<*SelectivelyDifferencedBoxJenkinsTestsPassedEigenvalueSummary>
-The factor analysis model from above can be applied to the 44 correlated criteria fulfilling clusters, as the factor correlation matrices of each cluster only had one eigenvalue greater than 1. 
+The factor analysis model from above can be applied to the 42 correlated criteria fulfilling clusters, as the factor correlation matrices of each cluster only had one eigenvalue greater than 1. 
 %</SelectivelyDifferencedBoxJenkinsTestsPassedEigenvalueSummary>

### 2020-08-10 manuscript.tex
Lunch!

@@
 
-\subsection{Notes}
-I have begun developing several techniques to address this but I think it may prove difficult to show anything strongly conclusive either way.
+% \subsection{Notes}
+% I have begun developing several techniques to address this but I think it may prove difficult to show anything strongly conclusive either way.
 I can give more details of this once we are agreed on the overall objectives.
@@
 In summary, extracellular recordings were made in 15 anaesthetised Wistar rats using 16 or 128 electrode silicon probes, inserted perpendicularly into the barrel cortex.
-Single whisker deflections were made at various frequencies (0.1-10Hz) in blocks, each containing typically 200 trials (sweeps) with an intertrial interval of 5 s (occasionally 10, 15, or 30 s).
+Single whisker deflections were made at various frequencies (0.066-10Hz) in blocks, each containing typically 200 trials (sweeps) with an inter trial interval of 5s (occasionally 10, 15, or 30 s).
 This was repeated for each of the selected target whiskers (maximum 4).
@@
 
-This section demonstrates that the responses of some neurons exhibit depressive latency adaptation, which should be controlled for when testing for correlative structure caused by a non-adaptive state.
+This section demonstrates that the responses of some neurons in all neuron-groups exhibit latency adaptation, which should be controlled for when testing for correlative structure caused by a non-adaptive state.
 Figure 2d shows the first spike times of 5 single units in response to repeated stimulations.
@@
 
-This section presents an example first spike pair response distribution for the purposes of understanding.
-Fig.
-3a shows the distribution of first spike pairs in response to a 4Hz stimulation of a single whisker for two units in L5AI and L4I of the corresponding principle column (over a 50ms window following cortical activation onset).
+For the purpose of understanding, this section presents an example distribution of first spike pairs for a pair of units (L5AI and L4I) in the principle column in response to a 4Hz stimulation of the corresponding principle whisker (over a 50ms window following cortical activation onset).
 Only responses on `conjunctive trials' (trials in which both units spiked for the single stimulus) are shown.
-
 The orange Gaussian is centered at the mean of the first spike pairs and has widths equal to 4 times the first spike time standard deviations of each unit (on conjunctive trial), illustrating the region from which 99.9\% of first spike pairs would be predicted to fall under the assumption of independence and normality (i.e.
@@
 This assumption fails to capture the distribution structure of this and many other spike pair distributions (Supplementary Vid.
-1) and very few different stimuli to be encoded with high-discriminability.
-
-The spike time differences have a standard deviation of X (Fig. 3b), which is also poor for discriminable encoding.
-
-First spike pairs in Fig.
-3a appear to be drawn from one of two clusters (blue and green dots) in response to the same stimulus.
-The green cluster appears correlated with non-unitary gradient, suggestive of an underlying factor, which covaries first spike times, such that distributions of differences would be smaller when factor-conditioned, offering a higher power for discriminable encoding.
-
-\subsection*{Clustering algorithm summary}
+1).
+% Moreover, very few different stimuli could be encoded with high-discriminability.
+
+% The spike time differences have a standard deviation of X (Fig. 3b), which is also poor for discriminable encoding.
+
+Instead, first spike pairs appear to come from one of two clusters (blue and green dots).
+Moveover, the green cluster appears correlated with non-unitary gradient, suggestive of some underlying continuous state, which covaries the first spike times of the two units from trial-to-trial.
+Moreover, the apparent non-unitary gradient of correlation suggests that the distribution of differences varies with respect to the value of this underlying state, such that the state-conditioned distribution of relative first spike differences would be more compact than the general case, thus offering a higher potential capacity for stimulus encoding if downstream neurons could respond in a state-dependent manner.
+
+% \subsection*{Clustering algorithm summary}
+
+\subsection*{Stage 1 Clusters - Correlation analysis}
+
 A two-stage algorithm was developed to quantify the occurrence and angle of correlated clusters.
-Stage 1 extracted well isolated \textit{`flat clusters'} defined by the $4\sigma$ boundary of flat 2D Gaussian ellipses (flat solid green ellipse in Fig.
-3a; see Methods \ref{methods:flat_clusters} for details).
+Stage 1 of the algorithm extracted well isolated clusters defined by the $4\sigma$ boundary of flat 2D Gaussian ellipses (see Methods \ref{methods:flat_clusters}; e.g.
+flat solid green ellipse in Fig.
+3a).
+\ExecuteMetaData[AnalysisOutputLatex.tex]{ClusterCorrelationSummary0}
+The algorithm was verified by confirming that the algorithm extracted correlated clusters at chance levels for multiple control data sets (see Methods \ref{methods:correlation_surrogate_control_datasets}; Supplementary Fig.
+3).
+
 % The algorithm applied the unsupervised DBSCAN clustering method to first spike response distributions for a range of values of the DBSCAN EPS free parameter.
@@
 Points in this bounding ellipse were assigned to the `flat cluster'.
-To assess the occurrence of correlated clusters without sample bias, repeated and similar flat clusters produced by non-identical DBSCAN clusterings were left in the final group of flat clusters (Methods \ref{methods:flat_clusters}).
-When the algorithm was verified on surrogate control datasets (see Methods \ref{methods:correlation_surrogate_control_datasets}) correlated clusters were found at chance levels (Supplementary Fig.
-3).
-
-Stage 2 of the clustering algorithm aimed to better estimate the shape of the distributions underlying correlated flat clusters (p-value < 0.005), by estimating an `angled cluster' ellipse (angled solid green ellipse in Fig.
-3a; see Methods \ref{methods:angled_clusters}).
-Repeats, similar and non-isolated angled clusters were discarded.
-\ExecuteMetaData[100msAnalysisOutputLatex.tex]{ClusterTimeSpanSummary} \ExecuteMetaData[AnalysisOutputLatex.tex]{AngledClustersUniquePairsSummary}
-X\%\todo[color=red!40]{Fill} of angled clusters contained spikes from all conjunctive trials (Fig. 3d).
-
-\subsection*{Correlation analysis}
-\ExecuteMetaData[AnalysisOutputLatex.tex]{ClusterCorrelationSummary1}
-\ExecuteMetaData[AnalysisOutputLatex.tex]{ClusterCorrelationSummary2}
+% To assess the occurrence of correlated clusters without sample bias, repeated and similar flat clusters produced by non-identical DBSCAN clusterings were left in the final group of flat clusters (Methods \ref{methods:flat_clusters}).
 
@@
 
-\subsection*{Cluster angle}
+\subsection*{Stage 2 Clusters - Angle analysis}
 \label{results:cluster_angle}
+
+To estimate the correlation angle of the correlated Stage 1 clusters, a corresponding `Stage 2' \textit{angled} $4\sigma$ ellipse was estimated for each correlated Stage 1 cluster (see Methods \ref{methods:angled_clusters}; e.g.
+angled solid green ellipse in Fig 3a).
+Stage 2 clusters which were non-isolated, repeats or contained smaller clusters were discarded.
+\ExecuteMetaData[100msAnalysisOutputLatex.tex]{ClusterTimeSpanSummary}
+
+\ExecuteMetaData[AnalysisOutputLatex.tex]{AngledClustersUniquePairsSummary}
+\ExecuteMetaData[AnalysisOutputLatex.tex]{ClusterCorrelationSummary2}
 
@@
 
-An algorithm was developed to estimate a mean angle and confidence interval for each angled cluster (see Methods \ref{methods:cluster_angle_confidence_interval} for details), to test whether the angles of angled clusters were significantly different from $45^{\circ}$ (possibly indicative of factor or trial-dependent spike time differences).
-The mean angles, $\theta$, of positively correlated angled clusters (between $0^{\circ}$ and $90^{\circ}$) were of interest.
-$\theta_{45} = min(\theta, 90-\theta)$ were calculated for analysis, as only the ordering of the units in the pair determined whether $\theta$ was above or below $45^{\circ}$.
-
+% X\%\todo[color=red!40]{Fill} of angled clusters contained spikes from all conjunctive trials (Fig. 3d).
+
+To test whether the cluster correlation angles were different from $45^{\circ}$, an algorithm was developed to estimate the angle $\theta$ of correlation for each Stage 2 cluster (see Methods \ref{methods:cluster_angle_confidence_interval}).
+A final $\theta_{45} = min(\theta, 90-\theta)$ angle was used for each Stage 2 cluster, as only the ordering of the units determined whether $\theta$ was above or below $45^{\circ}$.
+
+\ExecuteMetaData[AnalysisOutputLatex.tex]{ClusterCorrelationSummary1}
+Fig. 4a shows the estimated $\theta_{45}$ angles (with 95\% confidence intervals) for all positively correlated Stage 2 clusters.
 \ExecuteMetaData[AnalysisOutputLatex.tex]{OriginalAnglesSummary}
-The algorithm was tested on a surrogate control dataset to validate that the detection of such differences from $45^{\circ}$ were not artifacts of the algorithm (see Methods \ref{methods:angle_surrogate_control_dataset}; Supplementary Fig.
+This initial finding that shows that the distribution of relative first spike differences between two units depends on some some underlying processes.
+Following sections will aim to determine when neural adaptation and/or non-adaptative underlying states underlie such positive non-unitary correlations. 
+To validate that such differences from $45^{\circ}$ were not artifacts of the algorithm, the algorithm was tested on a surrogate control dataset (see Methods \ref{methods:angle_surrogate_control_dataset}; Supplementary Fig.
 6).
-The following subsections aim to determine when trial number and/or another factor underlie such non-unitary covariation.
+
+The non-unitary correlations cannot be explained by a model in which the second spiking neuron spikes a fixed time difference after the first. 
+\ExecuteMetaData[AnalysisOutputLatex.tex]{ClusterCorrelationSummary0p1}
+This strengthens the case that the spike timing correlations between pairs of neurons are co-modulated by some underlying processes.
 

### 2020-08-10 AnalysisOutputLatex.tex
End of day!

@@
 3h).
-Moreover, 72.8\% (118/162) of the original response distributions from which Stage 2 clusters were extracted were not correlated significantly (p>0.005) (Fig.
+Moreover, 59.9\% (97/162) of the original response distributions from which Stage 2 clusters were extracted were not correlated significantly (p>0.05) (Fig.
 3h).
@@
 %<*OriginalAnglesSummary>
-67.9\% (110/162) of $	heta_{45}$ angles were both significantly less than $45^{\circ}$ (p<0.025) and greater than $0^{\circ}$ (<0.025). 
+67.9\% (110/162) of $\theta_{45}$ angles were both significantly less than $45^{\circ}$ (p<0.025) and greater than $0^{\circ}$ (p<0.025). 
 %</OriginalAnglesSummary>
@@
 %<*OriginalTestsPassedSummary>
-Of the 162 angled clusters, 46 were determined to be stationary and sufficiently correlated by fulfilling the criteria described in the Methods \ref{methods:angle_and_factor_analysis_criteria}, such that an underlying factor could underlie the covariation.
+46 Stage 2 clusters were determined as stationary and correlated (see Methods \ref{methods:angle_and_factor_analysis_criteria} for criteria).
 %</OriginalTestsPassedSummary>
@@
 %<*OriginalTestsPassedAngleSummary>
-Of the 46 criteria fulfilling angled clusters, 30 had estimated angles significantly different from both $0^{\circ}$ (p<0.025; Fig.
-4f) and $45^{\circ}$ (p<0.025; Fig.
-4f).
-Of the 46 criteria fulfilling angled clusters, normality was not rejected for 20 clusters (Henze-Zirkler p > 0.05), of which 14 had an estimated angle significantly different from both $0^{\circ}$ (p-value < 0.025) and $45^{\circ}$ (p-value < 0.025).
+30 stationary Stage 2 clusters had estimated angles significantly different from both $45^{\circ}$ (p<0.025; Fig. 4f) and $0^{\circ}$ (p<0.025; Fig. 4f). 
 %</OriginalTestsPassedAngleSummary>
 
+%<*OriginalTestsPassedAndNormalAngleSummary>
+Normality was not rejected for 20 stationary Stage 2 clusters (Henze-Zirkler p > 0.05), of which 14 had an estimated angle significantly different from both $45^{\circ}$ (p-value < 0.025) and $0^{\circ}$ (p-value < 0.025).
+%</OriginalTestsPassedAndNormalAngleSummary>
+
 %<*OriginalTestsPassedFactorAnalysisSummary>
-Factor analysis could be applied to the 46 criteria fulfilling angled clusters, to give an estimate of the factor-conditioned cluster response distributions for the stationary angled clusters.
+As each Stage 2 cluster fulfilled the criteria of Methods \ref{methods:angle_and_factor_analysis_criteria}, factor analysis could be applied to the stationary Stage 2 clusters to give an estimate of state-conditioned response distributions for each cluster.
 %</OriginalTestsPassedFactorAnalysisSummary>

### 2020-08-10 manuscript.tex
End of day!

@@
 
-To estimate the correlation angle of the correlated Stage 1 clusters, a corresponding `Stage 2' \textit{angled} $4\sigma$ ellipse was estimated for each correlated Stage 1 cluster (see Methods \ref{methods:angled_clusters}; e.g.
+To better estimate the distribution underlying each correlated Stage 1 cluster and to estimate the cluster correlation angle, an \textit{angled} `Stage 2' ellipse was estimated for each Stage 1 cluster using a multi-stage bootstrap algorithm (see Methods \ref{methods:angled_clusters}; e.g.
 angled solid green ellipse in Fig 3a).
@@
 
-To test whether the cluster correlation angles were different from $45^{\circ}$, an algorithm was developed to estimate the angle $\theta$ of correlation for each Stage 2 cluster (see Methods \ref{methods:cluster_angle_confidence_interval}).
+To test whether the cluster correlation angles were different from $45^{\circ}$, the angle $\theta$ of correlation for each Stage 2 cluster was calculated (see Methods \ref{methods:cluster_angle_confidence_interval}).
 A final $\theta_{45} = min(\theta, 90-\theta)$ angle was used for each Stage 2 cluster, as only the ordering of the units determined whether $\theta$ was above or below $45^{\circ}$.
@@
 \ExecuteMetaData[AnalysisOutputLatex.tex]{OriginalAnglesSummary}
-This initial finding that shows that the distribution of relative first spike differences between two units depends on some some underlying processes.
-Following sections will aim to determine when neural adaptation and/or non-adaptative underlying states underlie such positive non-unitary correlations. 
-To validate that such differences from $45^{\circ}$ were not artifacts of the algorithm, the algorithm was tested on a surrogate control dataset (see Methods \ref{methods:angle_surrogate_control_dataset}; Supplementary Fig.
+This initial finding shows that for some pairs of units the distribution of relative first spike differences between two units depends on some underlying processes.
+Following sections aim to determine when neural adaptation and/or non-adaptative underlying states underlie such positive non-unitary correlations. 
+To validate that such differences from $45^{\circ}$ were not artifacts of the Stage 2 cluster estimation algorithm, the algorithm was tested on a surrogate control dataset (see Methods \ref{methods:angle_surrogate_control_dataset}; Supplementary Fig.
 6).
@@
 \ExecuteMetaData[AnalysisOutputLatex.tex]{ClusterCorrelationSummary0p1}
-This strengthens the case that the spike timing correlations between pairs of neurons are co-modulated by some underlying processes.
-
-\subsection*{Cluster stationarity and normality}
-
+This strengthens the case that the spike timing correlations between pairs of neurons are co-modulated by some underlying process.
+
+\subsection*{Cluster stationarity}
+
+This section tests both whether cluster spike times for individual units are affected by neural adaptation and also the degree to which neural adaptation explains cluster correlations.
 \ExecuteMetaData[AnalysisOutputLatex.tex]{ClusterSingleUnitTrialIndexLRSummary}
-Only a small amount of variance was explained by the linear models fit to predict the joint spike pair times from trial number (Fig.
-4d), suggesting that neural adaptation may not alone explain the covarying cluster spike times.
+\ExecuteMetaData[AnalysisOutputLatex.tex]{ClusterSingleUnitKPSSStationaritySummary}
+Only a small amount of variance was explained by the linear models fit to predict the joint spike pair times from trial number (Fig. 4d).
+
+% supporting that neural adaptation does not alone explain cluster correlations. 
 % More complex first spike time relationships with trial index were also observed. 
-\ExecuteMetaData[AnalysisOutputLatex.tex]{ClusterSingleUnitKPSSStationaritySummary}
-
-Beside general adaptive trends, some units exhibited observable short trial-by-trial sequences (3-4 trials) of highly correlated first spike times, before losing such correlations.
-Lagged autocorrelations were significantly above chance for a high number of lags (Supplementary Fig.
-4) supportive of general adaptive trends.
-Only the lag-2 partial autocorrelation was significantly above chance (p-value < 0.005; Supplementary Fig.
-4), representative of the additional trial-by-trial structure observed for very few single units.
-
-The cluster spike times of cluster unit pairs exhibit lagged cross-correlations above chance for lags 1-5 (Supplementary Fig.
-5), as expected for general adaptive trends.
-Only the lag-2 partial cross-correlation was above chance (Supplementary Fig.
-5).
-This suggests that the individual adaptive trends of cluster pairs explains the majority of above chance cross-correlations, whilst there may be some finer sequential structure between a few unit pairs.
+Lag-1 autocorrelations of single unit cluster spike times were above chance [STAT], as expected by the spike times of some units being affected by adaptation.
+Beside general adaptive trends, some single units exhibited short trial-by-trial sequences (3-4 trials) of autocorrelated first spike times, before losing such autocorrelations. 
+% Lagged autocorrelations were significantly above chance for a high number of lags (Supplementary Fig. 4) supportive of general adaptive trends. 
+ Only the lag-2 partial autocorrelation (lags 2-5 tested) was significantly above chance (p-value < 0.005; Supplementary Fig.
+4), representative of the minimal and temporary autocorrelations observed for several single units.
+
+Cluster spike times for cluster pairs showed above chance lag-1 cross-correlations, which could be explained by clusters for which the spike times of both units were modulated by adaptation.
+Only the lag-2 partial cross-correlation was slightly above chance (lags 2-5 tested) (Supplementary Fig.
+5), suggestive of minimal lagged cross-correlative structure between cluster spike times for some unit pairs after neural adaptation.
 
@@
 
+This section demonstrated heterogeneity in the non-stationarity of Stage 2 clusters.
+Particularly, either the spike times of both units were not affected by adaptation (i.e.
+stationary clusters) or one or both of the units have cluster spike times modulated by adaptation (i.e.
+non-stationary clusters).
+These two cases must be treated differently when testing for adaptation-independent noise correlations.
+
 \subsection*{Analysis of stationary clusters}
+This section characterises the occurrence, angle and potential state-dependent encoding capacity of stationary Stage 2 clusters.
+% Stationary clusters, for which the spike times of neither unit are modulated by neural adaptation must be modulated by some non-adaptative underlying continuous state. 
 \ExecuteMetaData[AnalysisOutputLatex.tex]{OriginalTestsPassedSummary}
 \ExecuteMetaData[AnalysisOutputLatex.tex]{OriginalTestsPassedAngleSummary}
-Fig.
-4g shows the histogram of variance explained by the stationary angled cluster correlations.
+Fig. 4g shows the proportion of variance explained by the stationary Stage 2 cluster correlations.
+\ExecuteMetaData[AnalysisOutputLatex.tex]{OriginalTestsPassedAndNormalAngleSummary}
+
  Autocorrelations, cross-correlations, partial autocorrelations and partial cross-correlations were at chance levels for the stationary clusters (Supplementary Fig.
-7).
+7), confirming that the stationary Stage 2 cluster correlations are not caused by neural adaptative trends.
 
@@
 
-\ExecuteMetaData[AnalysisOutputLatex.tex]{OriginalTestsPassedFactorAnalysisSummary} The following sections apply time series analysis techniques to remove the neural adaptive trends of the non-stationary units of the remaining clusters.
-The factor correlation matrix of all criteria fulfilling angled cluster had only one eigenvalue greater than 1, suggesting that a single factor underlied the pairwise cluster covariations.
-The following single factor model was therefore fit to the elements of each (zero-centered) cluster: $s_i = \lambda_i \eta + \sigma_i$ for $i \in [0, 1]$
+The factor correlation matrix of each stationary Stage 2 clusters had only one eigenvalue greater than 1, suggesting that a single factor underlied each cluster correlation.
+\ExecuteMetaData[AnalysisOutputLatex.tex]{OriginalTestsPassedFactorAnalysisSummary} 
+% The following sections apply time series analysis techniques to remove the neural adaptive trends of the non-stationary units of the remaining clusters.
+
+The following single factor model was fit to the elements of each (zero-centered) cluster: $s_i = \lambda_i \eta + \sigma_i$ for $i \in [0, 1]$ 
 % and $s_1 = \lambda_1 \eta + \sigma_1$, 
@@
 $\lambda_i$ is the gradient at which the spike time of unit $i$ varies w.r.t. the state $\eta$, 
-and $\sigma_i$ is the predicted standard deviation of $s_i$ given $\eta$. Bootstrapping (1000 iterations) was used to calculate means of these values, which were used in subsequent analyses.
+and $\sigma_i$ is the predicted standard deviation of $s_i$ given $\eta$. Bootstrapping (1000 iterations) was used to calculate mean estimates of these values.
 
@@
 5a shows the values predicted by the model for $\eta = [-4, 4]$ and $\sigma_i = 0$, re-centered around the cluster mean.
-$\theta_{45}$ angles of the factor lines were estimated using the same method used for PCA component angle estimation (Methods \ref{methods:cluster_angle_confidence_interval}) and were similar to the corresponding PCA $\theta_{45}$ angles (Fig.
-5b).
+$\theta_{45}$ angles of these factor anaylsis lines were also estimated
+% using the same method used for PCA component angle estimation 
+(see Methods \ref{methods:cluster_angle_confidence_interval}) and were similar to the initial Stage 2 $\theta_{45}$ angles (Fig.
+5b)\todo{Candidate to move to Supp Figs}.
 The purple ellipse in Fig.
-5a has widths equal to $4\sigma_0, 4\sigma_1$ where  $\sigma_0, \sigma_1$ are the standard deviations of the predicted factor ($\eta$) dependent response distribution.
-This allows visual comparison of the estimated factor-conditioned response distribution to the larger $4\sigma$ areas under the assumptions of normality and independence (orange), normal and independent clusters (green flat cluster) and normal correlated clusters (green angled cluster).
-
-The areas of the two-dimensional (i.e.
-two unit) factor dependent response distributions show a striking improvement in potential encoding power (Supplementary Vid.
+5a has widths equal to $4\sigma_0, 4\sigma_1$ where  $\sigma_0, \sigma_1$ are the standard deviations of the predicted state ($\eta$) dependent response distribution.
+
+The state-conditioned response distributions estimated by factor analysis allow visual (purple ellipses in Fig.
+5a and top left of Supplementary Vid.
+1) and quantitative (Fig.
+5c,d,e; Supplementary Fig.
+8) comparison of the estimated state-conditioned response distribution to the larger $4\sigma$ areas under fixed-state decoder assumptions of normality and independence (orange), normal and independent clusters (green flat cluster) and normal and correlated clusters (green angled cluster).
+The predicted state-conditioned response distributions are strikingly smaller than those predicted under the fixed-state assumptions, thus potentially offering an improved encoding capacity.
+
+The factor dependent predicted standard deviations, $\sigma$, are less than 1 for X\% of units and less than 2 for Y\% of units\todo[color=red!40]{Fill}. 
+As the factor analysis model assumes that noise for each unit given $\eta$ is normally distributed and independent between units, the state-dependent difference standard deviation predicted by the model is equal to \(\sigma_{diff} = \sqrt{\sigma_{0}^2 + \sigma_{1}^2}\).
+The estimated $\sigma_{diff}$ values are strikingly lower than the original standard deviation of differences (Fig.
+5e).
+
+% The areas of the two-dimensional (i.e.
+two unit) state-dependent response distributions show a striking improvement in potential encoding power (Supplementary Vid.
 1; Supplementary Fig.
 8).
-The standard deviations $\sigma_i$ of the factor analysis predicted response distributions for units $i = 0,1$ separately are also strikingly smaller than those assumed by the simpler assumptions (Fig.
+The standard deviations $\sigma_i$ of the factor analysis predicted state-conditioned response distributions for units $i = 0,1$ separately are also strikingly smaller than those assumed by the simpler assumptions (Fig.
 5c,d).
-The factor dependent predicted standard deviations, $\sigma$, are less than 1 for X\% of units and less than 2 for Y\% of units\todo[color=red!40]{Fill}.
 % This gives an estimate of potential encoding power for single units, given by the factor dependent response distributions. 
 % It is necessary to consider the results in terms of relative spike timing differences, which can be explicitly processed by downstream neurons.
-As the factor analysis model assumes that noise for each unit given $\eta$ is normally distributed and independent between units, the state-dependent difference standard deviation predicted by the model is equal to \(\sigma_{diff} = \sqrt{\sigma_{0}^2 + \sigma_{1}^2}\).
+
 % , where  $\sigma_0, \sigma_1$ are the standard deviations predicted by factor analysis given state $\eta$.
-The estimated $\sigma_{diff}$ values are strikingly lower than the original standard deviation of differences (Fig. 5e).
 
@@
 
+[UP TO HERE]
+
 To test whether a shared state could underlie covariation in the remaining non-stationary angled clusters, the non-stationary single unit cluster spike times were differenced and rescaled (a standard time-series analysis technique) to remove neural adaptive trends by subtracting the spike time on trial t by the spike time on trial t-1.

### 2020-08-11 AnalysisOutputLatex.tex
Great!

@@
 %<*OriginalTestsPassedSummary>
-46 Stage 2 clusters were determined as stationary and correlated (see Methods \ref{methods:angle_and_factor_analysis_criteria} for criteria).
+46 Stage 2 clusters were determined as stationary (see Methods \ref{methods:angle_and_factor_analysis_criteria} for criteria).
 %</OriginalTestsPassedSummary>
@@
 %<*OriginalTestsPassedFactorAnalysisSummary>
-As each Stage 2 cluster fulfilled the criteria of Methods \ref{methods:angle_and_factor_analysis_criteria}, factor analysis could be applied to the stationary Stage 2 clusters to give an estimate of state-conditioned response distributions for each cluster.
+As each Stage 2 cluster also fulfilled the criteria of Methods \ref{methods:angle_and_factor_analysis_criteria}, factor analysis could be applied to the stationary Stage 2 clusters to give an estimate of state-conditioned response distributions for each cluster.
 %</OriginalTestsPassedFactorAnalysisSummary>

### 2020-08-11 manuscript.tex
Great!

@@
 (d) First spike rasters following cortical onset for five unit stimulus combinations.
-(e) P-values of linear regressors fit to predict first spike time from trial index for all stimulus and unit combinations.
-(f) $R$ (green) and $R^2$ (blue) of linear regressors fit to predict first spike time from trial index for all stimulus and unit combinations.
-(g) First spike time KPSS p-values for all stimulus and unit combinations.
-Implementation of p-value estimation rounds all p-values $>0.1$ to 0.1.}
-\label{fig:universe}
-\end{figure}
-
-\subsection*{Single spike patterns}
-
-\todo{Add PSTH of single stimulus. i.e. 1Hz over 10 and all stimuli maybe}
-\todo{Do conclusion}
-
-This section justifies the analysis of first spike patterns as a cortical representation of single whisker deflections.
-Firstly, this section demonstrates that the majority of neurons respond to single whisker deflections with either 0 or 1 spike (Fig.
-2a).
-Spike count encoding therefore has a low capacity for discriminable stimulus encoding, made worse by high spike count variance (Fig.
-2b).
-The responses of each unit-type and layer group (averaged over stimuli) demonstrate a clear initial response, followed by a tailing-off (Fig.
-2c).
-The paper aims to better characterise this initial (largely single spike) response.
-
-\subsection*{Spike latency adaptation}
-
-\todo{Look at adaptation by neuron type and layer}
-\todo{Select final adaptation examples, with layers and stats.}
-\todo{Do adaptation conclusion}
-\todo{Make adaptation stats overview paragraph simpler, with nice single bar chart}
-
-This section demonstrates that the responses of some neurons in all neuron-groups exhibit latency adaptation, which should be controlled for when testing for correlative structure caused by a non-adaptive state.
-Figure 2d shows the first spike times of 5 single units in response to repeated stimulations.
 The L23E unit of Fig.
@@
 2d(iv, v) demonstrate more complex response-distribution structure and possible non-linear non-stationarity.
-
-Single unit first spike times are linearly correlated with trial number above chance (Fisher's method: $\chi^{2} (1786, N = 893) = 3672.69, p<.001$; Fig. 2e).
+(e) P-values of linear regressors fit to predict first spike time from trial index for all stimulus and unit combinations.
+(f) $R$ (green) and $R^2$ (blue) of linear regressors fit to predict first spike time from trial index for all stimulus and unit combinations.
+(g) First spike time KPSS p-values for all stimulus and unit combinations.
+Implementation of p-value estimation rounds all p-values $>0.1$ to 0.1.}
+\label{fig:universe}
+\end{figure}
+
+\subsection*{Single spike patterns}
+
+This section presents evidence that supports the analysis of first spike patterns as a cortical representation of single whisker deflections.
+Firstly, the majority of neurons respond to single whisker deflections with either 0 or 1 spike (Fig.
+2a) with high spike count variance (Fig.
+2b).
+Spike count encoding therefore has a low capacity for discriminable stimulus encoding.
+The responses of each unit-type and layer group (averaged over stimuli) demonstrate a clear initial response, followed by a tailing-off (Fig.
+2c).
+The paper aims to better characterise multi-neuron single spike patterns within this initial response.
+
+\subsection*{Spike latency adaptation}
+
+\todo{Select final adaptation examples, with layers and stats.}
+\todo{Do adaptation conclusion}
+
+This section demonstrates that a significant proportion of units exhibit spike latency adaptation, which must be controlled for when testing for non-adpatative noise correlations.
+In response to repetitions of single stimuli, single unit first spike time adaptative trends are heterogeneous (Fig 2d), with many units showing no adaptation.
+For repeated trials of single stimuli, the first spike times of an above-chance proportion of single units in 7 out of 8 unit groups are linearly correlated with trial index and are KPSS non-stationary (Fig.
+2e).
+$90\%$ (57/63) of these linear correlations with trial index were positive (Fig.
+2f), supporting depressive latency adaptation \cite{ahissar2000transformation, ahissar2001temporal, kheradpezhouh2017response}.\todo{Automate depressive latency sentence} Only the lag-2 partial autocorrelation was slightly above chance (lags 2-5 tested, Supplementary Fig.
+2), suggestive of additional structure in the trial-to-trial sequence of first spike times for a small number of units.
+These findings show that tests for non-adpatative noise correlations must control for latency adaptation.
+
+% first spike times of single units show largely heterogeneous relationships with trial index over repeated
+
+% Single unit first spike time relationships with trial index in response to repeated stimulations are highly heterogeneous (Fig 2d)
+
+% Figure 2d shows the first spike times of 5 single units in response to repeated stimulations.
+
+% (Fisher's method: $\chi^{2} (1786, N = 893) = 3672.69, p<.001$; Fig. 2e).
 % (Fisher's method: $\chi^{2} (1786, N = 893) = 3672.69, p = 1.027\textrm{e-132}$; Fig. 2e). 
-Of the 63 first spike response distributions that were significantly correlated with trial index (p<.005), $90\%$ (57/63) were positively correlated (Fig.
-2f), supporting depressive latency adaptation \cite{ahissar2000transformation, ahissar2001temporal, kheradpezhouh2017response}.
-The KPSS null hypothesis of stationarity was rejected for p-value < 0.05 for 19\% (168/893) first spike response distributions (Fig.
-2g).
-Autocorrelations of single-unit single-stimulus first spike times were above chance for lags 1-5 (Supplementary Fig.
-2), furthering support of general adaptive trends.
-Only the lag-2 partial autocorrelation was slightly above chance (Supplementary Fig.
-2), suggestive of additional structure in the trial-to-trial sequence of first spike times for a small number of units, beyond adaptive trends.
-% after accounting for general adaptive trends.
+%  The KPSS null hypothesis of stationarity was rejected for p-value < 0.05 for 19\% (168/893) first spike response distributions (Fig. 2g).
+
+% Autocorrelations of single-unit single-stimulus first spike times were above chance for lags 1-5 (Supplementary Fig. 2), furthering support of general adaptive trends.
 
@@
 
-\todo{Do introduction (objectives) and summary}
 \todo{Clear up final paragraph}
@@
 % Stationary clusters, for which the spike times of neither unit are modulated by neural adaptation must be modulated by some non-adaptative underlying continuous state. 
-\ExecuteMetaData[AnalysisOutputLatex.tex]{OriginalTestsPassedSummary}
+In total, \ExecuteMetaData[AnalysisOutputLatex.tex]{OriginalTestsPassedSummary}
 \ExecuteMetaData[AnalysisOutputLatex.tex]{OriginalTestsPassedAngleSummary}
@@
 
-The factor correlation matrix of each stationary Stage 2 clusters had only one eigenvalue greater than 1, suggesting that a single factor underlied each cluster correlation.
 \ExecuteMetaData[AnalysisOutputLatex.tex]{OriginalTestsPassedFactorAnalysisSummary} 
@@
 (6) Both units lag-1 autocorrelation p-value > 0.05 (this requirement was relaxed for the differenced clusters but not ARIMA clusters).
+(7) The factor correlation matrix had only one eigenvalue greater than 1, suggesting that a single factor underlied each cluster correlation.
 Bartlett spherecity tests that the correlation matrix is significantly different from the identity, and is a standard pre-test for factor analysis.

### 2020-08-11 manuscript.tex
Gut

@@
 % These population codes assume that trial-to-trial variability is either independent or correlated across the population \cite{averbeck2006neural, lin2015nature}. 
-Trial-to-trial spike count variability is in fact correlated across the population in early sensory cortices \cite{reyes2015high, lin2015nature} and was shown to be modulated by a two-dimensional continuous state \cite{lin2015nature}.
-It is possible that spike time variability is modulated across a population by the same or similar low-dimensional underlying state.
-This study aims to characterise the nature of trial-to-trial spike time correlations as a step towards confirming or rejecting the modulation of spike times by an underlying low-dimensional latent state.
+In a recent study in early visual cortex, trial-to-trial spike count variability was shown to be correlated across the population and modulated by an underlying low 2-dimensional continuous latent state \cite{lin2015nature}.
+Other studies have reported only weak spike count correlations in sensory cortices \cite{reyes2015high}.
+
+It is possible that spike time variability is caused and modulated across a population by a low-dimensional underlying state.
+This study aims to characterise the nature of trial-to-trial spike time noise correlations as a step towards confirming or rejecting the modulation of spike times by an underlying low-dimensional latent state.
 This offers a novel prediction of the encoding capacity of state-dependent spike time encoding.
+These results control for the occurrence of neural adaptation.
+The demonstrated nature of spike time noise correlations demonstrated offers an opportunity to  explain the report of only weak noise correlations in the previous characterisations of this data set \cite{reyes2015high}.
 Characterising spike timing codes is important to understanding a variety of dynamic, homeostatic and plastic processes, as well as properties of action potential patterns directly involved in cognition.
@@
 `Signal correlations' concern tuning similarity between neurons over multiple stimuli (see \cite{averbeck2006neural}).
-As discussed, previous studies have demonstrated that spike count noise correlations are modulated across a population  by low-dimensional continuous latent states \cite{lin2015nature}.
+Neural coding and noise correlations are classically considered with respect to fixed-state decoders \cite{averbeck2006neural, moreno2014information, stringer2019high}.
+For fixed-state decoders, noise correlations can either increase or decrease encoding capacity \cite{averbeck2006neural}.
+
+% As discussed, previous studies have demonstrated that spike count noise correlations are modulated across a population  by low-dimensional continuous latent states \cite{lin2015nature}. 
 % For example, a 2-dimensional continuous latent state modulates responses in V1 \cite{lin2015nature}.
 
-Neural coding and noise correlations are classically considered with respect to fixed-state decoders \cite{averbeck2006neural, moreno2014information, stringer2019high}.
 It is important to consider, that if downstream neurons are also modulated by the same population wide continuous latent state, then downstream `decoder' neurons could respond in a state-dependent manner.
@@
 % The stimulus-dependency of noise correlations is left to future studies.
-Characterising noise correlations within stimulus-response distributions is complicated by the heterogeneity of sample response distribution structure and neural adaptation, which occurs across species and modalities \cite{dragoi2002dynamics, ulanovsky2003processing, sharpee2006adaptive}.
+
+Regardless of whether decoding neurons in the brain respond in a fixed-state or state-dependent manner, characterising noise correlations within stimulus-response distributions is complicated by the heterogeneity of sample response distribution structure and neural adaptation, which occurs across species and modalities \cite{dragoi2002dynamics, ulanovsky2003processing, sharpee2006adaptive}.
 % That is, the response distributions are \textit{non-stationary} and change monotonically as a function of trial index for some neurons. 
@@
 The classical `stimulus-dependency' of noise correlations is left to future studies.
-A previous study analysing the same dataset found precise multi-neuron spike patterns that were intra- and inter-laminar, as well intra- and intra-columnar \cite{reyes2014laminar}.
-Another study analysing the data set found spike count noise correlations and could decode stimulus properties (stimulation frequency and location) from spike counts and spike patterns \cite{reyes2015high}.
+
+A previous study analysing the same data set found precise multi-neuron spike patterns that were intra- and inter-laminar, as well intra- and intra-columnar \cite{reyes2014laminar}.
+Another study analysing the data set could decode stimulus properties (stimulation frequency and location) from spike counts and spike patterns but found only weak spike count noise correlations \cite{reyes2015high}.
+The complex nature of spike time correlations uncovered in this study for the same data set may therefore offer an improved opportunity to study population-wide modulation by low-dimensional latent states.
 The new analysis uses time series analysis to control for neural adaptation over all stimulus trials.
-adaptation occurs in the barrel cortex to repetitive whisker stimulation for both spike count \cite{ahissar2000transformation, ahissar2001temporal, kheradpezhouh2017response, khatri2009stimulus, martin2014tactile, barros2019response} and spike latency \cite{ahissar2000transformation, ahissar2001temporal, kheradpezhouh2017response}, and was confirmed to occur for 7 out of 8 neuron groups in this study.
+Adaptation occurs in the barrel cortex to repetitive whisker stimulation for both spike count \cite{ahissar2000transformation, ahissar2001temporal, kheradpezhouh2017response, khatri2009stimulus, martin2014tactile, barros2019response} and spike latency \cite{ahissar2000transformation, ahissar2001temporal, kheradpezhouh2017response}, and was confirmed to occur for 7 out of 8 neuron groups in this study.
 
@@
 
-adaptation-controlled non-unitary correlated clusters appeared for pairs of neurons spatially separated over multiple barrels and cortical layers.
+Adaptation-controlled non-unitary correlated clusters appeared for pairs of neurons spatially separated over multiple barrels and cortical layers.
 This offers suggestive evidence of a shared low-dimensional continuous latent state affecting the spike times of spatially distanced neurons, which in turn could also modulate downstream neurons and allow them to respond in a state-dependent manner.
@@
 
-\subsection*{Example first spike pair response distribution}
+\subsection*{First spike pair response distributions}
 
@@
 
-\todo{Clear up final paragraph}
-\todo{Differences standard deviation}
-
+Analysing the joint responses of two or more units to repeated presentations of a single stimulus offers the opportunity to discern noise correlations caused by adaptative or non-adaptative processes.
 For the purpose of understanding, this section presents an example distribution of first spike pairs for a pair of units (L5AI and L4I) in the principle column in response to a 4Hz stimulation of the corresponding principle whisker (over a 50ms window following cortical activation onset).
 Only responses on `conjunctive trials' (trials in which both units spiked for the single stimulus) are shown.
-The orange Gaussian is centered at the mean of the first spike pairs and has widths equal to 4 times the first spike time standard deviations of each unit (on conjunctive trial), illustrating the region from which 99.9\% of first spike pairs would be predicted to fall under the assumption of independence and normality (i.e.
+
+First spike pairs appear to come from one of two clusters (blue and green dots) for presentations of the same stimulus.
+Moreover, the green cluster appears correlated with non-unitary gradient, suggestive that some underlying processes (adaptative or non-adaptative) may covary the first spike times of the two units.
+Such correlated clusters were observed in a large number of first spike pair distributions (Supplementary Vid.
+1).
+
+This paper aims to quantify the occurrence of such correlated clusters and determine when adaptative and/or non-adaptative processes underlie such noise correlations.
+Importantly, the apparent non-unitary gradient of correlation of the green cluster in Fig.
+3a suggests that the distribution of differences varies with respect to the possible adaptative and/or non-adaptative processes underlying the correlation.
+
+The orange Gaussian in Fig.
+3a is centered at the mean of the first spike pairs and has widths equal to 4 times the first spike time standard deviations of each unit (on conjunctive trial), illustrating the region from which 99.9\% of first spike pairs would be predicted to fall under the assumption of independence and normality (i.e.
 spike time jitter).
@@
 
-Instead, first spike pairs appear to come from one of two clusters (blue and green dots).
-Moveover, the green cluster appears correlated with non-unitary gradient, suggestive of some underlying continuous state, which covaries the first spike times of the two units from trial-to-trial.
 Moreover, the apparent non-unitary gradient of correlation suggests that the distribution of differences varies with respect to the value of this underlying state, such that the state-conditioned distribution of relative first spike differences would be more compact than the general case, thus offering a higher potential capacity for stimulus encoding if downstream neurons could respond in a state-dependent manner.

### 2020-08-11 manuscript.tex
Smashing!

@@
 
-\usepackage[colorinlistoftodos]{todonotes}
-
-% \usepackage[nomarkers,notablist,nofiglist]{endfloat}
-% \usepackage[nomarkers,notablist,nofiglist]{endfloat}
-% \usepackage[colorinlistoftodos,disable]{todonotes}
+% \usepackage[colorinlistoftodos]{todonotes}
+
+\usepackage[nomarkers,notablist,nofiglist]{endfloat}
+\usepackage[colorinlistoftodos,disable]{todonotes}
 
@@
 
-\begin{abstract}
-% http://blogs.nature.com/naturejobs/2014/11/03/how-to-get-published-in-high-impact-journals-big-research-and-better-writing/
-% \color{blue}`Do not undersell your research. “If you find something big, tell us!'\color{Black}
-
-REL: Efficient population coding of naturalistic whisker motion in the ventro-posterior medial thalamus based on precise spike timing
-
-Sensory stimulation induced patterns of action potentials recorded in adult rat cortex in vivo allow investigation of neural coding principles.
-
-Patterns of action potentials induced by sensory stimulation in vivo allow investigation of neural coding principles.
+% \begin{abstract}
+% % http://blogs.nature.com/naturejobs/2014/11/03/how-to-get-published-in-high-impact-journals-big-research-and-better-writing/
+% % \color{blue}`Do not undersell your research. “If you find something big, tell us!'\color{Black}
+
+% REL: Efficient population coding of naturalistic whisker motion in the ventro-posterior medial thalamus based on precise spike timing
+
+% Sensory stimulation induced patterns of action potentials recorded in adult rat cortex in vivo allow investigation of neural coding principles.
+
+% Patterns of action potentials induced by sensory stimulation in vivo allow investigation of neural coding principles.
 Trial-to-trial variability engenders debate over temporal and rate codes.
@@
 
-\end{abstract}
-
-\keywords{Neural coding, precise spike timing, time-warp, state-dependent processing, barrel cortex, noise correlations, neural adaptation, in vivo recording, feature-binding}
+% \end{abstract}
+
+% \keywords{Neural coding, precise spike timing, time-warp, state-dependent processing, barrel cortex, noise correlations, neural adaptation, in vivo recording, feature-binding}
 
@@
 
-It is possible that spike time variability is caused and modulated across a population by a low-dimensional underlying state.
-This study aims to characterise the nature of trial-to-trial spike time noise correlations as a step towards confirming or rejecting the modulation of spike times by an underlying low-dimensional latent state.
+It is possible that spike time variability is caused and modulated across a population by such a low-dimensional continuous latent state.
+This study aims to characterise the nature of trial-to-trial spike time noise correlations as a step towards confirming or rejecting the modulation of spike times by a continuous low-dimensional latent state.
 This offers a novel prediction of the encoding capacity of state-dependent spike time encoding.
-These results control for the occurrence of neural adaptation.
-The demonstrated nature of spike time noise correlations demonstrated offers an opportunity to  explain the report of only weak noise correlations in the previous characterisations of this data set \cite{reyes2015high}.
-Characterising spike timing codes is important to understanding a variety of dynamic, homeostatic and plastic processes, as well as properties of action potential patterns directly involved in cognition.
+These results control for neural adaptation.
+Characterising spike timing codes is important to understanding a variety of dynamic and plastic processes, including properties of action potential patterns directly involved in cognition.
 
@@
 
-Under such approaches, `noise correlations' refer to shared trial-to-trial variability between the responses of multiple neurons over repeated presentations of single stimulus \cite{averbeck2006neural}.
+Under such approaches, `noise correlations' refer to shared trial-to-trial variability between the responses of multiple neurons over repeated presentations of single stimuli \cite{averbeck2006neural}.
 `Signal correlations' concern tuning similarity between neurons over multiple stimuli (see \cite{averbeck2006neural}).
 Neural coding and noise correlations are classically considered with respect to fixed-state decoders \cite{averbeck2006neural, moreno2014information, stringer2019high}.
-For fixed-state decoders, noise correlations can either increase or decrease encoding capacity \cite{averbeck2006neural}.
 
@@
 
-Regardless of whether decoding neurons in the brain respond in a fixed-state or state-dependent manner, characterising noise correlations within stimulus-response distributions is complicated by the heterogeneity of sample response distribution structure and neural adaptation, which occurs across species and modalities \cite{dragoi2002dynamics, ulanovsky2003processing, sharpee2006adaptive}.
+Characterising noise correlations within stimulus-response distributions is complicated by the heterogeneity of sample response distribution structure and neural adaptation, which occurs across species and modalities \cite{dragoi2002dynamics, ulanovsky2003processing, sharpee2006adaptive}.
 % That is, the response distributions are \textit{non-stationary} and change monotonically as a function of trial index for some neurons. 
-Processes of adaptation, which we define as different to latent states underlying trial-to-trial noise correlations, could also lead to correlative structure in the stimulus-response distribution (Fig.
+Processes of adaptation, which we define as different to continuous latent states underlying trial-to-trial noise correlations, could also lead to correlative structure in the stimulus-response distribution (Fig.
 X).
@@
 
-The nature of first spike patterns has been explored in previous research.
+First spike patterns have been explored in previous research.
 First spike times have been observed with millisecond order precision relative to stimulus onset across species and modalities in sensory organs \cite{johansson2004first, uzzell2004precision, gollisch2008rapid}, thalamic areas \cite{reinagel2002precise, storchi2012comparison} and early sensory cortices \cite{reyes2014laminar}, as well as relative to oscillations \cite{havenith2011synchrony}.
@@
 Another study analysing the data set could decode stimulus properties (stimulation frequency and location) from spike counts and spike patterns but found only weak spike count noise correlations \cite{reyes2015high}.
-The complex nature of spike time correlations uncovered in this study for the same data set may therefore offer an improved opportunity to study population-wide modulation by low-dimensional latent states.
-The new analysis uses time series analysis to control for neural adaptation over all stimulus trials.
-Adaptation occurs in the barrel cortex to repetitive whisker stimulation for both spike count \cite{ahissar2000transformation, ahissar2001temporal, kheradpezhouh2017response, khatri2009stimulus, martin2014tactile, barros2019response} and spike latency \cite{ahissar2000transformation, ahissar2001temporal, kheradpezhouh2017response}, and was confirmed to occur for 7 out of 8 neuron groups in this study.
+The spike time noise correlations uncovered in this study offer a new opportunity to study population-wide modulation by low-dimensional latent states and controls for neural adaptation, which occurs in the barrel cortex to repetitive whisker stimulation for both spike count \cite{ahissar2000transformation, ahissar2001temporal, kheradpezhouh2017response, khatri2009stimulus, martin2014tactile, barros2019response} and spike latency \cite{ahissar2000transformation, ahissar2001temporal, kheradpezhouh2017response}.
+% and was confirmed to occur for 7 out of 8 neuron groups in this study.
 
@@
 
-For repeated trials of a single stimulus, the first spike times of two units were plotted against each other for trials on which they both spiked (see plots in the top left of Supplementary Vid.
-1).
-A novel algorithm extracted isolated clusters from these single stimulus first spike pair response distributions, demonstrating that temporally precise spike patterns fire on subsets of trials, as shown previously for the data set \cite{reyes2014laminar}.
-The novel algorithm, however, also showed that first spike pairs within these isolated clusters were positively correlated above chance-levels.
-
-Moreover, the correlation angle of many of the correlated clusters was different from $45^{\circ}$.
-This cannot be explained by a trial-to-trial shift in the onset of the stimulus or cortical activity, which would produce correlations with a unitary gradient (i.e.
+For repeated trials of a single stimulus, the first spike times of two units were plotted against each other for trials on which they both spiked.
+For example, Fig.
+X shows first spike pairs for a pair of units (L5AI and L4I) in the principle column in response to a 4Hz stimulation of the corresponding principle whisker.
+First spike pairs appear to come from one of two clusters (blue and green dots) for presentations of the same stimulus.
+Moreover, the green cluster appears correlated with non-unitary gradient (different from $45^{\circ}$).
+
+Such correlated clusters were observed in a large number of first spike pair distributions (Supplementary Vid.
+1) and represent reliable regions of the response distribution space that can be analysed to test the nature of spike timing noise correlations.
+The correlative structure could be caused by neural adaptation, a continuous non-adaptative latent state or a combination of the two.
+
+% The paper aims to test that the occurrence of correlated clusters is above chance and to test the degree to which such correlations are caused by adaptative and non-adaptative processes.
+
+% (Fig X; see plots in the top left of Supplementary Vid. 1).
+
+A novel algorithm extracted isolated clusters from single stimulus first spike pair response distributions, demonstrating that temporally precise spike patterns fire on subsets of trials, as shown previously for the data set \cite{reyes2014laminar}.
+The algorithm, however, also showed that first spike pairs within these isolated clusters were positively correlated above chance-levels.
+A large proportion of these noise correlations would have been missed without clustering.
+Moreover, many of the correlated clusters had non-unitary correlation angles.
+
+Importantly, clusters remained correlated with non-unitary gradient after controlling for neural adaptation,
+% using time series analysis, 
+giving evidence of precise non-adaptive noise correlations.
+Such non-unitary spike time correlation angles cannot be explained by a trial-to-trial shift in the onset of the stimulus or cortical activity, which would produce correlations with a unitary gradient (i.e.
 a $45^{\circ}$ angle).
-Instead the `non-unitary' correlation gradients (different from $45^{\circ}$) may be indicative that correlations are due to an internal shared latent state.
-Clusters remained correlated with non-unitary gradient after controlling for neural adaptation using time series analysis, confirming the correlated clusters as evidence of precise non-adaptive noise correlations.
-A large proportion of these noise correlations would have been missed without clustering.
-
-Importantly, the observation of adaptation-independent correlated clusters with non-unitary gradient shows that the distribution of decodable first spike time relative differences for pairs of neurons are `warped' from trial-to-trial by some `non-adaptive' continuous latent state.
-This phenomenon is referred to as `adaptation-independent clustered time-warped spike timing'.
-This differs from previous `time-warp' studies, which aim to align spike trains over trials and do not consider discriminable encoding, clustering, adaptation and different warping functions for different neurons \cite{williams2020discovering}.
-One previous study has modelled time-warp-invariant integration \cite{gutig2009time}, although neurons need not respond identically to warps of the same pattern.
-Rather, neurons should process relative spike patterns in a state-dependent manner to utilise the improved state-dependent coding capacity.
-
-Adaptation-controlled non-unitary correlated clusters appeared for pairs of neurons spatially separated over multiple barrels and cortical layers.
-This offers suggestive evidence of a shared low-dimensional continuous latent state affecting the spike times of spatially distanced neurons, which in turn could also modulate downstream neurons and allow them to respond in a state-dependent manner.
+Instead the `non-unitary' correlation gradients may be indicative of covariation by a non-adaptative continuous latent state that varies from trial-to-trial.
+
+This novel finding of `adaptation-independent clustered time-warped spike-timing noise correlations' shows that the distribution of decodable first spike time relative differences for pairs of neurons are `warped' from trial-to-trial by some `non-adaptive' continuous latent state.
+% This phenomenon is referred to as `adaptation-independent clustered time-warped spike-timing noise correlations'. 
+This differs from previous `time-warp' studies, which aim to align spike trains over trials and do not simultaneously consider stimulus encoding, clustering, adaptation and different warping functions for different neurons \cite{williams2020discovering}.
+
+The characterisation of adaptation-independent clustered spike-timing noise correlations is useful under the fixed-state decoder interpretations where noise correlations can either increase or decrease encoding capacity \cite{averbeck2006neural}.
+More interestingly, these noise correlations open the possibility of state-dependent spike time coding.
 Factor analysis provided an estimate of state-dependent response distributions of first spike pairs (and relative first spike differences) for each adaptation-controlled correlated cluster.
@@
 
-The number of low-dimensional continuous latent states underlying the correlative structure, however, must be confirmed.
-Estimating whether cluster correlations observed for different pairs of neurons is modulated by a single or low-dimensional shared state was made difficult by variability between clusters in the trials from which cluster samples came [See notes below].
-The work of this paper in extracting and characterising adaptation-independent correlated clusters provides a necessary component step towards confirming or rejecting state-dependent spike-time coding, and is also important under the classical fixed-state decoder interpretation.
+Adaptation-independent non-unitary correlated clusters appeared for pairs of neurons spatially separated over multiple barrels and cortical layers.
+This offers suggestive evidence of a shared low-dimensional continuous latent state affecting the spike times of spatially distanced neurons, which in turn could also modulate downstream neurons and allow them to respond in a state-dependent manner.
+% The number of low-dimensional continuous latent states underlying the correlative structure, however, must be confirmed. 
+Estimating whether cluster correlations observed for different pairs of neurons are modulated by a single or low-dimensional shared state was made difficult by variability between clusters in the trials from which cluster samples came [See notes below].
+The work of this paper in extracting and characterising adaptation-independent correlated clusters provides a necessary component step towards confirming or rejecting state-dependent spike-time coding.
 The results show that consideration of clustering, correlative structure and adaptation, together, are necessary to explain a large proportion of neural variability.
+
+% Whether caused by adaptative and/or non-adaptative processes, the presence of correlated clusters with non-unitary correlation angles would be evidence that the distribution of relative spike time differences that representations of a stimulus come from are not fixed.
+If a continuous non-adaptative latent state underlies the correlations, then the state-dependent distribution of relative differences depends on the value of this state on a single trial.
+If neural adaptation underlies such correlations then the, than the the distribution of relative differences on a single trial covers a smaller region of space than the region from which responses come from over all trials.
+[HERE] If a combination of an adaptative and non-adaptative continuous latent state underlies the correlations then..
+
+% The orange Gaussian in Fig.
+3a is centered at the mean of the first spike pairs and has widths equal to 4 times the first spike time standard deviations of each unit (on conjunctive trial), illustrating the region from which 99.9\% of first spike pairs would be predicted to fall under the assumption of independence and normality (i.e.
+spike time jitter).
+This assumption fails to capture the distribution structure of this and many other spike pair distributions (Supplementary Vid.
+1).
+
+% Moreover, the apparent non-unitary gradient of correlation suggests that the distribution of differences varies with respect to the value of this underlying state, such that the state-conditioned distribution of relative first spike differences would be more compact than the general case, thus offering a higher potential capacity for stimulus encoding if downstream neurons could respond in a state-dependent manner.
 
@@
 
-\subsection*{First spike pair response distributions}
+% \subsection*{First spike pair response distributions}
 
@@
 
+\subsection*{Stage 1 Clusters - Correlation analysis}
+
 Analysing the joint responses of two or more units to repeated presentations of a single stimulus offers the opportunity to discern noise correlations caused by adaptative or non-adaptative processes.
-For the purpose of understanding, this section presents an example distribution of first spike pairs for a pair of units (L5AI and L4I) in the principle column in response to a 4Hz stimulation of the corresponding principle whisker (over a 50ms window following cortical activation onset).
-Only responses on `conjunctive trials' (trials in which both units spiked for the single stimulus) are shown.
-
-First spike pairs appear to come from one of two clusters (blue and green dots) for presentations of the same stimulus.
-Moreover, the green cluster appears correlated with non-unitary gradient, suggestive that some underlying processes (adaptative or non-adaptative) may covary the first spike times of the two units.
-Such correlated clusters were observed in a large number of first spike pair distributions (Supplementary Vid.
+A two-stage algorithm was developed to quantify the occurrence and angle of correlated clusters observed in first spike response distributions for pairs of units (Fig.
+X; top left of Supplementary Vid.
 1).
-
-This paper aims to quantify the occurrence of such correlated clusters and determine when adaptative and/or non-adaptative processes underlie such noise correlations.
-Importantly, the apparent non-unitary gradient of correlation of the green cluster in Fig.
-3a suggests that the distribution of differences varies with respect to the possible adaptative and/or non-adaptative processes underlying the correlation.
-
-The orange Gaussian in Fig.
-3a is centered at the mean of the first spike pairs and has widths equal to 4 times the first spike time standard deviations of each unit (on conjunctive trial), illustrating the region from which 99.9\% of first spike pairs would be predicted to fall under the assumption of independence and normality (i.e.
-spike time jitter).
-This assumption fails to capture the distribution structure of this and many other spike pair distributions (Supplementary Vid.
-1).
-% Moreover, very few different stimuli could be encoded with high-discriminability.
-
-% The spike time differences have a standard deviation of X (Fig. 3b), which is also poor for discriminable encoding.
-
-Moreover, the apparent non-unitary gradient of correlation suggests that the distribution of differences varies with respect to the value of this underlying state, such that the state-conditioned distribution of relative first spike differences would be more compact than the general case, thus offering a higher potential capacity for stimulus encoding if downstream neurons could respond in a state-dependent manner.
-
-% \subsection*{Clustering algorithm summary}
-
-\subsection*{Stage 1 Clusters - Correlation analysis}
-
-A two-stage algorithm was developed to quantify the occurrence and angle of correlated clusters.
 Stage 1 of the algorithm extracted well isolated clusters defined by the $4\sigma$ boundary of flat 2D Gaussian ellipses (see Methods \ref{methods:flat_clusters}; e.g.
@@
 
+\subsection*{Time warp integration}
+
+One previous study has modelled time-warp-invariant integration \cite{gutig2009time}, although neurons need not respond identically to warps of the same pattern.
+Rather, neurons should process relative spike patterns in a state-dependent manner to utilise the improved state-dependent coding capacity.
+
 \subsection*{Underlying state}

### 2020-08-11 manuscript.tex
gut

@@
 
-Such correlated clusters were observed in a large number of first spike pair distributions (Supplementary Vid.
-1) and represent reliable regions of the response distribution space that can be analysed to test the nature of spike timing noise correlations.
-The correlative structure could be caused by neural adaptation, a continuous non-adaptative latent state or a combination of the two.
-
-% The paper aims to test that the occurrence of correlated clusters is above chance and to test the degree to which such correlations are caused by adaptative and non-adaptative processes.
-
-% (Fig X; see plots in the top left of Supplementary Vid. 1).
-
+Correlated clusters were observed in a large number of first spike pair distributions (Supplementary Vid.
+1).
 A novel algorithm extracted isolated clusters from single stimulus first spike pair response distributions, demonstrating that temporally precise spike patterns fire on subsets of trials, as shown previously for the data set \cite{reyes2014laminar}.
 The algorithm, however, also showed that first spike pairs within these isolated clusters were positively correlated above chance-levels.
+These correlated clusters represent reliable regions of response distribution space that allow analysis of spike timing noise correlations possibly caused by neural adaptation, a continuous non-adaptative latent state or a combination of both.
 A large proportion of these noise correlations would have been missed without clustering.
-Moreover, many of the correlated clusters had non-unitary correlation angles.
-
+
+% The paper aims to test that the occurrence of correlated clusters is above chance and to test the degree to which such correlations are caused by adaptative and non-adaptative processes.
+
+% (Fig X; see plots in the top left of Supplementary Vid. 1).
+
+Interestingly, many of the correlated clusters had non-unitary correlation angles. 
 Importantly, clusters remained correlated with non-unitary gradient after controlling for neural adaptation,
@@
 a $45^{\circ}$ angle).
-Instead the `non-unitary' correlation gradients may be indicative of covariation by a non-adaptative continuous latent state that varies from trial-to-trial.
-
-This novel finding of `adaptation-independent clustered time-warped spike-timing noise correlations' shows that the distribution of decodable first spike time relative differences for pairs of neurons are `warped' from trial-to-trial by some `non-adaptive' continuous latent state.
+Instead the `non-unitary' correlation gradients are indicative of covariation by a non-adaptative continuous latent state that varies from trial-to-trial.
+
+This novel finding of `adaptation-independent clustered time-warp spike-timing noise correlations' shows that the distribution of decodable first spike time relative differences for pairs of neurons are `warped' from trial-to-trial by some `non-adaptive' continuous latent state.
 % This phenomenon is referred to as `adaptation-independent clustered time-warped spike-timing noise correlations'. 

### 2020-08-11 AnalysisOutputLatex.tex
End of day!

@@
 %<*ClusterCorrelationSummary0p1>
-Moreover, for Stage 2 clusters, spike pair differences were correlated with the spike time of the first unit in the pair (Fisher's method: $\chi^{2}(324, N = 162) = 7832.7, $ $p < .001$; Fig.
-3g), showing that correlations were not explained by a model of the form $s_1 = s_0 + d + independent\_noise$ where $d$ is a fixed difference.
+spike pair differences were correlated with the spike time of the first unit in the pair for Stage 2 clusters (Fisher's method: $\chi^{2}(324, N = 162) = 7832.7, $ $p < .001$; Fig.
+3g), shows that correlations are not explained by a model of the form $s_1 = s_0 + d + independent\_noise$ where $d$ is a fixed difference.
 %</ClusterCorrelationSummary0p1>
@@
 %<*ClusterCorrelationSummary2>
-80.9\% (131/162) of the Stage 2 clusters had correlations of higher significance than correlations calculated for all unclustered first spike pairs in the original response distribution (Fig.
+80.9\% (131/162) of the Stage 2 clusters had correlations of higher significance than correlations calculated for all unclustered first spike pairs in the originating response distribution (Fig.
 3h).
@@
 %<*SelectivelyDifferencedTestsPassedNewSummary>
-In total, 83 clusters had differencing applied to atleast one unit.
-35 of these were correlated and fulfilled the criteria.
-23 of these clusters had estimated angles significantly greater than $0^{\circ}$ (p-value < 0.025; Fig 6b) and less than $45^{\circ}$ (p-value < 0.025; Fig 6b), of which 12 normality was not rejected for (Henze-Zirkler p < 0.05).
-These angles are similar to the original angles before differencing (Fig.
-6c), showing that the differencing operation does not largely alter cluster angle.
+In total, 83 clusters had differencing applied to the cluster spike times of at least one unit.
+35 of these fulfilled the stationarity and correlation criteria (Methods \ref{methods:angle_and_factor_analysis_criteria}).
+23 of these clusters had estimated angles significantly less than $45^{\circ}$ (p-value < 0.025; Fig 6b) and greater than $0^{\circ}$ (p-value < 0.025; Fig 6b), of which for 12 normality was not rejected for (Henze-Zirkler p > 0.05).
+The differencing operation did not largely alter the cluster angle of the criteria fulfilling differenced clusters (Fig.
+6c).
 Fig.
@@
 %<*SelectivelyDifferencedTestsPassedEigenvalueSummary>
-The factor analysis model from above can be applied to the 35 correlated criteria fulfilling differenced clusters, as the factor correlation matrices of each cluster only had one eigenvalue greater than 1.
+The factor analysis model from above can be applied to the 35 correlated criteria fulfilling differenced clusters. 
 %</SelectivelyDifferencedTestsPassedEigenvalueSummary>

### 2020-08-11 manuscript.tex
End of day!

@@
 3).
+The correlated clusters can be analysed to test whether the correlations are explained by adaptative or non-adaptative processes.
 
@@
 \ExecuteMetaData[AnalysisOutputLatex.tex]{ClusterCorrelationSummary2}
+These results suggest show that clustering is necessary to uncover spike time noise correlations between a large proportion of unit pairs.
 
@@
 
-The non-unitary correlations cannot be explained by a model in which the second spiking neuron spikes a fixed time difference after the first. 
+Finally, the non-unitary correlation angles and the fact that
+% cannot be explained by a model in which the second spiking neuron spikes a fixed time difference after the first. 
 \ExecuteMetaData[AnalysisOutputLatex.tex]{ClusterCorrelationSummary0p1}
-This strengthens the case that the spike timing correlations between pairs of neurons are co-modulated by some underlying process.
+This strengthens the case that the spike timing correlations between pairs of neurons are co-modulated by some underlying process or processes.
 
@@
 
-This section tests both whether cluster spike times for individual units are affected by neural adaptation and also the degree to which neural adaptation explains cluster correlations.
+This section tests both whether cluster spike times for individual units are affected by neural adaptation. 
+% and also the degree to which neural adaptation explains cluster correlations.
 \ExecuteMetaData[AnalysisOutputLatex.tex]{ClusterSingleUnitTrialIndexLRSummary}
@@
 % More complex first spike time relationships with trial index were also observed. 
-Lag-1 autocorrelations of single unit cluster spike times were above chance [STAT], as expected by the spike times of some units being affected by adaptation.
-Beside general adaptive trends, some single units exhibited short trial-by-trial sequences (3-4 trials) of autocorrelated first spike times, before losing such autocorrelations. 
+% Lag-1 autocorrelations of single unit cluster spike times were above chance [STAT], as expected by neural adaptation affecting the spike times of some units.
+Beside general adaptive trends, some single units exhibited short and occasional trial-by-trial sequences (3-4 trials) of autocorrelated first spike times.
 % Lagged autocorrelations were significantly above chance for a high number of lags (Supplementary Fig. 4) supportive of general adaptive trends. 
 Only the lag-2 partial autocorrelation (lags 2-5 tested) was significantly above chance (p-value < 0.005; Supplementary Fig.
-4), representative of the minimal and temporary autocorrelations observed for several single units.
-
-Cluster spike times for cluster pairs showed above chance lag-1 cross-correlations, which could be explained by clusters for which the spike times of both units were modulated by adaptation.
+4), representative of these short and occasional lagged autocorrelations.
 Only the lag-2 partial cross-correlation was slightly above chance (lags 2-5 tested) (Supplementary Fig.
 5), suggestive of minimal lagged cross-correlative structure between cluster spike times for some unit pairs after neural adaptation.
+Such highly infrequent and short lag-2 auto and cross-correlations could be ignored when determining cluster stationarity.
+
+% Cluster spike times for cluster pairs showed above chance lag-1 cross-correlations, which could be explained by clusters for which the spike times of both units were modulated by adaptation.
 
@@
 This section demonstrated heterogeneity in the non-stationarity of Stage 2 clusters.
-Particularly, either the spike times of both units were not affected by adaptation (i.e.
-stationary clusters) or one or both of the units have cluster spike times modulated by adaptation (i.e.
+Particularly, either the spike times of both units were not modulated by adaptation (i.e.
+stationary clusters) or the cluster spike times of one or both of the units are adaptation-modulated (i.e.
 non-stationary clusters).
-These two cases must be treated differently when testing for adaptation-independent noise correlations.
+In the latter case, neural adaptation must be accounted for when testing for adaptation-independent noise correlations.
 
@@
 7), confirming that the stationary Stage 2 cluster correlations are not caused by neural adaptative trends.
+\ExecuteMetaData[AnalysisOutputLatex.tex]{OriginalTestsPassedFactorAnalysisSummary} 
+% The following sections apply time series analysis techniques to remove the neural adaptive trends of the non-stationary units of the remaining clusters.
 
@@
 
-\ExecuteMetaData[AnalysisOutputLatex.tex]{OriginalTestsPassedFactorAnalysisSummary} 
-% The following sections apply time series analysis techniques to remove the neural adaptive trends of the non-stationary units of the remaining clusters.
-
 The following single factor model was fit to the elements of each (zero-centered) cluster: $s_i = \lambda_i \eta + \sigma_i$ for $i \in [0, 1]$ 
@@
 8) comparison of the estimated state-conditioned response distribution to the larger $4\sigma$ areas under fixed-state decoder assumptions of normality and independence (orange), normal and independent clusters (green flat cluster) and normal and correlated clusters (green angled cluster).
-The predicted state-conditioned response distributions are strikingly smaller than those predicted under the fixed-state assumptions, thus potentially offering an improved encoding capacity.
+The predicted state-conditioned response distributions are strikingly smaller than those predicted under the fixed-state assumptions, thus offering an improved encoding capacity if downstream neurons can respond in a state-dependent manner.
 
@@
 
-[UP TO HERE]
-
-To test whether a shared state could underlie covariation in the remaining non-stationary angled clusters, the non-stationary single unit cluster spike times were differenced and rescaled (a standard time-series analysis technique) to remove neural adaptive trends by subtracting the spike time on trial t by the spike time on trial t-1.
+To test whether a non-adaptative latent state could also modulate the non-stationary Stage 2 clusters, non-stationary single unit cluster spike times were differenced and rescaled (a standard time-series analysis technique) to remove neural adaptive trends (see Methods \ref{methods:differencing}).
+% by subtracting the spike time on trial t by the spike time on trial t-1.
 % Clusters which did not fulfil the criteria after differencing were removed from further analysis. 
 % Differencing a time series reduces the number of samples by 1 (Methods \ref{methods:angle_and_factor_analysis_criteria}). 
-The green angled cluster of Fig.
-3a fulfils the criteria after the spike times of its non-stationary unit are differenced (Fig.
+As expected by differencing, the spike times of the differenced unit show negative lag-1 and 2 partial autocorrelations (Fig. 6a; Supplementary Fig. 9?). 
+% Due to the expected negative autocorrelations introduced as expected by differencing (Supplementary Fig. 9), 
+The autocorrelation criterion (criterion 6, Methods \ref{methods:angle_and_factor_analysis_criteria}) for cluster stationarity was therefore relaxed.
+The following section applies a more complex time series model to remove the negative autocorrelations.
+The assessment of `raw' differences in this section, however, allows a direct assessment of cluster angles after trend removal.
+
+The green Stage 2 cluster of Fig.
+3a fulfils the stationarity and correlation criteria (Methods \ref{methods:angle_and_factor_analysis_criteria}) after the spike times of its non-stationary unit are differenced (Fig.
 6a).
-Application of the angle estimation technique estimated that the angle was X\todo[color=red!40]{Fill} plus minus Y (Fig.
-6a), similar to the original angle and suggestive of an underlying factor after the removal of trial dependence.
-As expected by differencing, the spike times of the differenced unit show negative lag-1 autocorrelations and negative lag-2 partial autocorrelations (Fig.
-6a).
+The $\theta_{45}$ angle was calculated as X\todo[color=red!40]{Fill} plus minus Y (Fig.
+6a), similar to the original angle and suggestive of covariation by an underlying state after the removal of the non-stationary unit's adaptative trend.
+
 % Such negative autocorrelations are expected by differencing, and are removed in the next section by fitting ARIMA models.
 
-Due to the expected negative autocorrelations introduced by differencing (Supplementary Fig.
-9), the criteria for cluster unit autocorrelation p-values < 0.05, was relaxed for the differenced clusters (criterion 6, Methods \ref{methods:angle_and_factor_analysis_criteria}).
 \ExecuteMetaData[AnalysisOutputLatex.tex]{SelectivelyDifferencedTestsPassedNewSummary}
@@
 Again, the factor analysis angles show close similarity to the PCA angles (Fig. 6e).
-The standard deviations of the factor-dependent response distributions (estimated by factor analysis) again offer a strikingly higher potential for encoding than simpler neural coding assumptions (Fig.
+The standard deviations of the state-dependent response distributions estimated by factor analysis again offer a strikingly higher potential for encoding than simpler neural coding assumptions (Fig.
 6a; Supplementary Vid 1; Fig.

### 2020-08-12 AnalysisOutputLatex.tex
Bravo!

@@
 Normality was not rejected for 18 of these clusters (Henze-Zirkler p > 0.05).
-27 of the correlated criteria fulfilling clusters had estimated angles significantly greater than $0^{\circ}$ (p-value < 0.025; Fig.
-7b) and less than $45^{\circ}$ (p-value < 0.025; Fig.
+27 of the correlated criteria fulfilling clusters had estimated angles significantly less than $45^{\circ}$ (p-value < 0.025; Fig.
+7b) and greater than $0^{\circ}$ (p-value < 0.025; Fig.
 7b), for which 10 normality was not rejected for (Henze-Zeckler p-value > 0.05).
@@
 %<*SelectivelyDifferencedBoxJenkinsTestsPassedEigenvalueSummary>
-The factor analysis model from above can be applied to the 42 correlated criteria fulfilling clusters, as the factor correlation matrices of each cluster only had one eigenvalue greater than 1. 
+The factor analysis model from above can be applied to the 42 correlated criteria fulfilling clusters. 
 %</SelectivelyDifferencedBoxJenkinsTestsPassedEigenvalueSummary>

### 2020-08-12 manuscript.tex
Bravo!

@@
 
-\usepackage[nomarkers,notablist,nofiglist]{endfloat}
+% \usepackage[nomarkers,notablist,nofiglist]{endfloat}
 \usepackage[colorinlistoftodos,disable]{todonotes}
@@
 
-\title{evidence of adaptation-independent time-warped spike timing noise correlations in cortical networks}
+\title{evidence of adaptation-independent time-warp spike timing noise correlations in cortical networks}
 
@@
 
-%   \And
-%   Irina Higgins\\
-%   DeepMind
-
   \And
@@
 
-% \begin{abstract}
+\begin{abstract}
 % % http://blogs.nature.com/naturejobs/2014/11/03/how-to-get-published-in-high-impact-journals-big-research-and-better-writing/
@@
 
-% Patterns of action potentials induced by sensory stimulation in vivo allow investigation of neural coding principles.
+Patterns of action potentials induced by sensory stimulation in vivo allow investigation of neural coding principles.
 Trial-to-trial variability engenders debate over temporal and rate codes.
@@
 
-% \end{abstract}
+\end{abstract}
 
@@
 Whilst precise temporal spike patterns have been observed \cite{uzzell2004precision, johansson2004first, gollisch2008rapid, storchi2012comparison, reinagel2002precise, reyes2014laminar}, a large proportion of spike time variability remains unexplained.
-Such variability leads to the common assumption of population codes based on spike counts \cite{reyes2015high, lin2015nature} or trial-averaged firing rates \cite{hong2016explicit, chang2017code, kar2019evidence, stringer2019high}.
+Such variability leads to the common assumption of population spike count \cite{reyes2015high, lin2015nature} or trial-averaged firing rate \cite{hong2016explicit, chang2017code, kar2019evidence, stringer2019high} codes.
+Previous studies have shown that spike count variability is shared across the population \cite{scholvinck2015cortical, lin2015nature}.
+As single cortical neurons can generate highly reliable spike trains to direct stimulation in vitro \cite{mainen1995reliability}, in vivo shared variability is likely to arise from variability in synaptic inputs, reflecting cortical network dynamics \cite{carandini2004amplification}.
+
+In early visual cortex, trial-to-trial spike count variability is correlated across the population and shown to be modulated by two global factors, which fluctuate the level of excitability from trial-to-trial \cite{lin2015nature}.
+Firstly, an additive factor is explained by the interaction of evoked activity with population-wide spontaneous activity \cite{arieli1996dynamics, scholvinck2015cortical}.
+A second multiplicative factor also modulates the trial-to-trial fluctuation in excitability across the population, and has been proposed to be caused by top-down modulation...
+Fluctuations in spontaneous activity have also been shown to gate sensory-evoked responses \cite{luczak2013gating}.
+
+Fluctuations in trial-to-trial excitability are referred to herein as trial-to-trial fluctuations in continuous cortical state \cite{scholvinck2015cortical}, which may be multi-dimensional and can be separated into short and long-timescale fluctuations.
+In this paper, long-timescale refer to large changes taking place over multiple trials that are caused by neural adaptation in response to repetitive stimulation or possibly due to changes in the level of anaesthesia.
+Short-timescale fluctuations refer to random trial-to-trial fluctuations in excitability around the current general level of excitability.
+
+% Other studies have reported only weak spike count correlations in sensory cortices \cite{reyes2015high}.
+
+Rather than spike count noise correlations, this study aims to characterise the nature of trial-to-trial spike time noise correlations as a step towards confirming or rejecting the modulation of spike times by short-timescale fluctuations of a low-dimensional continuous cortical state.
+This offers a novel prediction of the encoding capacity of state-dependent spike time encoding, which controls for longer timescale changes in excitability such as neural adaptation.
+Characterising spike timing codes is important to understanding a variety of dynamic and plastic processes, including properties of action potential patterns directly involved in cognition.
+
+%  low 2-dimensional continuous latent state .
+
+% It is possible that spike time variability is caused and modulated across a population by such a low-dimensional continuous latent state.
+
 % \cite{gerstner1997neural}, which have capacity beyond the original spike count \cite{}. 
@@
 % These population codes assume that trial-to-trial variability is either independent or correlated across the population \cite{averbeck2006neural, lin2015nature}.
-In a recent study in early visual cortex, trial-to-trial spike count variability was shown to be correlated across the population and modulated by an underlying low 2-dimensional continuous latent state \cite{lin2015nature}.
-Other studies have reported only weak spike count correlations in sensory cortices \cite{reyes2015high}.
-
-It is possible that spike time variability is caused and modulated across a population by such a low-dimensional continuous latent state.
-This study aims to characterise the nature of trial-to-trial spike time noise correlations as a step towards confirming or rejecting the modulation of spike times by a continuous low-dimensional latent state.
-This offers a novel prediction of the encoding capacity of state-dependent spike time encoding.
-These results control for neural adaptation.
-Characterising spike timing codes is important to understanding a variety of dynamic and plastic processes, including properties of action potential patterns directly involved in cognition.
 
@@
 
-To remove autocorrelations in cluster unit time series (particularly negative autocorrelations introduced by differencing) and to return the differenced time series to the original scale without division by 2, autoregressive (AR) and/or moving average (MA) models were fit to differenced correlated criteria fulfilling angled clusters, using an automated version of the Box-Jenkins method (Methods \ref{methods:automated_box_jenkins}).
-
+To remove autocorrelations in cluster unit time series (particularly negative autocorrelations introduced by differencing) and to return the differenced time series to the original scale without rescaling, autoregressive (AR) and/or moving average (MA) models were fit to differenced clusters, using an automated version of the Box-Jenkins method (Methods \ref{methods:automated_box_jenkins}).
  \ExecuteMetaData[AnalysisOutputLatex.tex]{SelectivelyDifferencedBoxJenkinsDifferencedSummary}

### 2020-08-12 manuscript.tex
Lunch!

@@
 
-% \usepackage[nomarkers,notablist,nofiglist]{endfloat}
+\usepackage[nomarkers,notablist,nofiglist]{endfloat}
 \usepackage[colorinlistoftodos,disable]{todonotes}
@@
 \section{Introduction}
+
 The neural coding problem addresses the question of how information is encoded by patterns of action potentials in the nervous system \cite{gerstner1997neural, johnson2000neural, stein2005neuronal, gerstner2014neuronal}.
 Whilst precise temporal spike patterns have been observed \cite{uzzell2004precision, johansson2004first, gollisch2008rapid, storchi2012comparison, reinagel2002precise, reyes2014laminar}, a large proportion of spike time variability remains unexplained.
-Such variability leads to the common assumption of population spike count \cite{reyes2015high, lin2015nature} or trial-averaged firing rate \cite{hong2016explicit, chang2017code, kar2019evidence, stringer2019high} codes.
-Previous studies have shown that spike count variability is shared across the population \cite{scholvinck2015cortical, lin2015nature}.
-As single cortical neurons can generate highly reliable spike trains to direct stimulation in vitro \cite{mainen1995reliability}, in vivo shared variability is likely to arise from variability in synaptic inputs, reflecting cortical network dynamics \cite{carandini2004amplification}.
+Such variability leads to the common assumption that the brain uses a population code based on spike counts \cite{reyes2015high, lin2015nature} or trial-averaged firing rates \cite{hong2016explicit, chang2017code, kar2019evidence, stringer2019high}.
+Previous studies have shown that spike count variability is shared across the population in early sensory cortices \cite{scholvinck2015cortical, lin2015nature}.
+Single cortical neurons can generate highly reliable spike trains to direct stimulation in vitro \cite{mainen1995reliability}.
+In vivo shared variability is therefore likely to arise from variability in synaptic inputs, reflecting cortical network dynamics \cite{carandini2004amplification}.
 
 In early visual cortex, trial-to-trial spike count variability is correlated across the population and shown to be modulated by two global factors, which fluctuate the level of excitability from trial-to-trial \cite{lin2015nature}.
-Firstly, an additive factor is explained by the interaction of evoked activity with population-wide spontaneous activity \cite{arieli1996dynamics, scholvinck2015cortical}.
-A second multiplicative factor also modulates the trial-to-trial fluctuation in excitability across the population, and has been proposed to be caused by top-down modulation...
+The first of these is an additive factor, which is explained by the interaction of evoked activity with population-wide spontaneous activity \cite{arieli1996dynamics, scholvinck2015cortical}.
+The second is a multiplicative factor, which also modulates the trial-to-trial fluctuation in excitability across the population, and has been proposed to be caused by top-down feedback from higher cortices \cite{lin2015nature}.
 Fluctuations in spontaneous activity have also been shown to gate sensory-evoked responses \cite{luczak2013gating}.
 
-Fluctuations in trial-to-trial excitability are referred to herein as trial-to-trial fluctuations in continuous cortical state \cite{scholvinck2015cortical}, which may be multi-dimensional and can be separated into short and long-timescale fluctuations.
-In this paper, long-timescale refer to large changes taking place over multiple trials that are caused by neural adaptation in response to repetitive stimulation or possibly due to changes in the level of anaesthesia.
+Fluctuations in trial-to-trial excitability are referred to herein as trial-to-trial fluctuations in continuous cortical state \cite{scholvinck2015cortical}, which may be multi-dimensional and can be separated into short and long-timescale changes.
+In this paper, long-timescale fluctuations refer to large changes taking place over multiple trials caused by processes such as neural adaptation in response to repetitive stimulation or changes in the level of anaesthesia.
 Short-timescale fluctuations refer to random trial-to-trial fluctuations in excitability around the current general level of excitability.
@@
 
-Rather than spike count noise correlations, this study aims to characterise the nature of trial-to-trial spike time noise correlations as a step towards confirming or rejecting the modulation of spike times by short-timescale fluctuations of a low-dimensional continuous cortical state.
-This offers a novel prediction of the encoding capacity of state-dependent spike time encoding, which controls for longer timescale changes in excitability such as neural adaptation.
+This study aims to characterise the non-trivial nature of trial-to-trial spike time (rather than spike count) noise correlations, as a component step towards investigating the modulation of spike times by short-timescale fluctuations of a low-dimensional continuous cortical state.
+% which would introduce the possibility of state-dependent encoding and decoding (see below). 
+The characterisation controls for longer timescale changes in excitability to estimate the short-timescale state-modulated representation of a stimulus at a given point in time. 
+% and a novel prediction of the encoding capacity under a state-dependent spike time encoding regime. 
 Characterising spike timing codes is important to understanding a variety of dynamic and plastic processes, including properties of action potential patterns directly involved in cognition.
@@
 % Due to response variability, characterising the neural representation of a single stimulus is a matter of characterising the \textit{distribution} of possible neural responses for the stimulus. 
-Particularly, a multi-neuron response can be expressed as a point in multi-dimensional space, where each dimension represents the value of a spike train metric.
+A multi-neuron response can be expressed as a point in space, where each dimension represents the value of a spike train metric.
 Responses from multiple presentations of a stimulus form a sample stimulus-response distribution in this space, which can be used to estimate an underlying response distribution from which single-trial responses are `drawn' (Fig.
 1a).
+
 % Estimating the underlying response distribution (including correlative structure) for single stimuli is complicated by neural adaptation and heterogeneity of sample response distribution structure.
@@
 1a).
-Neural coding is classically interested in determining the dimensions in which response distributions do not overlap \cite{averbeck2006neural}.
-
-Under such approaches, `noise correlations' refer to shared trial-to-trial variability between the responses of multiple neurons over repeated presentations of single stimuli \cite{averbeck2006neural}.
+Neural coding and noise correlations are classically considered with respect to fixed-state decoders \cite{averbeck2006neural, moreno2014information, stringer2019high}.
+`Noise correlations' refer to shared trial-to-trial variability between the responses of multiple neurons over repeated presentations of single stimuli \cite{averbeck2006neural}.
 `Signal correlations' concern tuning similarity between neurons over multiple stimuli (see \cite{averbeck2006neural}).
-Neural coding and noise correlations are classically considered with respect to fixed-state decoders \cite{averbeck2006neural, moreno2014information, stringer2019high}.
+
+% Neural coding is classically interested in determining the dimensions in which response distributions do not overlap \cite{averbeck2006neural}.
+
+% Under such approaches,
 
@@
 
-It is important to consider, that if downstream neurons are also modulated by the same population wide continuous latent state, then downstream `decoder' neurons could respond in a state-dependent manner.
-When the continuous latent state has a certain value on a single trial, a response comes from a \textit{state-conditioned} stimulus-response distribution, which would occupy a smaller region of response distribution space than the region sampled over multiple trials (Fig.
+It is important to consider that if downstream neurons are also modulated by the same population-wide short-timescale fluctuating state, then downstream `decoder' neurons could respond in a state-dependent manner.
+When the short-timescale fluctuating state has a certain value on a single trial, a response comes from a \textit{state-conditioned} stimulus-response distribution, which would occupy a smaller region of response distribution space than the region sampled over multiple trials (Fig.
 1c).
@@
 % That is, the response distributions are \textit{non-stationary} and change monotonically as a function of trial index for some neurons. 
-Processes of adaptation, which we define as different to continuous latent states underlying trial-to-trial noise correlations, could also lead to correlative structure in the stimulus-response distribution (Fig.
-X).
-Adaptation must therefore be controlled for when characterising adaptation-independent noise correlations.
+% Processes of adaptation, which we define as different to continuous latent states underlying trial-to-trial noise correlations, could also 
+Long-term changes in cortical state, such as adpatation, could lead to correlative structure in the stimulus-response distribution (Fig.
+X), and must be controlled for when characterising short-timescale or `adaptation-independent' noise correlations.
 
@@
 
-Using a large data set of spike sorted recordings in anesthetized rat barrel cortex, this paper aimed to precisely characterise the nature of spike time noise correlations for first-spike patterns, towards estimating the potential encoding capacity of a state-dependent spike time encoding and determining the dimensionality of such an underlying latent state.
+Using a large data set of spike sorted recordings in anesthetized rat barrel cortex, this paper aimed to precisely characterise the nature of short-timescale spike time noise correlations for first-spike patterns, whilst controlling for neural adaptation, which occurs in the barrel cortex to repetitive whisker stimulation for both spike count \cite{ahissar2000transformation, ahissar2001temporal, kheradpezhouh2017response, khatri2009stimulus, martin2014tactile, barros2019response} and spike latency \cite{ahissar2000transformation, ahissar2001temporal, kheradpezhouh2017response}.
+% towards estimating the potential encoding capacity of a state-dependent spike time encoding and determining the dimensionality of such an underlying latent state.
 The classical `stimulus-dependency' of noise correlations is left to future studies.
-
 A previous study analysing the same data set found precise multi-neuron spike patterns that were intra- and inter-laminar, as well intra- and intra-columnar \cite{reyes2014laminar}.
 Another study analysing the data set could decode stimulus properties (stimulation frequency and location) from spike counts and spike patterns but found only weak spike count noise correlations \cite{reyes2015high}.
-The spike time noise correlations uncovered in this study offer a new opportunity to study population-wide modulation by low-dimensional latent states and controls for neural adaptation, which occurs in the barrel cortex to repetitive whisker stimulation for both spike count \cite{ahissar2000transformation, ahissar2001temporal, kheradpezhouh2017response, khatri2009stimulus, martin2014tactile, barros2019response} and spike latency \cite{ahissar2000transformation, ahissar2001temporal, kheradpezhouh2017response}.
+% The spike time noise correlations uncovered in this study 
+% offer a new opportunity to study population-wide modulation by low-dimensional latent states and 
+% controls for neural adaptation, which occurs in the barrel cortex to repetitive whisker stimulation for both spike count \cite{ahissar2000transformation, ahissar2001temporal, kheradpezhouh2017response, khatri2009stimulus, martin2014tactile, barros2019response} and spike latency \cite{ahissar2000transformation, ahissar2001temporal, kheradpezhouh2017response}.
 % and was confirmed to occur for 7 out of 8 neuron groups in this study.
@@
 A novel algorithm extracted isolated clusters from single stimulus first spike pair response distributions, demonstrating that temporally precise spike patterns fire on subsets of trials, as shown previously for the data set \cite{reyes2014laminar}.
-The algorithm, however, also showed that first spike pairs within these isolated clusters were positively correlated above chance-levels.
-These correlated clusters represent reliable regions of response distribution space that allow analysis of spike timing noise correlations possibly caused by neural adaptation, a continuous non-adaptative latent state or a combination of both.
+It was also shown, however, that first spike pairs within these isolated clusters were positively correlated above chance-levels.
+These correlated clusters represent reliable regions of response distribution space that allow analysis of short- and long-timescale spike timing noise correlations.
 A large proportion of these noise correlations would have been missed without clustering.
@@
 Interestingly, many of the correlated clusters had non-unitary correlation angles. 
-Importantly, clusters remained correlated with non-unitary gradient after controlling for neural adaptation,
+Importantly, many clusters remained correlated with non-unitary gradient after controlling for longer-timescale neural adaptation,
 % using time series analysis, 
-giving evidence of precise non-adaptive noise correlations.
-Such non-unitary spike time correlation angles cannot be explained by a trial-to-trial shift in the onset of the stimulus or cortical activity, which would produce correlations with a unitary gradient (i.e.
-a $45^{\circ}$ angle).
-Instead the `non-unitary' correlation gradients are indicative of covariation by a non-adaptative continuous latent state that varies from trial-to-trial.
-
-This novel finding of `adaptation-independent clustered time-warp spike-timing noise correlations' shows that the distribution of decodable first spike time relative differences for pairs of neurons are `warped' from trial-to-trial by some `non-adaptive' continuous latent state.
+indicative of short-timescale fluctuating state that affects different neurons differently.
+% giving evidence of precise non-adaptive noise correlations
+% Such non-unitary spike time correlation angles cannot be 
+This cannot explained by a trial-to-trial shift in the onset of the stimulus or cortical activity, which would produce correlations with a unitary gradient (i.e. a $45^{\circ}$ angle).
+
+This novel finding of `adaptation-independent clustered time-warp spike-timing noise correlations' shows that the distribution of decodable first spike time relative differences for pairs of neurons at a point in time are `warped' from trial-to-trial by some short-timescale fluctuating continuous cortical state.
 % This phenomenon is referred to as `adaptation-independent clustered time-warped spike-timing noise correlations'. 
@@
 % The number of low-dimensional continuous latent states underlying the correlative structure, however, must be confirmed. 
-Estimating whether cluster correlations observed for different pairs of neurons are modulated by a single or low-dimensional shared state was made difficult by variability between clusters in the trials from which cluster samples came [See notes below].
-The work of this paper in extracting and characterising adaptation-independent correlated clusters provides a necessary component step towards confirming or rejecting state-dependent spike-time coding.
+Estimating whether cluster correlations observed for different pairs of neurons are modulated by a single or low-dimensional shared state was made difficult by low trial overlap between cluster.
+The work of this paper
+% in extracting and characterising adaptation-independent correlated clusters provides 
+a useful component step towards investigating this. 
+% confirming or rejecting state-dependent spike-time coding. 
 The results show that consideration of clustering, correlative structure and adaptation, together, are necessary to explain a large proportion of neural variability.

### 2020-08-12 manuscript.tex
Great!

@@
 
-In early visual cortex, trial-to-trial spike count variability is correlated across the population and shown to be modulated by two global factors, which fluctuate the level of excitability from trial-to-trial \cite{lin2015nature}.
+In early visual cortex, trial-to-trial spike count variability is correlated across the population and is modulated by two global factors, which fluctuate the level of excitability from trial-to-trial \cite{lin2015nature}.
 The first of these is an additive factor, which is explained by the interaction of evoked activity with population-wide spontaneous activity \cite{arieli1996dynamics, scholvinck2015cortical}.
@@
 Fluctuations in trial-to-trial excitability are referred to herein as trial-to-trial fluctuations in continuous cortical state \cite{scholvinck2015cortical}, which may be multi-dimensional and can be separated into short and long-timescale changes.
-In this paper, long-timescale fluctuations refer to large changes taking place over multiple trials caused by processes such as neural adaptation in response to repetitive stimulation or changes in the level of anaesthesia.
+In this paper, long-timescale fluctuations refer to large changes taking place over multiple trials caused by processes such as neural adaptation in response to repetitive stimulation, which occurs across species and modalities \cite{dragoi2002dynamics, ulanovsky2003processing, sharpee2006adaptive}, or changes in the level of anaesthesia.
 Short-timescale fluctuations refer to random trial-to-trial fluctuations in excitability around the current general level of excitability.
@@
 % which would introduce the possibility of state-dependent encoding and decoding (see below). 
-The characterisation controls for longer timescale changes in excitability to estimate the short-timescale state-modulated representation of a stimulus at a given point in time. 
-% and a novel prediction of the encoding capacity under a state-dependent spike time encoding regime. 
+The characterisation controls for longer timescale changes in excitability to estimate how stimulus representations are modulated by trial-to-trial fluctuations in state at a given point in time. 
+% This offers a novel prediction of the encoding capacity under a state-dependent spike time encoding regime. 
 Characterising spike timing codes is important to understanding a variety of dynamic and plastic processes, including properties of action potential patterns directly involved in cognition.
@@
 % Due to response variability, characterising the neural representation of a single stimulus is a matter of characterising the \textit{distribution} of possible neural responses for the stimulus. 
-A multi-neuron response can be expressed as a point in space, where each dimension represents the value of a spike train metric.
-Responses from multiple presentations of a stimulus form a sample stimulus-response distribution in this space, which can be used to estimate an underlying response distribution from which single-trial responses are `drawn' (Fig.
+A multi-neuron response can be expressed as a point in space, where each dimension represents a spike train metric.
+Responses to multiple stimulus presentations form a sample stimulus-response distribution in this space, which can be used to estimate an underlying response distribution from which single-trial responses are `drawn' (Fig.
+1a).
+% Information theoretic approaches aim to understand 
+If two sensory stimuli have a cognitively relevant difference, downstream neurons must be able to respond differently to samples from the response distribution of either stimulus.
+This is possible if the response distributions are highly non-overlapping in dimensions that downstream neurons are sensitive to, such as relative spike time differences (\cite{gasparini2006state, branco2010dendritic, branco2011synaptic}; Fig.
 1a).
@@
 
-% Information theoretic approaches aim to understand 
-If two sensory stimuli have a cognitively relevant difference, downstream neurons must be able to respond differently to samples from the response distribution of either stimulus.
-This is possible if the response distributions are highly non-overlapping in dimensions that downstream neurons are sensitive to, such as relative spike time differences (\cite{gasparini2006state, branco2010dendritic, branco2011synaptic}; Fig.
-1a).
-Neural coding and noise correlations are classically considered with respect to fixed-state decoders \cite{averbeck2006neural, moreno2014information, stringer2019high}.
-`Noise correlations' refer to shared trial-to-trial variability between the responses of multiple neurons over repeated presentations of single stimuli \cite{averbeck2006neural}.
-`Signal correlations' concern tuning similarity between neurons over multiple stimuli (see \cite{averbeck2006neural}).
+Neural coding classically considers such response-distributions with respect to fixed-state decoders \cite{averbeck2006neural, moreno2014information, stringer2019high}.
+`Noise correlations' refer to shared trial-to-trial variability between the responses of multiple neurons over repeated presentations of single stimuli, whilst `signal correlations' concern tuning similarity between neurons.
+With respect to fixed-state decoders, noise correlations can either increase or decrease encoding capacity \cite{averbeck2006neural}.
 
@@
 It is important to consider that if downstream neurons are also modulated by the same population-wide short-timescale fluctuating state, then downstream `decoder' neurons could respond in a state-dependent manner.
-When the short-timescale fluctuating state has a certain value on a single trial, a response comes from a \textit{state-conditioned} stimulus-response distribution, which would occupy a smaller region of response distribution space than the region sampled over multiple trials (Fig.
+When the fluctuating state has a value on a single trial, a response comes from a \textit{state-conditioned} stimulus-response distribution, which would occupy a smaller region of response distribution space than the region sampled over multiple trials (Fig.
 1c).
 State-dependent decoding would therefore offer an improved encoding capacity, compared to the classical fixed-state decoder interpretation.
-
-% Correlative structure between the responses of different neurons in single stimulus response distributions indicates the presence of trial-to-trial noise shared between the responses of multiple neurons.
-These are classically referred to as `noise correlations' \cite{averbeck2006neural}, rather than
 
@@
 
-Characterising noise correlations within stimulus-response distributions is complicated by the heterogeneity of sample response distribution structure and neural adaptation, which occurs across species and modalities \cite{dragoi2002dynamics, ulanovsky2003processing, sharpee2006adaptive}.
+Characterising noise correlations is complicated by the heterogeneity of sample response distribution structure and long-timescale changes 
 % That is, the response distributions are \textit{non-stationary} and change monotonically as a function of trial index for some neurons. 
 % Processes of adaptation, which we define as different to continuous latent states underlying trial-to-trial noise correlations, could also 
-Long-term changes in cortical state, such as adpatation, could lead to correlative structure in the stimulus-response distribution (Fig.
-X), and must be controlled for when characterising short-timescale or `adaptation-independent' noise correlations.
+which, could cause correlative structure in the stimulus-response distribution (Fig. X), and must be controlled for when characterising `adaptation-independent' noise correlations.
 
@@
 
-Using a large data set of spike sorted recordings in anesthetized rat barrel cortex, this paper aimed to precisely characterise the nature of short-timescale spike time noise correlations for first-spike patterns, whilst controlling for neural adaptation, which occurs in the barrel cortex to repetitive whisker stimulation for both spike count \cite{ahissar2000transformation, ahissar2001temporal, kheradpezhouh2017response, khatri2009stimulus, martin2014tactile, barros2019response} and spike latency \cite{ahissar2000transformation, ahissar2001temporal, kheradpezhouh2017response}.
+Using a large data set of spike sorted recordings in anesthetized rat barrel cortex, this paper aimed to precisely characterise the nature of short-timescale spike time noise correlations for first-spike patterns, by controlling for neural adaptation, which occurs in the barrel cortex to repetitive whisker stimulation for both spike count \cite{ahissar2000transformation, ahissar2001temporal, kheradpezhouh2017response, khatri2009stimulus, martin2014tactile, barros2019response} and spike latency \cite{ahissar2000transformation, ahissar2001temporal, kheradpezhouh2017response}.
 % towards estimating the potential encoding capacity of a state-dependent spike time encoding and determining the dimensionality of such an underlying latent state.
@@
 % using time series analysis, 
-indicative of short-timescale fluctuating state that affects different neurons differently.
+indicative of short-timescale fluctuating state, which varies the latency of different neurons at a different rate.
 % giving evidence of precise non-adaptive noise correlations
@@
 
-The characterisation of adaptation-independent clustered spike-timing noise correlations is useful under the fixed-state decoder interpretations where noise correlations can either increase or decrease encoding capacity \cite{averbeck2006neural}.
+The characterisation of adaptation-independent clustered spike-timing noise correlations is useful under the fixed-state decoder interpretations.
 More interestingly, these noise correlations open the possibility of state-dependent spike time coding.
-Factor analysis provided an estimate of state-dependent response distributions of first spike pairs (and relative first spike differences) for each adaptation-controlled correlated cluster.
-This gave a prediction of the resolution of possible state-dependent encoding, which surpasses previously proposed coding forms.
+Factor analysis provided an estimate of state-dependent response distributions of first spike pairs (and relative first spike differences) for each adaptation-controlled correlated cluster, giving a prediction of the capacity of state-dependent encoding, which surpasses previously proposed coding forms.
 
 Adaptation-independent non-unitary correlated clusters appeared for pairs of neurons spatially separated over multiple barrels and cortical layers.
-This offers suggestive evidence of a shared low-dimensional continuous latent state affecting the spike times of spatially distanced neurons, which in turn could also modulate downstream neurons and allow them to respond in a state-dependent manner.
+This offers suggestive evidence of a low-dimensional fluctuating state affecting the spike times of spatially distanced neurons, which in turn could modulate downstream neurons and cause them to respond in a state-dependent manner.
 % The number of low-dimensional continuous latent states underlying the correlative structure, however, must be confirmed. 
-Estimating whether cluster correlations observed for different pairs of neurons are modulated by a single or low-dimensional shared state was made difficult by low trial overlap between cluster.
-The work of this paper
+Estimating whether cluster correlations observed for different pairs of neurons are modulated by a single or low-dimensional shared state was made difficult by low trial overlap between clusters.
+This paper provides
 % in extracting and characterising adaptation-independent correlated clusters provides 
-a useful component step towards investigating this. 
+a useful step towards investigating this and shows that consideration of clustering, correlative structure and adaptation, together, are necessary to explain a large proportion of neural variability.
 % confirming or rejecting state-dependent spike-time coding.
-The results show that consideration of clustering, correlative structure and adaptation, together, are necessary to explain a large proportion of neural variability.
 

### 2020-08-12 manuscript.tex
End of day!

@@
 Previous studies have shown that spike count variability is shared across the population in early sensory cortices \cite{scholvinck2015cortical, lin2015nature}.
-Single cortical neurons can generate highly reliable spike trains to direct stimulation in vitro \cite{mainen1995reliability}.
-In vivo shared variability is therefore likely to arise from variability in synaptic inputs, reflecting cortical network dynamics \cite{carandini2004amplification}.
+As single cortical neurons can generate highly reliable spike trains to direct stimulation in vitro \cite{mainen1995reliability}, shared variability in vivo is likely to arise from variability in synaptic inputs reflecting cortical network dynamics \cite{carandini2004amplification}.
 
 In early visual cortex, trial-to-trial spike count variability is correlated across the population and is modulated by two global factors, which fluctuate the level of excitability from trial-to-trial \cite{lin2015nature}.
-The first of these is an additive factor, which is explained by the interaction of evoked activity with population-wide spontaneous activity \cite{arieli1996dynamics, scholvinck2015cortical}.
-The second is a multiplicative factor, which also modulates the trial-to-trial fluctuation in excitability across the population, and has been proposed to be caused by top-down feedback from higher cortices \cite{lin2015nature}.
-Fluctuations in spontaneous activity have also been shown to gate sensory-evoked responses \cite{luczak2013gating}.
-
-Fluctuations in trial-to-trial excitability are referred to herein as trial-to-trial fluctuations in continuous cortical state \cite{scholvinck2015cortical}, which may be multi-dimensional and can be separated into short and long-timescale changes.
-In this paper, long-timescale fluctuations refer to large changes taking place over multiple trials caused by processes such as neural adaptation in response to repetitive stimulation, which occurs across species and modalities \cite{dragoi2002dynamics, ulanovsky2003processing, sharpee2006adaptive}, or changes in the level of anaesthesia.
-Short-timescale fluctuations refer to random trial-to-trial fluctuations in excitability around the current general level of excitability.
+The first of these factors is explained by an additive interaction of evoked activity with population-wide spontaneous activity \cite{arieli1996dynamics, scholvinck2015cortical}.
+The second is a multiplicative factor proposed to be caused by top-down feedback from higher cortices \cite{lin2015nature}.
+Changes in excitability are herein referred to as changes in a continuous cortical state \cite{scholvinck2015cortical}.
+
+% Fluctuations in spontaneous activity also gate sensory-evoked responses \cite{luczak2013gating}.
+
+% which may be multi-dimensional. 
+Changes in such cortical state can happen on long or short-timescales, and can be either specific or general across the population (Fig.
+0a).
+In this paper, long-timescale state changes refer to slow changes in neuronal excitability caused by processes such as changes in the level of anaesthesia or neural adaptation \cite{dragoi2002dynamics, ulanovsky2003processing, sharpee2006adaptive}.
+Such processes are more specific than general across the population.
+Short-timescale state changes refer to random trial-to-trial fluctuations in excitability around the longer-timescale modulated state.
 
@@
 
-This study aims to characterise the non-trivial nature of trial-to-trial spike time (rather than spike count) noise correlations, as a component step towards investigating the modulation of spike times by short-timescale fluctuations of a low-dimensional continuous cortical state.
+% Study of the modulation of spike times by short-timescale changes in state are limited. 
+% Towards investigating whether spike times are modulated by short-timescale fluctuations of a low-dimensional cortical state, this study aims to characterise the nature of spike time noise correlations, whilst controlling for long-timescale changes in excitability.
+
+The paper takes a step towards uncovering the dimensionality of possible population-wide short-timescale state changes, by better characterising pairwise spike time noise correlations, whilst controlling for long-timescale state changes.
+% This study aims to characterise the non-trivial nature of trial-to-trial spike time noise correlations, as a component step towards investigating the modulation of spike times by short-timescale fluctuations of a low-dimensional continuous cortical state.
 % which would introduce the possibility of state-dependent encoding and decoding (see below). 
-The characterisation controls for longer timescale changes in excitability to estimate how stimulus representations are modulated by trial-to-trial fluctuations in state at a given point in time. 
-% This offers a novel prediction of the encoding capacity under a state-dependent spike time encoding regime. 
+% The characterisation controls for longer timescale changes in excitability to estimate how stimulus representations are modulated by trial-to-trial fluctuations in state at a given point in time. 
+This offers a novel prediction of the encoding capacity under a state-dependent coding regime. 
 Characterising spike timing codes is important to understanding a variety of dynamic and plastic processes, including properties of action potential patterns directly involved in cognition.
+Responses following sensory stimulation can be analysed to explore such principles of neural coding.
 
 %  low 2-dimensional continuous latent state .  
-
 % It is possible that spike time variability is caused and modulated across a population by such a low-dimensional continuous latent state.
@@
 
-Responses following sensory stimulation can be analysed to explore such principles of neural coding. 
 % Due to response variability, characterising the neural representation of a single stimulus is a matter of characterising the \textit{distribution} of possible neural responses for the stimulus. 
-A multi-neuron response can be expressed as a point in space, where each dimension represents a spike train metric.
-Responses to multiple stimulus presentations form a sample stimulus-response distribution in this space, which can be used to estimate an underlying response distribution from which single-trial responses are `drawn' (Fig.
+A multi-neuron stimulus response can be expressed as a point in multi-dimensional space, where each dimension represents the value of a spike train metric.
+Responses to multiple stimulus presentations form a sample stimulus-response distribution in this space.
+Neural coding classically considers such response-distributions with respect to fixed-state decoders \cite{averbeck2006neural, moreno2014information, stringer2019high}.
+All samples from the sample stimulus-response distribution are therefore used to estimate an underlying response distribution from which single-trial responses are `drawn' (Fig.
 1a).
+
 % Information theoretic approaches aim to understand 
@@
 1a).
+`Noise correlations' refer to shared trial-to-trial variability over repeated presentations of single stimuli, whilst `signal correlations' concern tuning similarity between neurons.
+With respect to fixed-state decoders, noise correlations can either increase or decrease encoding capacity \cite{averbeck2006neural}.
 
@@
 % which approximates an underlying stimulus-response distribution.
+
+It is important to consider that if downstream neurons are also modulated by the same population-wide short-timescale fluctuating state, then downstream `decoder' neurons could respond in a state-dependent manner.
+When the fluctuating state has a value on a single trial, a response comes from a \textit{state-conditioned} stimulus-response distribution, which would occupy a smaller region of response distribution space than the region sampled over multiple trials (Fig.
+1c).
+State-dependent decoding therefore offers an improved coding capacity.
+% , compared to the classical fixed-state decoder interpretation.
+
+Long-timescale changes in excitability make it difficult to accurately characterise the stimulus response distribution from which samples are drawn at a point in time. 
+% including noise correlation structure caused by short-timescale fluctuations in excitability. 
+Long-timescale changes are also likely to cause correlative structure in the stimulus-response distribution that could be misinterpreted as noise correlations caused by a short-timescale fluctuating state.
+If the effect of long-timescale changes in excitability can be removed or controlled for, it may be possible to give an estimate of the response distribution on a single trial.
+This could enable characterisation of noise correlations, which in turn could uncover short-timescale fluctuations in a population-wide state.
+Removing the effect of long-timescale changes and characterising subsequent adaptation-controlled response distributions, is made difficult by heterogeneity in single unit processes of adaptation and the structure of original sample response distributions.
+
+% That is, the response distributions are \textit{non-stationary} and change monotonically as a function of trial index for some neurons. 
+% Processes of adaptation, which we define as different to continuous latent states underlying trial-to-trial noise correlations, could also
 
@@
 
-Neural coding classically considers such response-distributions with respect to fixed-state decoders \cite{averbeck2006neural, moreno2014information, stringer2019high}.
-`Noise correlations' refer to shared trial-to-trial variability between the responses of multiple neurons over repeated presentations of single stimuli, whilst `signal correlations' concern tuning similarity between neurons.
-With respect to fixed-state decoders, noise correlations can either increase or decrease encoding capacity \cite{averbeck2006neural}.
-
 % Neural coding is classically interested in determining the dimensions in which response distributions do not overlap \cite{averbeck2006neural}.
@@
 % For example, a 2-dimensional continuous latent state modulates responses in V1 \cite{lin2015nature}.
-
-It is important to consider that if downstream neurons are also modulated by the same population-wide short-timescale fluctuating state, then downstream `decoder' neurons could respond in a state-dependent manner.
-When the fluctuating state has a value on a single trial, a response comes from a \textit{state-conditioned} stimulus-response distribution, which would occupy a smaller region of response distribution space than the region sampled over multiple trials (Fig.
-1c).
-State-dependent decoding would therefore offer an improved encoding capacity, compared to the classical fixed-state decoder interpretation.
 
@@
 
-Characterising noise correlations is complicated by the heterogeneity of sample response distribution structure and long-timescale changes 
-% That is, the response distributions are \textit{non-stationary} and change monotonically as a function of trial index for some neurons. 
-% Processes of adaptation, which we define as different to continuous latent states underlying trial-to-trial noise correlations, could also 
-which, could cause correlative structure in the stimulus-response distribution (Fig. X), and must be controlled for when characterising `adaptation-independent' noise correlations.
-
 The topographic representation of each single whisker on the snout of a rodent in the somatosensory cortex, the so-called barrel cortex, represents an excellent model to study noise correlations for controlled sensory stimulation \cite{feldmeyer2013barrel}.
@@
 
-Using a large data set of spike sorted recordings in anesthetized rat barrel cortex, this paper aimed to precisely characterise the nature of short-timescale spike time noise correlations for first-spike patterns, by controlling for neural adaptation, which occurs in the barrel cortex to repetitive whisker stimulation for both spike count \cite{ahissar2000transformation, ahissar2001temporal, kheradpezhouh2017response, khatri2009stimulus, martin2014tactile, barros2019response} and spike latency \cite{ahissar2000transformation, ahissar2001temporal, kheradpezhouh2017response}.
+Using a large dataset of spike sorted recordings in anesthetized rat barrel cortex, this paper aimed to precisely characterise the nature of short-timescale spike time noise correlations for first-spike patterns, by controlling for neural adaptation, which occurs in the barrel cortex to repetitive whisker stimulation for both spike count \cite{ahissar2000transformation, ahissar2001temporal, kheradpezhouh2017response, khatri2009stimulus, martin2014tactile, barros2019response} and spike latency \cite{ahissar2000transformation, ahissar2001temporal, kheradpezhouh2017response}.
 % towards estimating the potential encoding capacity of a state-dependent spike time encoding and determining the dimensionality of such an underlying latent state.
 The classical `stimulus-dependency' of noise correlations is left to future studies.
-A previous study analysing the same data set found precise multi-neuron spike patterns that were intra- and inter-laminar, as well intra- and intra-columnar \cite{reyes2014laminar}.
-Another study analysing the data set could decode stimulus properties (stimulation frequency and location) from spike counts and spike patterns but found only weak spike count noise correlations \cite{reyes2015high}.
+A previous study analysing the same dataset found precise multi-neuron spike patterns that were intra- and inter-laminar, as well intra- and intra-columnar \cite{reyes2014laminar}.
+Another study analysing the dataset could decode stimulus properties (stimulation frequency and location) from spike counts and spike patterns but found only weak spike count noise correlations \cite{reyes2015high}.
 % The spike time noise correlations uncovered in this study 
@@
 
-For repeated trials of a single stimulus, the first spike times of two units were plotted against each other for trials on which they both spiked.
+% \section{Results}
+
+For repeated trials of a single stimulus condition, the first spike times of two units were plotted against each other for trials on which they both spiked.
 For example, Fig.
 X shows first spike pairs for a pair of units (L5AI and L4I) in the principle column in response to a 4Hz stimulation of the corresponding principle whisker.
-First spike pairs appear to come from one of two clusters (blue and green dots) for presentations of the same stimulus.
-Moreover, the green cluster appears correlated with non-unitary gradient (different from $45^{\circ}$).
+First spike pairs appear to come from one of two clusters (blue and green dots) for presentations of the same stimulus condition.
+Moreover, the green cluster appears correlated.
 
@@
 1).
-A novel algorithm extracted isolated clusters from single stimulus first spike pair response distributions, demonstrating that temporally precise spike patterns fire on subsets of trials, as shown previously for the data set \cite{reyes2014laminar}.
+A novel algorithm extracted isolated clusters from single stimulus first spike pair response distributions, demonstrating that temporally precise spike patterns fire on subsets of trials, as shown previously for the dataset \cite{reyes2014laminar}.
 It was also shown, however, that first spike pairs within these isolated clusters were positively correlated above chance-levels.
@@
 
-% The paper aims to test that the occurrence of correlated clusters is above chance and to test the degree to which such correlations are caused by adaptative and non-adaptative processes.
-
-% (Fig X; see plots in the top left of Supplementary Vid. 1).
-
-Interestingly, many of the correlated clusters had non-unitary correlation angles. 
-Importantly, many clusters remained correlated with non-unitary gradient after controlling for longer-timescale neural adaptation,
-% using time series analysis, 
-indicative of short-timescale fluctuating state, which varies the latency of different neurons at a different rate.
-% giving evidence of precise non-adaptive noise correlations
-% Such non-unitary spike time correlation angles cannot be 
+Interestingly, many of the correlated clusters had non-unitary (different from $45^{\circ}$) correlation angles (including the green cluster in Fig. X).
+Importantly, many clusters remained correlated with non-unitary gradient after controlling for longer-timescale processes (including neural adaptation).
+Such non-unitary gradient correlations are indicative of a short-timescale fluctuating state varying the spike times of two neurons at a different rate, such that the relative difference between the spike times varies with respect to this state.
 This cannot explained by a trial-to-trial shift in the onset of the stimulus or cortical activity, which would produce correlations with a unitary gradient (i.e. a $45^{\circ}$ angle).
 
-This novel finding of `adaptation-independent clustered time-warp spike-timing noise correlations' shows that the distribution of decodable first spike time relative differences for pairs of neurons at a point in time are `warped' from trial-to-trial by some short-timescale fluctuating continuous cortical state.
+This novel finding is referred to as `adaptation-independent clustered time-warp spike time noise correlations'.
+Here, time-warp refers to the finding that pairwise relative spike time differences vary with respect to a state.
+% Particularly, within cluster spike times vary at different rates w.r.t. an underlying state, such that relative differences vary with res
+% by some short-timescale fluctuating continuous cortical state. 
+The total number of low-dimensional states underlying the correlative structure of different pairs across the population, however, must be confirmed. 
 % This phenomenon is referred to as `adaptation-independent clustered time-warped spike-timing noise correlations'. 
-This differs from previous `time-warp' studies, which aim to align spike trains over trials and do not simultaneously consider stimulus encoding, clustering, adaptation and different warping functions for different neurons \cite{williams2020discovering}.
-
-The characterisation of adaptation-independent clustered spike-timing noise correlations is useful under the fixed-state decoder interpretations.
-More interestingly, these noise correlations open the possibility of state-dependent spike time coding.
-Factor analysis provided an estimate of state-dependent response distributions of first spike pairs (and relative first spike differences) for each adaptation-controlled correlated cluster, giving a prediction of the capacity of state-dependent encoding, which surpasses previously proposed coding forms.
-
-Adaptation-independent non-unitary correlated clusters appeared for pairs of neurons spatially separated over multiple barrels and cortical layers.
-This offers suggestive evidence of a low-dimensional fluctuating state affecting the spike times of spatially distanced neurons, which in turn could modulate downstream neurons and cause them to respond in a state-dependent manner.
-% The number of low-dimensional continuous latent states underlying the correlative structure, however, must be confirmed. 
-Estimating whether cluster correlations observed for different pairs of neurons are modulated by a single or low-dimensional shared state was made difficult by low trial overlap between clusters.
-This paper provides
-% in extracting and characterising adaptation-independent correlated clusters provides 
-a useful step towards investigating this and shows that consideration of clustering, correlative structure and adaptation, together, are necessary to explain a large proportion of neural variability.
-% confirming or rejecting state-dependent spike-time coding.
+The findings differ from previous `time-warp' studies, which aim to align spike trains over trials and do not simultaneously consider potential encoding capacity, clustering, adaptation and different warping functions for different neurons \cite{williams2020discovering}.
+
+This characterisation is useful to the fixed-state decoder interpretation.
+More interestingly, these noise correlations open the possibility of state-dependent spike-time coding.
+Noise correlations were found for pairs of neurons separated over multiple barrels and cortical layers, giving evidence of a spatially distanced short-timescale fluctuating state, which could also cause downstream neurons to respond in a state-dependent manner.
+Analysis estimated state-dependent relative first spike difference response distributions for each adaptation-controlled correlated cluster, giving a prediction of the capacity of state-dependent encoding (based on distribution size), which surpasses previously proposed coding forms.
+
+% Whilst stimulus-dependency of noise correlations is not investigated in this study, estimates of potential encoding capacity are made by analysis of state-dependent response distribution size.
 
@@
 
-% Using a large data set of spike sorted recordings in anesthetized rat barrel cortex, the paper aims to address the following questions:
+% Using a large dataset of spike sorted recordings in anesthetized rat barrel cortex, the paper aims to address the following questions:
 
@@
 \ExecuteMetaData[AnalysisOutputLatex.tex]{ClusterCorrelationSummary0}
-The algorithm was verified by confirming that the algorithm extracted correlated clusters at chance levels for multiple control data sets (see Methods \ref{methods:correlation_surrogate_control_datasets}; Supplementary Fig.
+The algorithm was verified by confirming that the algorithm extracted correlated clusters at chance levels for multiple control datasets (see Methods \ref{methods:correlation_surrogate_control_datasets}; Supplementary Fig.
 3).
@@
 Testing the state-dependent decoding performance of the clusters is left for future studies, as it would require explicit knowledge of the underlying shared state or an estimate from the spike times of other neurons (which is difficult for the current extracted cluster counts per stimulus in single experimental sessions).
+Estimating whether cluster correlations observed for different pairs of neurons are modulated by a single or low-dimensional shared state was made difficult by low trial overlap between clusters. 
 Neuronal subthreshold membrane potentials would be a candidate for such state, as they are correlated between neurons across areas \cite{poulet2008internal} and affect action potential initiation \cite{petersen2003interaction}.
 The LFP may offer a useful proxy for the subthreshold membrane potential.
+
+This paper provides a useful step towards investigating this and shows that consideration of clustering, correlative structure and adaptation, together, are necessary to explain a large proportion of neural variability.
 
@@
 Here we present evidence of such possible codes.
-
-\subsection*{Non-stationarity}
-
-Future studies of non-stationary time-warped continua should carefully consider non-stationarity.
-Adaptation describes the `continuous recalibration of the sensory system to compensate for the changes in the statistics of the input stream' \cite{kheradpezhouh2017response}, with specific functional roles including gain modulation \cite{ahissar2000transformation, heiss2008shift, khatri2009stimulus, mohar2013opposite}, efficiency \cite{sharpee2006adaptive} and detectability of novel stimuli \cite{dragoi2002dynamics, ulanovsky2003processing}.
-The rate and degree of adaptation depends on stimulus parameters, neuron location and neuron properties (see \cite{kheradpezhouh2017response} for review).
-Adaptation modulates correlations and response heterogeneity (see \cite{kheradpezhouh2017response} for review), and depends on brain state and anaesthesia \cite{castro2004absence, katz2012trial}.
-Adaptation from repetitive stimulation changes excitatory-inhibitory balance \cite{higley2006balanced, heiss2008shift, malina2013imbalance} thought to be caused by short term synaptic depression (due to neurotransmitter depletion) observed in trigeminal complex \cite{mohar2013opposite}, thalamus \cite{deschenes2003relay} and cortex \cite{higley2006balanced, heiss2008shift, malina2013imbalance} (mostly at thalamocortical synapses \cite{chung2002short}).
-
-In mouse barrel cortex, spiking statistics change through development \cite{van2017layer}.
-Functional imaging in adult mice shows day-to-day maintenance of population activity \cite{margolis2012reorganization, mayrhofer2015sparse} but variability in the day-to-day stability of single unit representations \cite{mayrhofer2015sparse}.
-Functional imaging shows that in addition to tuning sensory maps, plasticity and sensory experience affects response gain, variability, firing correlations, and top-down modulation by task context (see \cite{lemessurier2018plasticity}).
-
-Mechanisms of attention \cite{cohen2010neuronal} and plasticity \cite{lemessurier2018plasticity}.
-% Collectively, these studies characterise the barrel cortex as a highly dynamic system, with representations continuously changing.
 

### 2020-08-18 manuscript.tex
Good!

@@
 
-% \usepackage[colorinlistoftodos]{todonotes}
-
-\usepackage[nomarkers,notablist,nofiglist]{endfloat}
+\usepackage[colorinlistoftodos]{todonotes}
+
+% \usepackage[nomarkers,notablist,nofiglist]{endfloat}
 \usepackage[colorinlistoftodos,disable]{todonotes}
@@
 The neural coding problem addresses the question of how information is encoded by patterns of action potentials in the nervous system \cite{gerstner1997neural, johnson2000neural, stein2005neuronal, gerstner2014neuronal}.
-Whilst precise temporal spike patterns have been observed \cite{uzzell2004precision, johansson2004first, gollisch2008rapid, storchi2012comparison, reinagel2002precise, reyes2014laminar}, a large proportion of spike time variability remains unexplained.
-Such variability leads to the common assumption that the brain uses a population code based on spike counts \cite{reyes2015high, lin2015nature} or trial-averaged firing rates \cite{hong2016explicit, chang2017code, kar2019evidence, stringer2019high}.
-Previous studies have shown that spike count variability is shared across the population in early sensory cortices \cite{scholvinck2015cortical, lin2015nature}.
-As single cortical neurons can generate highly reliable spike trains to direct stimulation in vitro \cite{mainen1995reliability}, shared variability in vivo is likely to arise from variability in synaptic inputs reflecting cortical network dynamics \cite{carandini2004amplification}.
-
-In early visual cortex, trial-to-trial spike count variability is correlated across the population and is modulated by two global factors, which fluctuate the level of excitability from trial-to-trial \cite{lin2015nature}.
+Responses following sensory stimulation can be analysed to explore principles of neural coding.
+Whilst precise temporal spike patterns have been observed \cite{uzzell2004precision, johansson2004first, gollisch2008rapid, storchi2012comparison, reinagel2002precise, reyes2014laminar}, a large proportion of spike time variability remains unexplained, leading to the common assumption that the brain uses a population code based on spike counts \cite{reyes2015high, lin2015nature} or trial-averaged firing rates \cite{hong2016explicit, chang2017code, kar2019evidence, stringer2019high}.
+Characterising spike timing codes is important to understanding a variety of dynamic and plastic processes, including properties of action potential patterns directly involved in cognition.
+
+As single cortical neurons can generate highly reliable spike trains to direct stimulation in vitro \cite{mainen1995reliability}, variability in vivo is likely to reflect 
+% arise from variability in synaptic inputs reflecting 
+cortical network dynamics \cite{carandini2004amplification}. 
+% Previous studies have shown that spike count variability is shared across the population in early sensory cortices . 
+For example, trial-to-trial spike count variability is shared across the population \cite{scholvinck2015cortical, lin2015nature} in early visual cortex.
+Particularly, fluctuations in the level of excitability from trial-to-trial was shown to be modulated by two global factors \cite{lin2015nature}.
 The first of these factors is explained by an additive interaction of evoked activity with population-wide spontaneous activity \cite{arieli1996dynamics, scholvinck2015cortical}.
 The second is a multiplicative factor proposed to be caused by top-down feedback from higher cortices \cite{lin2015nature}.
+
+% Fluctuations in spontaneous activity also gate sensory-evoked responses \cite{luczak2013gating}.
+
+% which may be multi-dimensional. 
 Changes in excitability are herein referred to as changes in a continuous cortical state \cite{scholvinck2015cortical}.
-
-% Fluctuations in spontaneous activity also gate sensory-evoked responses \cite{luczak2013gating}.
-
-% which may be multi-dimensional. 
-Changes in such cortical state can happen on long or short-timescales, and can be either specific or general across the population (Fig.
-0a).
-In this paper, long-timescale state changes refer to slow changes in neuronal excitability caused by processes such as changes in the level of anaesthesia or neural adaptation \cite{dragoi2002dynamics, ulanovsky2003processing, sharpee2006adaptive}.
-Such processes are more specific than general across the population.
+Such changes in continuous cortical state can occur on long or short-timescales, and may be specific to neurons or neuron-types or general across the population.
+% , and may occur across the population, or specifically for neurons or neuron-types (Fig. 0a). 
+In this paper, long-timescale state changes refer to slow changes in neuronal excitability caused by processes such as neural adaptation \cite{dragoi2002dynamics, ulanovsky2003processing, sharpee2006adaptive} or changes in the level of anaesthesia.
 Short-timescale state changes refer to random trial-to-trial fluctuations in excitability around the longer-timescale modulated state.
-
-% Other studies have reported only weak spike count correlations in sensory cortices \cite{reyes2015high}.
-
-% Study of the modulation of spike times by short-timescale changes in state are limited. 
-% Towards investigating whether spike times are modulated by short-timescale fluctuations of a low-dimensional cortical state, this study aims to characterise the nature of spike time noise correlations, whilst controlling for long-timescale changes in excitability.
-
-The paper takes a step towards uncovering the dimensionality of possible population-wide short-timescale state changes, by better characterising pairwise spike time noise correlations, whilst controlling for long-timescale state changes.
-% This study aims to characterise the non-trivial nature of trial-to-trial spike time noise correlations, as a component step towards investigating the modulation of spike times by short-timescale fluctuations of a low-dimensional continuous cortical state.
-% which would introduce the possibility of state-dependent encoding and decoding (see below). 
-% The characterisation controls for longer timescale changes in excitability to estimate how stimulus representations are modulated by trial-to-trial fluctuations in state at a given point in time. 
-This offers a novel prediction of the encoding capacity under a state-dependent coding regime. 
-Characterising spike timing codes is important to understanding a variety of dynamic and plastic processes, including properties of action potential patterns directly involved in cognition.
-Responses following sensory stimulation can be analysed to explore such principles of neural coding.
-
-%  low 2-dimensional continuous latent state .  
-% It is possible that spike time variability is caused and modulated across a population by such a low-dimensional continuous latent state.
-
-% \cite{gerstner1997neural}, which have capacity beyond the original spike count \cite{}. 
-% Continuous firing rates (averaged over time, trial or population ) 
-% These population codes assume that trial-to-trial variability is either independent or correlated across the population \cite{averbeck2006neural, lin2015nature}.
-
-% Continuous firing rates (averaged over time, trial or population \cite{gerstner1997neural}) have capacity beyond the single trial discrete spike count from which they are calculated, leaving a full understanding of neural coding and variability undetermined.
-
-% , under the premise that variability cancels out across a population \cite{abbott1999effect, moreno2014information, stringer2019high} or that variability is shared across the population and modulated from trial-to-trial by some underlying low-dimensional latent state \cite{lin2015nature}.
-
-% Characterising responses may also uncover response properties involved in behaviour.
-
-% Due to response variability, characterising the neural representation of a single stimulus is a matter of characterising the \textit{distribution} of possible neural responses for the stimulus. 
-A multi-neuron stimulus response can be expressed as a point in multi-dimensional space, where each dimension represents the value of a spike train metric.
-Responses to multiple stimulus presentations form a sample stimulus-response distribution in this space.
-Neural coding classically considers such response-distributions with respect to fixed-state decoders \cite{averbeck2006neural, moreno2014information, stringer2019high}.
-All samples from the sample stimulus-response distribution are therefore used to estimate an underlying response distribution from which single-trial responses are `drawn' (Fig.
-1a).
-
-% Information theoretic approaches aim to understand 
-If two sensory stimuli have a cognitively relevant difference, downstream neurons must be able to respond differently to samples from the response distribution of either stimulus.
-This is possible if the response distributions are highly non-overlapping in dimensions that downstream neurons are sensitive to, such as relative spike time differences (\cite{gasparini2006state, branco2010dendritic, branco2011synaptic}; Fig.
-1a).
-`Noise correlations' refer to shared trial-to-trial variability over repeated presentations of single stimuli, whilst `signal correlations' concern tuning similarity between neurons.
-With respect to fixed-state decoders, noise correlations can either increase or decrease encoding capacity \cite{averbeck2006neural}.
-
-% Estimating the underlying response distribution (including correlative structure) for single stimuli is complicated by neural adaptation and heterogeneity of sample response distribution structure.
-% which approximates an underlying stimulus-response distribution.
-
-It is important to consider that if downstream neurons are also modulated by the same population-wide short-timescale fluctuating state, then downstream `decoder' neurons could respond in a state-dependent manner.
-When the fluctuating state has a value on a single trial, a response comes from a \textit{state-conditioned} stimulus-response distribution, which would occupy a smaller region of response distribution space than the region sampled over multiple trials (Fig.
-1c).
-State-dependent decoding therefore offers an improved coding capacity.
-% , compared to the classical fixed-state decoder interpretation.
-
-Long-timescale changes in excitability make it difficult to accurately characterise the stimulus response distribution from which samples are drawn at a point in time. 
-% including noise correlation structure caused by short-timescale fluctuations in excitability. 
-Long-timescale changes are also likely to cause correlative structure in the stimulus-response distribution that could be misinterpreted as noise correlations caused by a short-timescale fluctuating state.
-If the effect of long-timescale changes in excitability can be removed or controlled for, it may be possible to give an estimate of the response distribution on a single trial.
-This could enable characterisation of noise correlations, which in turn could uncover short-timescale fluctuations in a population-wide state.
-Removing the effect of long-timescale changes and characterising subsequent adaptation-controlled response distributions, is made difficult by heterogeneity in single unit processes of adaptation and the structure of original sample response distributions.
-
-% That is, the response distributions are \textit{non-stationary} and change monotonically as a function of trial index for some neurons. 
-% Processes of adaptation, which we define as different to continuous latent states underlying trial-to-trial noise correlations, could also
+Analysis of neuronal responses over multiple trials may allow inference of a fluctuating shared short-timescale cortical state, with possible implications for neural coding.
 
@@
 \includegraphics[width=\textwidth]{Figure1.pdf}
-\caption{Explanatory Figures. 
-(a) Left: For two neurons (x and y-axis), illustration of sample (dots) and boundary of underlying (circle) first spike stimulus-response distributions for two hypothetical stimuli (red and gold).
+\caption{Adaptation-independent noise correlations, underlying state, response distributions.
+(a) Left: Blue lines show the average response of two neurons in response to multiple presentation of a single stimulus.
+Neuron 1 shows a consistent average response, whilst the responses of neuron 2 decrease over presentations, demonstrating a long-timescale response change.
+Pink points show the actual responses of each neuron on each trial.
+Centre: Plots the  By removing the long-timescale adaptative trend for neuron 2, it is poss}
+
+% (a) Left: For two neurons (x and y-axis), illustration of sample (dots) and boundary of underlying (circle) first spike stimulus-response distributions for two hypothetical stimuli (red and gold).
 Right: Illustration of corresponding relative first spike differences.
-(b) Illustration of trial dependence of a sampled response distribution (dots). Circle illustrates the boundary of a possible trial-dependent response distribution on a single trial. 
-(c) Illustration of boundary of state-dependent first spike response distribution for a single state value.
+% (b) Illustration of trial dependence of a sampled response distribution (dots). Circle illustrates the boundary of a possible trial-dependent response distribution on a single trial. 
+% (c) Illustration of boundary of state-dependent first spike response distribution for a single state value.
 The green line shows the mean points that would be predicted for a continuous set of state values.
-(d) Illustration of superposition of two single spike pattern feature representations.
-(e) Illustration of `stationary clustered time-warped precise spike timing'.
+% (d) Illustration of superposition of two single spike pattern feature representations.
+% (e) Illustration of `stationary clustered time-warped precise spike timing'.
 Left: Responses form several clear clusters.
@@
 
+A multi-neuron stimulus response can be expressed as a point in multi-dimensional space, where each dimension represents the value of a spike train metric.
+Responses to multiple stimulus presentations form a sample stimulus-response distribution in this space.
+Neural coding classically considers such response-distributions with respect to fixed-state decoders \cite{averbeck2006neural, moreno2014information, stringer2019high}.
+The sample stimulus-response distribution is therefore used to estimate an underlying response distribution from which single-trial responses are `drawn' (Fig.
+1a).
+
+% Information theoretic approaches aim to understand 
+If two sensory stimuli have a cognitively relevant difference, downstream neurons must be able to respond differently to samples from the response distribution of either stimulus.
+This is possible if the response distributions are highly non-overlapping in dimensions that downstream neurons are sensitive to, such as relative spike time differences (\cite{gasparini2006state, branco2010dendritic, branco2011synaptic}; Fig.
+1a).
+For a single stimulus, trial-to-trial variability shared between multiple neurons (i.e.
+`noise correlations') corresponds with correlative structure in the stimulus response distribution\footnote{Noise correlations differ to `signal correlations', which refer to tuning similarity between neurons.}.
+With respect to fixed-state decoders, noise correlations can either increase or decrease encoding capacity \cite{averbeck2006neural}.
+
+% Estimating the underlying response distribution (including correlative structure) for single stimuli is complicated by neural adaptation and heterogeneity of sample response distribution structure.
+% which approximates an underlying stimulus-response distribution.
+
+However, short-timescale noise correlations are suggestive of a short-timescale fluctuating state, which covaries neuronal responses and would thus explain a proportion of neural variability.
+If downstream neurons were also modulated by the same population-wide short-timescale fluctuating state, then downstream `decoder' neurons could respond in a state-dependent manner.
+When the fluctuating state has a value on a single trial, a response comes from a \textit{state-conditioned} stimulus-response distribution, which would occupy a smaller region of response distribution space than the region sampled over multiple trials (Fig.
+1c).
+State-dependent decoding therefore offers an improved coding capacity.
+
+This paper aims to precisely characterise shared short-timescale trial-to-trial spike-timing response variability as a step towards inferring the dimensionality of shared short-timescale trial-to-trial state fluctuations (Fig.
+1a).
+This is complicated by changes to neural responses caused by longer-timescale processes, which are therefore controlled for in this paper.
+This offers a novel prediction of the encoding capacity under a state-dependent coding regime.
+
+% Long-timescale changes are also likely to cause correlative structure in the stimulus-response distribution that could be misinterpreted as noise correlations caused by a short-timescale fluctuating state.
+If the effect of long-timescale changes in excitability can be removed or controlled for, it may be possible to give an estimate of the response distribution on a single trial.
+This could enable characterisation of noise correlations, which in turn could uncover short-timescale fluctuations in a population-wide state.
+Removing the effect of long-timescale changes and characterising subsequent adaptation-controlled response distributions, is made difficult by heterogeneity in single unit processes of adaptation and the structure of original sample response distributions.
+
+% That is, the response distributions are \textit{non-stationary} and change monotonically as a function of trial index for some neurons. 
+% Processes of adaptation, which we define as different to continuous latent states underlying trial-to-trial noise correlations, could also
+
 % Neural coding is classically interested in determining the dimensions in which response distributions do not overlap \cite{averbeck2006neural}.
@@
 Analysis estimated state-dependent relative first spike difference response distributions for each adaptation-controlled correlated cluster, giving a prediction of the capacity of state-dependent encoding (based on distribution size), which surpasses previously proposed coding forms.
+
+% Other studies have reported only weak spike count correlations in sensory cortices \cite{reyes2015high}.
+
+% Study of the modulation of spike times by short-timescale changes in state are limited. 
+% Towards investigating whether spike times are modulated by short-timescale fluctuations of a low-dimensional cortical state, this study aims to characterise the nature of spike time noise correlations, whilst controlling for long-timescale changes in excitability.
+
+% The paper takes a step towards uncovering the dimensionality of possible population-wide short-timescale state changes, by better characterising pairwise spike time noise correlations, whilst controlling for long-timescale state changes.
+
+% This study aims to characterise the non-trivial nature of trial-to-trial spike time noise correlations, as a component step towards investigating the modulation of spike times by short-timescale fluctuations of a low-dimensional continuous cortical state.
+% which would introduce the possibility of state-dependent encoding and decoding (see below). 
+% The characterisation controls for longer timescale changes in excitability to estimate how stimulus representations are modulated by trial-to-trial fluctuations in state at a given point in time.
+
+%  low 2-dimensional continuous latent state .  
+% It is possible that spike time variability is caused and modulated across a population by such a low-dimensional continuous latent state.
+
+% \cite{gerstner1997neural}, which have capacity beyond the original spike count \cite{}. 
+% Continuous firing rates (averaged over time, trial or population ) 
+% These population codes assume that trial-to-trial variability is either independent or correlated across the population \cite{averbeck2006neural, lin2015nature}.
+
+% Continuous firing rates (averaged over time, trial or population \cite{gerstner1997neural}) have capacity beyond the single trial discrete spike count from which they are calculated, leaving a full understanding of neural coding and variability undetermined.
+
+% , under the premise that variability cancels out across a population \cite{abbott1999effect, moreno2014information, stringer2019high} or that variability is shared across the population and modulated from trial-to-trial by some underlying low-dimensional latent state \cite{lin2015nature}.
+
+% Characterising responses may also uncover response properties involved in behaviour.
+
+% Due to response variability, characterising the neural representation of a single stimulus is a matter of characterising the \textit{distribution} of possible neural responses for the stimulus.
 

### 2020-08-18 manuscript.tex
Lunch

@@
 
-However, short-timescale noise correlations are suggestive of a short-timescale fluctuating state, which covaries neuronal responses and would thus explain a proportion of neural variability.
-If downstream neurons were also modulated by the same population-wide short-timescale fluctuating state, then downstream `decoder' neurons could respond in a state-dependent manner.
+However, short-timescale noise correlations are suggestive of a short-timescale fluctuating state, which covaries neuronal responses.
+If downstream neurons were also modulated by the same short-timescale fluctuating state, then downstream `decoder' neurons could respond in a state-dependent manner.
 When the fluctuating state has a value on a single trial, a response comes from a \textit{state-conditioned} stimulus-response distribution, which would occupy a smaller region of response distribution space than the region sampled over multiple trials (Fig.
 1c).
-State-dependent decoding therefore offers an improved coding capacity.
-
-This paper aims to precisely characterise shared short-timescale trial-to-trial spike-timing response variability as a step towards inferring the dimensionality of shared short-timescale trial-to-trial state fluctuations (Fig.
+State-dependent decoding would therefore offer an improved coding capacity.
+
+This paper aims to precisely characterise shared short-timescale trial-to-trial spike-timing noise correlations as a step towards inferring the dimensionality of shared short-timescale trial-to-trial state fluctuations (Fig.
 1a).
-This is complicated by changes to neural responses caused by longer-timescale processes, which are therefore controlled for in this paper.
-This offers a novel prediction of the encoding capacity under a state-dependent coding regime.
+If the state fluctuations were low dimensional across a population, then it is more likely that decoding neurons within or outside the population would be modulated by the same fluctuations, and could therefore respond with improved coding capacity in a state-dependent manner.
+Characterisation of noise correlations is complicated by changes to neural responses caused by longer-timescale processes.
+Such longer-timescale changes are therefore are controlled for in this paper, offering a prediction of encoding capacity under a state-dependent coding regime.
 

### 2020-08-18 manuscript.tex
Leaving library

@@
 Changes in excitability are herein referred to as changes in a continuous cortical state \cite{scholvinck2015cortical}.
-Such changes in continuous cortical state can occur on long or short-timescales, and may be specific to neurons or neuron-types or general across the population.
+Such changes in continuous cortical state can occur on long or short-timescales, and may be specific to neurons or neuron-types or general across the population (Fig. 1a).
 % , and may occur across the population, or specifically for neurons or neuron-types (Fig. 0a). 
-In this paper, long-timescale state changes refer to slow changes in neuronal excitability caused by processes such as neural adaptation \cite{dragoi2002dynamics, ulanovsky2003processing, sharpee2006adaptive} or changes in the level of anaesthesia.
-Short-timescale state changes refer to random trial-to-trial fluctuations in excitability around the longer-timescale modulated state.
-Analysis of neuronal responses over multiple trials may allow inference of a fluctuating shared short-timescale cortical state, with possible implications for neural coding.
+In this paper, long-timescale state changes refer to slow changes in neuronal excitability caused by processes such as neural adaptation \cite{dragoi2002dynamics, ulanovsky2003processing, sharpee2006adaptive} or changes in the level of anaesthesia (Fig.
+1a).
+Short-timescale state changes refer to random trial-to-trial fluctuations in excitability around the longer-timescale modulated state (Fig.
+1a).
+Analysis of neuronal responses over multiple trials may allow inference of a fluctuating shared short-timescale cortical state, with possible implications for neural coding (Fig.
+1a).
 
@@
 A multi-neuron stimulus response can be expressed as a point in multi-dimensional space, where each dimension represents the value of a spike train metric.
-Responses to multiple stimulus presentations form a sample stimulus-response distribution in this space.
+Responses to multiple stimulus presentations form a sample stimulus-response distribution in this space (Fig.
+1b).
 Neural coding classically considers such response-distributions with respect to fixed-state decoders \cite{averbeck2006neural, moreno2014information, stringer2019high}.
 The sample stimulus-response distribution is therefore used to estimate an underlying response distribution from which single-trial responses are `drawn' (Fig.
-1a).
+1b).
 
@@
 This is possible if the response distributions are highly non-overlapping in dimensions that downstream neurons are sensitive to, such as relative spike time differences (\cite{gasparini2006state, branco2010dendritic, branco2011synaptic}; Fig.
-1a).
-For a single stimulus, trial-to-trial variability shared between multiple neurons (i.e.
-`noise correlations') corresponds with correlative structure in the stimulus response distribution\footnote{Noise correlations differ to `signal correlations', which refer to tuning similarity between neurons.}.
+1c).
+For a single stimulus, trial-to-trial variability shared between multiple neurons (`noise correlations'\footnote{Noise correlations differ to `signal correlations', which refer to tuning similarity between neurons.}) corresponds with correlative structure in the stimulus response distribution when the dimensions correspond to spike train metrics for different individual neurons (Fig.
+1d).
 With respect to fixed-state decoders, noise correlations can either increase or decrease encoding capacity \cite{averbeck2006neural}.
@@
 However, short-timescale noise correlations are suggestive of a short-timescale fluctuating state, which covaries neuronal responses.
-If downstream neurons were also modulated by the same short-timescale fluctuating state, then downstream `decoder' neurons could respond in a state-dependent manner.
+If downstream neurons were also modulated by the same short-timescale fluctuating state, then downstream `decoding' neurons could respond in a state-dependent manner.
 When the fluctuating state has a value on a single trial, a response comes from a \textit{state-conditioned} stimulus-response distribution, which would occupy a smaller region of response distribution space than the region sampled over multiple trials (Fig.
-1c).
+1e).
 State-dependent decoding would therefore offer an improved coding capacity.
 
-This paper aims to precisely characterise shared short-timescale trial-to-trial spike-timing noise correlations as a step towards inferring the dimensionality of shared short-timescale trial-to-trial state fluctuations (Fig.
-1a).
-If the state fluctuations were low dimensional across a population, then it is more likely that decoding neurons within or outside the population would be modulated by the same fluctuations, and could therefore respond with improved coding capacity in a state-dependent manner.
-Characterisation of noise correlations is complicated by changes to neural responses caused by longer-timescale processes.
+If the state fluctuations were low dimensional across a population, then it is more likely that decoding neurons within or outside the population would be modulated by the fluctuations modulating afferently connected encoding neurons, and could therefore respond with improved coding capacity in a state-dependent manner.
+This paper aims to precisely characterise shared short-timescale trial-to-trial spike-timing noise correlations, towards a better understanding of stimulus representations and as an important step towards inferring the dimensionality of shared short-timescale trial-to-trial state fluctuations.
+Characterisation of noise correlations is complicated by longer-timescale changes to neural responses.
 Such longer-timescale changes are therefore are controlled for in this paper, offering a prediction of encoding capacity under a state-dependent coding regime.
